---
title: "Hemochromatosis exploration in FinnGen"
author: "Jarkko Toivonen"
date: "`r Sys.Date()`"
format: 
  html:
    df-print: kable
    toc: true
    toc-depth: 4
    embed-resources: true
---

## TODO

- DONE Other diagnoses before and after first hc diagnosis
- DONE Enrichment/european MAF for snips
- DONE Download document, smooth distributions
- Confidence intervals for dosage counts?
- DONE More snips in the analysis and do LD and HWE checks for them

## Setup

```{r setup}
library(tidyverse)
library(glue)
library(lubridate)
library(patchwork)
# Data manipulation related
#library(data.table)
library(dbplyr)
library(openxlsx)
library(broom)
library(rsample)

# BigQuery connection related
library(bigrquery)
library(DBI)

source("common.R")

release <- "r12"; Release <- str_to_upper(release)

result_path <- file.path("../results", release)
cat(sprintf("result_path=%s", result_path), 
    sprintf("release=%s", release),
    sep = "\n",
    file = "../data/bash_init")  # Store the variables for bash chunks

data_path <- "../data"
figure_path <- file.path(result_path, "figures")
table_path <- file.path(result_path, "tables")
c(figure_path, table_path) %>% discard(dir.exists) %>% walk(~dir.create(., recursive = TRUE))  # Create dirs if they don't exist
recompute <- FALSE
save_figs <- FALSE
save_tables <- FALSE


#no_individual_level_data <- TRUE
allow_individual_level_data <- FALSE
ldstore <- "~/usr/bin/ldstore_v2.0_x86_64"
bgenix <- "~/usr/bin/bgenix"
```

## Functions



```{r}
#| code-fold: true
# cut_helper reduces the number of breaks so that the bin sizes are all at least 10.
# It achieves this by dropping the small bins from both ends of the distributions,
# and forming new bins to this more narrow range.
# This is suboptimal solution.
cut_helper <- function(v) {
  # Get the count in each bin
  x <- table(cut(v, breaks=30)) %>%
    as_tibble(.name_repair="unique") %>% 
    rename(f="...1")
  # Extract the start and end point of each bin
  x2 <- x %>% 
    extract(col = f, regex = "^\\((.*),(.*)]$", into=c("start", "end")) %>%
    mutate(across(c(start, end), as.double))
  x3 <- x2 %>% filter(n >= 10)   # Drop small bins
  #x3 <- x2 %>% filter(n >= 5)
  n <- nrow(x3)  # number of bins
  r <- c(min(x3$start), max(x3$end))  # New, more narrow range of data
  bin_width <- (r[2] - r[1]) / n      
  breaks <- seq(r[1], r[2], length.out=n+1) # Cut the range into bins
  list(r=r, n=n, breaks=breaks)
}
my_histogram <- function(df, name="CHILD_WEIGHT") {
  #L <- cut_helper(df, name)
  L <- cut_helper(df[[name]])
  df %>% ggplot(aes(.data[[name]])) + geom_histogram(breaks=L$breaks)
}

```

## Load data sets

### Extract hemochromatosis cases

There seems to be problem with the BigQuery system. The following doesn't work.

```{r, eval=FALSE}
library(bigrquery)
library(dplyr)
library(dbplyr)
# Running some authentication code:
projectid <- "fg-production-sandbox-5" # NB: ADD YOUR SANDBOX NUMBER!
options(gargle_oauth_cache = FALSE)
bq_auth(scopes = "https://www.googleapis.com/auth/bigquery")

# Making a connection object that is used to connect to the tables:

con = dbConnect(
  bigrquery::bigquery(),
  project = "finngen-production-library",
  dataset = if (release == "r12") "sandbox_tools_r12" else "sandbox_tools_r13",
  billing = projectid
)

if (release == "r12") {
  endpoint = tbl(con, "endpoint_cohorts_r12_v1")
} else {
  endpoint = tbl(con, "endpoint_cohorts_r13_v3")
}
#endpoint = tbl(con, "finngen_r12_service_sector_detailed_longitudinal_v1")
#info = tbl(con, "finngenid_info_r11_v1")
#E4_IRON_MET
```

```{r, eval=FALSE}
tmp <- endpoint %>% filter(ENDPOINT == "E4_IRON_MET") %>% 
  count(CONTROL_CASE_EXCL) %>% 
  collect()
tmp
```

BigQuery doesn't seem to work, I will subset endpoints from command line.

```{r}
#endpoint_dir <- "/finngen/library-red/finngen_R12/endpoint_cohorts_v1.0"
if (release == "r12") {
  input_file <- "/finngen/library-red/finngen_R12/endpoint_cohorts_v1.0/endpoint_cohorts_r12_v1.csv.gz"
} else {
  input_file <- "/finngen/library-red/finngen_R13/endpoint_cohorts_v3.0/finngen_R13_endpoint_cohorts_r13_v3.csv.gz"
}
output_file <- file.path(result_path, "hemochromatosis.tsv.gz")
cmd <- glue("zcat {input_file} | awk 'NR==1{{print}}{{if ($2 == \"E4_IRON_MET\") print}}' | gzip > {output_file}")
print(cmd)
if (!file.exists(output_file))
  system(cmd)
```

|Field|Description|
|---|---| 
| FINNGENID | Subject id|  
| ENDPOINT | endpoint name|
| CONTROL_CASE_EXCL | Indicates if patient is control=0, case=1, excluded control=2 |
| AGE | If case, the age of the first event now with three decimals to provide more accuracy |
| APPROX_EVENT_DAY | approximate date of the first event |
| NEVT | If case, the number of events |

I assume that the AGE is at death or EOF for controls!

```{r Read hemochromatosis}
filename <- file.path(result_path, "hemochromatosis.tsv.gz")
hc <- read_tsv(filename, n_max = Inf, col_types = cols(
  FINNGENID = col_character(),
  ENDPOINT = col_character(),
  CONTROL_CASE_EXCL = col_double(),
  AGE = col_double(),
  APPROX_EVENT_DAY = col_date(),
  NEVT = col_double()))
```

Number of controls, cases and exclusions.

```{r}
nrow(hc)   # This is the full R12 data
```


```{r}
#hc %>% select(CONTROL_CASE_EXCL) %>% table()
# 0 == control
# 1 == case
# 2 == exclusion
hc %>% pull(CONTROL_CASE_EXCL) %>% as.factor() %>% fct_count(prop=TRUE)
```

```{r}
hc <- hc %>% 
  rename(hemochromatosis = CONTROL_CASE_EXCL,
         sample = FINNGENID) %>%
  filter(hemochromatosis != 2) %>%
  mutate(hemochromatosis = as.factor(case_when(hemochromatosis == 1 ~ "case",
                                               TRUE  ~  "control"))) %>%
  select(-ENDPOINT)
```

### Venesections

What codes are used for phlebotomy and venesection?

```{bash}
grep -Ei "FG_CODE1|phlebotomy|venesection" /finngen/library-green/medical_codes/fg_codes_info_v7/fg_codes_info.csv > /tmp/procedures.tsv
```

```{r}
procedure_codes <- read_tsv("/tmp/procedures.tsv", show_col_types = FALSE)
procedure_codes %>% select(-c(FG_CODE2:FG_CODE9, extra_info)) # Drop the NA columns
```

Get the venesections and phlebotomies from detailed longitudinal file.

```{bash}
. ../data/bash_init  # Read some bash variables
#OUTDIR=/home/ivm/projects/hemochromatosis/results
if [[ $release == "r12" ]] ; then
  filename=/finngen/library-red/finngen_R12/phenotype_1.0/data/finngen_R12_detailed_longitudinal_1.0.txt.gz
else
  filename=/finngen/library-red/finngen_R13/phenotype_1.0/data/finngen_R13_detailed_longitudinal_1.0.txt.gz
fi
result=$result_path/venesections.tsv
zresult=$result.gz
# awk is slow, let's use grep instead. BigQuery would be even better, or parquet
if [[ ! -f $zresult ]] ; then
  echo Creating file $zresult from $filename
  zcat  $filename | head -n 1 > $result
  zcat  $filename | grep -Ew "TPH00|SPAT1074" >> $result
  gzip $result
fi
```

The above may give warning "gzip: stdout: Broken pipe", but that is nothing to worry about.

### Dosages

Define interesting snips.

```{r Snips}
# Hemochromatosis mutations
# rs1799945, c.187C>G, p.His63Asp (H63D)
# rs1800562, c.845G>A, p. Cys282Tyr (C282Y)
snips <- tribble(#~id, ~name, ~rsid, ~chr, ~pos, ~ref, ~alt,
  ~SNPID, ~name, ~rsid, ~chromosome, ~position, ~first, ~second,
  "chr6_26071867_A_C", "H1-2",      "rs9968910", 6, 26071867, "A", "C",
  "chr6_26090951_C_G", "H63D",      "rs1799945", 6, 26090951, "C", "G",
  "chr6_26090957_A_T", "S65C",      "rs1800730", 6, 26090957, "A", "T",
  "chr6_26092913_G_A", "C282Y",     "rs1800562", 6, 26092913, "G", "A",
  "chr6_32636375_A_G", "supersnip", "rs9272324", 6, 32636375, "A", "G")
snip_names <- snips %>% pull(name)
conv <- snips %>% select(name, SNPID) %>% deframe()
snips_filename <- file.path(data_path, "snips.tsv")
if (! file.exists(snips_filename))
  write_tsv(snips %>% select(SNPID), snips_filename, col_names = FALSE)
```

Extract the variants.

```{r Extract variants}
# qctool -g input.bgen -s input.sample -og output.vcf -incl-samples-where 'ID_1 = 123455'
# /finngen/library-red/finngen_R12/bgen_2.0/data/chrom/finngen_R12_6.bgen
if (release == "r12") {
  bgen_base <- "/finngen/library-red/finngen_R12/bgen_2.0/data/chrom"
} else {
  bgen_base <- "/finngen/library-red/finngen_R13/bgen_1.0/data/chrom"
}
input_bgen <- file.path(bgen_base, glue("finngen_{str_to_upper(release)}_6.bgen"))
input_sample <- file.path(bgen_base, glue("finngen_{str_to_upper(release)}_6.bgen.sample"))
output <- file.path(result_path, "snips.bgen")

# Use bgenix, it is faster
cmd <- glue("{bgenix} -g {input_bgen} -incl-rsids {snips_filename} > {output}")
# qctool is slower
#cmd <- glue("qctool -g {input_bgen} -s {input_sample} -og {output} -ofiletype bgen -incl-snpids {snips_filename}")
# The -incl-rsids option didn't seem to work. Maybe the rsids are not stored in the FinnGen bgens

# One can list the snip names in a bgen file with command "bgenix -g input.bgen -list"
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
# Create the index
cmd <- glue("{bgenix} -g {output} -index")
print(cmd)
if (!file.exists(file.path(result_path, "snips.bgen.bgi"))) {
  system(cmd)
}
```

qctool took 9m16.454s
bgenix took 0m4.236s

Extract dosages.

```{r Extract dosages}
input_bgen <- file.path(result_path, "snips.bgen")
#input_sample <- "../results/snips.sample"   # This file does not exist
output <- file.path(result_path, "snips.dosage")
cmd <- glue("qctool -g {input_bgen} -og {output} -ofiletype dosage -incl-snpids {snips_filename}")
#cmd <- glue("qctool -g {input_bgen} -s {input_sample} -og {output} -ofiletype dosage -incl-snpids {snips_filename}")
# The -incl-rsids option didn't seem to work. Maybe the rsids are not stored in the FinnGen bgens
# One can list the snip names in a bgen file with command "bgenix -g input.bgen -list"
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
```

Read dosages.

```{r Read snip dosages}
dosages_filename <- file.path(result_path, "snips.dosage")
dosages <- read_delim(dosages_filename, delim=" ", col_types = cols(position = col_integer(), .default = col_character()))
dosages <- dosages %>% 
  select(-c(chromosome, rsid, position, alleleA, alleleB)) %>% 
  pivot_longer(cols=-c("SNPID"), names_to = "sample", values_to = "dosage") %>% 
  mutate(dosage = as.integer(dosage)) %>%
  inner_join(snips %>% select(SNPID, rsid, name), by="SNPID")
dosages_wide <- dosages %>% pivot_wider(names_from=name, values_from=dosage, id_cols = sample)
```

Compute snpstats using qctool.

```{r Compute snpstats}
input_file <- file.path(result_path, "snips.bgen")
output_file <- file.path(result_path, "snp-stats.txt")
cmd <- glue("qctool -g {input_file} -snp-stats -osnp {output_file}")
print(cmd)
if (! file.exists(output_file)) {
  system(cmd)
}
```

Read snp-stats.

```{r}
stats <- read_delim(output_file, delim=" ", comment="#")
```



```{r}
stats %>% 
  select(-c(rsid, chromosome, position, alleleA, alleleB, comment, minor_allele, major_allele)) %>%
  pivot_longer(cols=-alternate_ids) %>% 
  pivot_wider(names_from=alternate_ids) %>%
  rename(all_of(conv)) %>%
  print(n=Inf)
```

There is more evidence of C282Y being in linkage disequilibrium. Are all five snips in
Hardy-Weinberg disequilibrium?

### Certain hla allele dosages

Extract dosages of the following alleles.

```{r}
hla_alleles <- tibble(allele = c(
  "F*01:01",
  "F*01:03",
  "G*010101c",
  "G*UTR-4",
  "G*01:01:01:05",
  "A*03:01",
  "C*07:02",
  "B*07:02"
))
```

```{r Extract nonclassical HLA dosages}
hla_alleles_filename <- file.path(data_path, "hla_alleles.tsv")
if (! file.exists(hla_alleles_filename))
  write_tsv(hla_alleles %>% select(allele), hla_alleles_filename, col_names = FALSE)
input_bgen <- file.path(data_path, "R12_ncHLA.bgen")
#input_sample <- "../results/snips.sample"   # This file does not exist
output <- file.path(result_path, "hla_5_alleles.dosage")
cmd <- glue("qctool -g {input_bgen} -og {output} -ofiletype dosage -incl-rsids {hla_alleles_filename}")
# One can list the snip names in a bgen file with command "bgenix -g input.bgen -list"
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
```

```{r Extract classical HLA dosages}
hla_alleles_filename <- file.path(data_path, "hla_alleles.tsv")
if (! file.exists(hla_alleles_filename))
  write_tsv(hla_alleles %>% select(allele), hla_alleles_filename, col_names = FALSE)
input_bgen <- glue("/finngen/library-red/finngen_{Release}/hla_1.0/bgen/{Release}_HLA.bgen")
#input_sample <- "../results/snips.sample"   # This file does not exist
output <- file.path(result_path, "hla_3_alleles.dosage")
cmd <- glue("qctool -g {input_bgen} -og {output} -ofiletype dosage -incl-rsids {hla_alleles_filename}")
# One can list the snip names in a bgen file with command "bgenix -g input.bgen -list"
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
```

Read dosages. 

```{r Read hla dosages}
classical_hla_filename <- file.path(result_path, "hla_3_alleles.dosage")
nonclassical_hla_filename <- file.path(result_path, "hla_5_alleles.dosage")
classical_hla_dosages <- read_delim(classical_hla_filename, delim=" ", col_types = cols(position = col_integer(), .default = col_character()))
nonclassical_hla_dosages <- read_delim(nonclassical_hla_filename, delim=" ", col_types = cols(position = col_integer(), .default = col_character()))
hla_dosages <- bind_rows(classical_hla_dosages, nonclassical_hla_dosages) %>% 
  select(-c(chromosome, rsid, position, alleleA, alleleB)) %>% 
  pivot_longer(cols=-c("SNPID"), names_to = "sample", values_to = "dosage") %>% 
  mutate(dosage = as.integer(dosage)) #%>%
  #inner_join(hla_alleles %>% select(allele, rsid, name), by="SNPID")
hla_dosages_wide <- hla_dosages %>% pivot_wider(names_from=SNPID, values_from=dosage, id_cols = sample)
hla_dosages_wide <- hla_dosages_wide %>% mutate(sample = str_remove(sample, "_.*$"))
```

### HLA snip dosages

```{r}
hc_significant <- read_csv("../data/E4_IRON_MET_R12_traditional.csv")
hc_significant <- hc_significant %>% 
  rename(pos_hg38 = "pos...2",
         rsid = rsids) %>% 
  mutate(SNPID = sprintf("chr%i_%i_%s_%s", chrom, pos_hg38, ref, alt), .before = 1)
hc_significant
```



```{r}
tmp_conv <- c(
  rs114179634 = "HLA1",
  rs9261354   = "HLA4",
  rs3094093   = "HLA2")
hla_snip_info <- hc_significant %>% 
  filter(chrom==6 & 28510120 <= pos_hg38 & pos_hg38 <= 33480577, # HLA area
         pval < 5e-8) %>% arrange(pos_hg38) %>%
    mutate(name = tmp_conv[rsid], .after = SNPID)
filename <- file.path(data_path, "hla_snip_info.tsv")
write_tsv(hla_snip_info, filename)
hla_snip_info
```

The snip rs9261354 is correlated with other HLA snip and causes high VIF if included in the model. Hence, I won't
include it in the model. In addition, the snip is not available in UKBB.

Write snip ids to a temp file.

```{r}
snip_id_filename <- "/tmp/hla_snips_ids.tsv"
if (! file.exists(snip_id_filename))
  write_tsv(hla_snip_info %>% select(SNPID), snip_id_filename, col_names = FALSE)
```

Extract the variants with bgenix.

```{r Extract variants2}
#bgen_base <- "/finngen/library-red/finngen_R12/bgen_2.0/data/chrom"
if (release == "r12") {
  input <- "/finngen/library-red/finngen_R12/bgen_2.0/data/chrom/finngen_R12_6.bgen"
} else {
  input <- "/finngen/library-red/finngen_R13/bgen_1.0/data/chrom/finngen_R13_6.bgen"
}
input_sample <- paste0(input, ".sample")
output <- file.path(result_path, "hla_snips.bgen")

cmd <- glue("{bgenix} -g {input} -incl-rsids {snip_id_filename} > {output}")
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
# Create the index
cmd <- glue("{bgenix} -g {output} -index")
print(cmd)
if (!file.exists(file.path(result_path, "hla_snips.bgen.bgi"))) {
  system(cmd)
}
```

qctool took 9m16.454s
bgenix took 0m4.236s

Extract dosages.

```{r Extract dosages2}
input_bgen <- file.path(result_path, "hla_snips.bgen")
#input_sample <- "../results/snips.sample"   # This file does not exist
output <- file.path(result_path, "hla_snips.dosage")
cmd <- glue("qctool -g {input_bgen} -og {output} -ofiletype dosage -incl-snpids {snip_id_filename}")
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
```


Read dosages.

```{r Read dosages2}
filename <- file.path(table_path, "tidy_hla_dosages.tsv.gz")
if (! file.exists(filename)) {
  tmp <- read_delim(output, delim=" ", col_types = cols(position = col_integer(), .default = col_character()))
  tmp <- tmp %>% 
    select(-c(chromosome, rsid, position, alleleA, alleleB)) %>% 
    inner_join(hla_snip_info %>% select(SNPID, rsid), by="SNPID") %>%
    select(-SNPID) %>%
    pivot_longer(cols=-c("rsid"), names_to = "sample", values_to = "dosage") %>% 
    mutate(dosage = as.integer(dosage)) #%>%
  #inner_join(credible_snips %>% select(SNPID="locus_id", name), by="SNPID")
  #dosages_wide <- dosages %>% pivot_wider(names_from=name, values_from=dosage, id_cols = sample)
  hla_snips <- tmp %>% pivot_wider(names_from=rsid, values_from=dosage, id_cols = sample)
  write_tsv(hla_snips, filename)
} else  {
  hla_snips <- read_tsv(filename)
}
hla_snip_names <- hla_snip_info$rsid
hla_snips %>% select(-sample) %>% mutate(across(everything(), as.factor)) %>% summary()
```



### HC credible snips

```{r}
tmp <- read_csv("../data/E4_IRON_MET_R12_finemapping.csv")
tmp <- tmp %>%
  rename(pos_hg38 = pos)
tmp
```

```{r}
credible_snip_info <- tribble(
  ~SNPID, ~name, ~rsid, ~chromosome, ~pos_hg38, ~first, ~second,
  "chr6_21989753_T_C", "credible1", "rs181949568",  6, 21989753, "T", "C",
  "chr6_22583313_C_A", "credible2", "rs1284018747", 6, 22583313, "C", "A",
  "chr6_23835557_A_T", "credible3", "rs1300021441", 6, 23835557, "A", "T"
)
credible_snip_info <- credible_snip_info %>%
  right_join(tmp, by=c(SNPID="locus_id", "pos_hg38"))
credible_snip_info <- credible_snip_info %>% filter(good_cs) # Drop one bad credible set representative
filename <- file.path(data_path, "credible_snip_info.tsv")
write_tsv(credible_snip_info, filename)
credible_snip_info
credible_snip_names <- credible_snip_info$name
```



Write snip ids to a temp file.

```{r}
snip_id_filename <- "/tmp/credible_snips_ids.tsv"
if (! file.exists(snip_id_filename))
  write_tsv(credible_snip_info %>% select(SNPID), snip_id_filename, col_names = FALSE)
```

Extract the credible variants with bgenix.

```{r Extract variants3}
if (release == "r12") {
  input <- "/finngen/library-red/finngen_R12/bgen_2.0/data/chrom/finngen_R12_6.bgen"
} else {
  input <- "/finngen/library-red/finngen_R13/bgen_1.0/data/chrom/finngen_R13_6.bgen"
}
#bgen_base <- "/finngen/library-red/finngen_R12/bgen_2.0/data/chrom"
#input_bgen <- file.path(bgen_base, "finngen_R12_6.bgen")
input_sample <- paste0(input, ".sample")
output <- file.path(result_path, "credible_snips.bgen")

cmd <- glue("{bgenix} -g {input} -incl-rsids {snip_id_filename} > {output}")
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
# Create the index
cmd <- glue("{bgenix} -g {output} -index")
print(cmd)
if (!file.exists(file.path(result_path, "credible_snips.bgen.bgi"))) {
  system(cmd)
}
```

qctool took 9m16.454s
bgenix took 0m4.236s

Extract dosages.

```{r Extract dosages3}
input_bgen <- file.path(result_path, "credible_snips.bgen")
#input_sample <- "../results/snips.sample"   # This file does not exist
output <- file.path(result_path, "credible_snips.dosage")
cmd <- glue("qctool -g {input_bgen} -og {output} -ofiletype dosage -incl-snpids {snip_id_filename}")
print(cmd)
if (!file.exists(output)) {
  system(cmd)
}
```

Read dosages.

```{r Read dosages3}
filename <- file.path(table_path, "tidy_credible.dosages.tsv.gz")
if (! file.exists(filename)) {
  tmp <- read_delim(output, delim=" ", col_types = cols(position = col_integer(), .default = col_character()))
  tmp <- tmp %>% 
    select(-c(chromosome, rsid, position, alleleA, alleleB)) %>% 
    inner_join(credible_snip_info %>% select(SNPID, name), by="SNPID") %>%
    pivot_longer(cols=-c("SNPID", "name"), names_to = "sample", values_to = "dosage") %>% 
    mutate(dosage = as.integer(dosage)) #%>%
  #inner_join(credible_snips %>% select(SNPID="locus_id", name), by="SNPID")
  #tmp_wide <- tmp %>% pivot_wider(names_from=name, values_from=dosage, id_cols = sample)
  credible_snips <- tmp %>% pivot_wider(names_from=name, values_from=dosage, id_cols = sample)
  write_tsv(credible_snips, filename)
} else {
  credible_snips <- read_tsv(filename)
}
credible_snips %>% select(-sample) %>% mutate(across(everything(), as.factor)) %>% summary()
```

```{r}
#filename <- file.path(table_path, "tidy_credible.dosages.tsv.gz")
#credible_snips <- read_tsv(filename, show_col_types = FALSE)
#credible_snip_names <- c("credible1", "credible2", "credible3")
#credible_snips %>% select(-sample) %>% mutate(across(everything(), as.factor)) %>% summary()
```

### Linkage disequilirium

#### LDs using ldstore.

Describe variants.

```{r}
# rsid chromosome position allele1 allele2
# rs1 10 1 T C
# rs2 10 1 A G
# rs3 10 1 G A
zfile <- snips %>% select(rsid=SNPID, chromosome, position, allele1=first, allele2=second) %>%
  mutate(chromosome = paste0("chr", chromosome))
zfilename <- file.path(result_path, "ldstore_snips.z")
if (! file.exists(zfilename))
  write_delim(zfile, zfilename, delim = " ")
```

Describe other input.

```{r}
# z;bgen;bgi;bcor;ld;n_samples
# dataset1.z;dataset1.bgen;dataset1.bgen.bgi;dataset1.bcor;;5363
# dataset2.z;dataset2.bgen;dataset2.bgen.bgi;;dataset2.ld;5363
masterfile <- tibble(z    = zfilename, 
                     bgen = file.path(result_path, "snips.bgen"), 
                     bgi  = file.path(result_path, "snips.bgen.bgi"), 
                     bcor = file.path(result_path, "ldstore_snips.bcor"),
                     ld   = file.path(result_path, "ldstore_snips.ld"),
                     n_samples = if_else(release == "r12", 520210, 519972))
masterfilename <- file.path(result_path, "ldstore_snips.master")
if (! file.exists(masterfilename))
  write_delim(masterfile, masterfilename, delim = ";")
```

```{r}
cmd <- glue("{ldstore} --in-files {masterfilename} --write-bcor --read-only-bgen")
print(cmd)
if (! file.exists(masterfile$bcor))
  system(cmd)
cmd <- glue("{ldstore} --in-files {masterfilename} --bcor-to-text")
print(cmd)
if (! file.exists(masterfile$ld))
  system(cmd)
```

Read the LD values.

```{r}
filename <- file.path(result_path, "ldstore_snips.ld")
ld2 <- read_delim(filename, delim=" ", col_names = FALSE)  # These are Pearson correlations 
ld2 <- ld2[,1:(ncol(ld2)-1)] # Remove empty column
```



```{r}
#labels <- snips$name %>%  #paste(snips$name, snips$rsid, sep="\n") %>% 
#  factor(levels=c("H1-2", "H63D", "C282Y"))
labels <- snips$name
snip_order <- c("H1-2", "H63D", "S65C", "C282Y", "supersnip")
ld <- ld2 %>% set_names(labels) %>% 
  mutate(x=factor(labels, snip_order)) %>% 
  pivot_longer(cols=-x, names_to = "y") %>%
  mutate(y = factor(y, snip_order))
summary(ld)
```


```{r}
ld %>%
  mutate(y=fct_rev(y)) %>%
  ggplot(aes(x, y, fill=value**2)) + 
  geom_raster() +
  geom_text(aes(label=sprintf("%.2f", value**2)), color="white") +
  scale_fill_continuous(breaks=seq(0, 1, length.out = 5), limits=c(0,1)) +
  scale_x_discrete(expand=expansion(add=0)) +
  scale_y_discrete(expand=expansion(add=0)) +
  labs(x="SNP 1", y="SNP 2", title="Linkage disequilibrium using ldstore", fill=bquote(R^2))
```

These numbers are not the same as in ldlink.nih.gov with Finnish panel.

No LD between H63D and S65C even though they are only 6 bp apart!

Should I restrict the samples to hemochromatosis cases and controls?

#### LD through cor

Lets compute this using a different method.

```{r}
pearson <- my_correlation(dosages_wide %>% select(-sample), snip_order)
pearson
```

```{r}
calc_element("geom", theme_gray())  # Note the output is in millimeters, even though input is in points
```


```{r}
plot_correlation(pearson) #+ theme(geom.text = element_geom(fontsize = 11))
```

Same result as from ldstore. Ok, ldlink.nih.gov also gives R2 value 0.0006 for snips H63D and S65C using FIN population. 
But note that S65C nucleotide T appears only once!!!


### Enrichment

Use tabix to quickly read the enrichment annotation file.

```{r Extract enrichement information}
input_filename <- "/finngen/library-green/finngen_R12/finngen_R12_analysis_data/annotations/R12_annotated_variants_v0.small.gz"
#output_filename <- tempfile()
output_filename <- file.path(table_path, "enrichment.tsv")
snip_positions <- snips %>% glue_data("{chromosome}:{position}-{position}") %>% paste(collapse=" ")
cmd <- glue("tabix -h {input_filename} {snip_positions} > {output_filename}")
print(cmd)
if (!file.exists(output_filename)) {
  system(cmd)
}
```

Read enrichment information.

```{r Read enrichment information}
enrichment <- read_tsv(output_filename)
enrichment %>% 
  select("#variant", INFO, AF, AC_Het, AC_Hom, most_severe, gene_most_severe, rsid, 
         GENOME_enrichment_nfee, GENOME_enrichment_nfe) %>%
  relocate("#variant", GENOME_enrichment_nfee, GENOME_enrichment_nfe)
```

The snips don't seem to be very Finnish specific.

### Cohort info

Load cohort info.

```{r Load cohort info}
filename <- glue("/finngen/library-red/finngen_{Release}/phenotype_1.0/data/finngen_{Release}_cohort_1.0.txt.gz")
cohort <- read_tsv(filename, col_names=c("sample", "cohort"))
```

Define blood donor variable.

```{r}
bd <- cohort %>% mutate(blood_donor = cohort=="BLOOD SERVICE BIOBANK") %>% select(-cohort)
bd %>% select(blood_donor) %>% table()
```

### Imputed sex

```{r}
if (release == "r12") {
  filename <- "/finngen/library-red/finngen_R12/analysis_covariates/R12_COV_V2.FID.txt.gz"
} else {
  filename <- "/finngen/library-red/finngen_R13/analysis_covariates/data/R13_COV_V0.FID.txt.gz"
}
sex <- read_tsv(filename, col_select = c("FID", "SEX", "SEX_IMPUTED"))
sex <- sex %>% rename(sample=FID)
```

No problems with sex:

```{r}
sex %>% select(-sample) %>% table(useNA = "always")
```

```{r}
sex <- sex %>% select(-SEX_IMPUTED)
```



### Plot hemochromatosis

```{r}
helper <- function(df, par) {
  palette <- scales::hue_pal()(2)
  df <- df %>% 
    filter(hemochromatosis == par)
  breaks <- cut_helper_variable_width(df$AGE)$breaks
  g <- df %>%
    ggplot(aes(AGE)) + 
    #(if (allow_individual_level_data) geom_histogram() else geom_density(aes(y=after_stat(count)), color=NA)) + 
    #(if (allow_individual_level_data) geom_histogram() else 
      geom_histogram(breaks=breaks, color="black", 
                     fill=if_else(par=="case", palette[1], palette[2])) + 
    geom_text(data = df %>%
                count() %>%
                #count(hemochromatosis) %>% 
                mutate(x=0.2, y=0.9, label=glue("n={n}")), 
              mapping=aes(I(x), I(y), label=label),
              vjust=0, color="black", inherit.aes = FALSE
    ) +
    labs(x="Age", y="Count", title=glue("Hemochromatosis {par}s"), subtitle="Age at first event, death or EOF") #+
    #facet_wrap(~ hemochromatosis, scales = "free")
  g
}
g1 <- helper(hc, "case")
g2 <- helper(hc, "control")
min_bin_size(g1)
min_bin_size(g2)
g1+g2
```

Is the average diagnosis age different between males and females?

```{r}
tmp <- hc %>%
  inner_join(sex, by="sample") %>%
  filter(hemochromatosis == "case")
tmp %>% group_by(SEX) %>% summarise(age = mean(AGE))
male_age   <- tmp %>% filter(SEX == "male") %>% pull(AGE)
female_age <- tmp %>% filter(SEX == "female") %>% pull(AGE)
t.test(male_age, female_age)
```

```{r}
#hc %>% ggplot(aes(NEVT)) + geom_histogram() + scale_y_continuous(trans = scales::log1p_trans())+
#  labs(x="Number of events per person")
breaks <- cut_helper_variable_width(hc$NEVT)$breaks
g <- hc %>% ggplot(aes(NEVT)) + 
  geom_histogram(breaks=breaks, color="white") + 
  geom_text(aes(label=after_stat(count)), stat="bin", breaks=breaks, color="white", angle=90, hjust=1.1, vjust=0.5) +
  scale_y_continuous(trans = scales::log1p_trans()) +
  labs(x="Number of events per person", y="Number of persons in bin", 
       title="Hemochromatosis events",
       subtitle="Variable width histogram")
min_bin_size(g)
g
```

```{r}
tmp <- hc %>% drop_na(APPROX_EVENT_DAY)
breaks <- cut_helper_variable_width(tmp$APPROX_EVENT_DAY)$breaks
g <- tmp %>% 
  ggplot(aes(APPROX_EVENT_DAY)) + 
  (if (allow_individual_level_data) 
    geom_density(aes(y=after_stat(count))) else
    #geom_histogram() else 
    geom_histogram(breaks=breaks)
  ) +
  geom_vline(xintercept = as_date("1996-08-01"), color="red") +
  annotate("text", x=lubridate::as_date("1990-01-01"), y=0.09, label=glue("n={nrow(tmp)}")) +
  labs(y = if_else(allow_individual_level_data, "Density", "Count"),
       title="First hemochromatosis events", subtitle="Variable width histogram")
min_bin_size(g)
g
```

```{r}
tmp <- hc %>% drop_na(APPROX_EVENT_DAY)
g <- tmp %>% 
  mutate(rank = rank(APPROX_EVENT_DAY)) %>%
  ggplot(aes(APPROX_EVENT_DAY, rank)) + 
  geom_point() +
  geom_vline(xintercept = as_date("1996-08-01"), color="red") +
  annotate("text", x=lubridate::as_date("1990-01-01"), y=100, label=glue("n={nrow(tmp)}")) +
  labs(y = "Count",
       title="First hemochromatosis events", subtitle="Cumulative events")
g
```

### Blood donation data

```{r Read blood donation data}
filename <- if (release == "r12") "/finngen/library-red/finngen_R11/FRCBS_blood_donor_data_1.0/data/finngen_R11_FRCBS_Blood_Donor_Donations_1.0.txt.gz" else "/finngen/library-red/finngen_R13/FRCBS_blood_donor_data/data/finngen_R13_FRCBS_Blood_Donor_Donations_1.0.txt.gz"
donation <- read_tsv(filename)
donation <- donation %>% 
  mutate(across(BloodDonationTypeKey, as.factor)) %>%
  rename(sample = FINNGENID,
         status = BloodDonationTypeKey)
```

```{r}
summary(donation)
```

Number of blood donors.

```{r}
n_distinct(donation$sample)
```

```{r}
g <- donation %>% 
  filter(lubridate::as_date("1995-01-01") <= APPROX_DONATION_DATE) %>% # To get rid of long left tail
  ggplot(aes(APPROX_DONATION_DATE)) + geom_histogram(bins=30) +
  labs(title="Blood donations")
min_bin_size(g)
g
```

The area under the density curve is the number of successful Hb measurements.

```{r}
tmp <- donation %>% filter(!is.na(Hb)) 
breaks <- cut_helper_variable_width(tmp, "Hb")$breaks
g <- tmp %>% 
  ggplot(aes(Hb)) + 
  #geom_density(aes(y=after_stat(count), linetype="dotted"), key_glyph="path") + 
  (if (allow_individual_level_data) 
    geom_freqpoly(aes(linetype="solid")) else
    geom_histogram(breaks = breaks)
  ) +
  #annotate("text", x=75, y=10000, label=glue("n={nrow(tmp)}")) +
  annotate("text", x=I(0.25), y=I(0.5), label=glue("n={nrow(tmp)}")) +
  labs(y=if_else(allow_individual_level_data, "Density", "Count"),
       title="Hemoglobin measurements", subtitle="Variable width histogram") +
  scale_linetype_identity(name = "Curve type",
                          breaks = c("solid", "dotted"),
                          labels = c("freqpoly", "density"),
                          guide = "legend")
min_bin_size(g)
g
```

### Save extracted datas

```{r}
filename <- file.path(result_path, "all_datas.rds")
if (! file.exists(filename) || recompute) {
  L <- list(dosages=dosages, cohort=cohort, bd=bd, sex=sex, hc=hc, donation=donation)
  saveRDS(L, filename)
}
```

```{r, eval=FALSE}
L <- readRDS(file.path(result_path, "all_datas.rds"))
dosages <- L$dosages
cohort <- L$cohort
bd <- L$bd
sex <- L$sex
hc <- L$hc
donation <- L$donation
```

## Analyze data

Combine datasets.

```{r}
hc_full <- hc %>%
  select(sample, hemochromatosis, age=AGE, approx_event_date = APPROX_EVENT_DAY) %>%
  inner_join(bd, by="sample") %>%
  inner_join(sex %>% rename(sex=SEX), by="sample") %>%
  mutate(sex = as.factor(sex))
filename <- file.path(result_path, "hemochromatosis_full.rds")
if (! file.exists(filename) || recompute) {
  saveRDS(hc_full, file=filename)
}
```


```{r}
hc_full %>% count(hemochromatosis) %>% mutate(freq = n / sum(n))
```

### Prevalences

```{r}
hc_full %>% 
  count(sex) %>%
  mutate(freq = n / sum(n),
         label = sprintf("n=%i\n%.1f%%", n, freq*100)) %>%
  ggplot(aes(sex, y=n, fill=sex)) + 
  geom_col() + 
  geom_text(aes(y=n/2, label=label))
```

```{r}
part1 <- hc_full %>% 
  mutate(blood_donor = if_else(blood_donor, "Blood donor", "Non-donor")) %>%
  count(blood_donor, hemochromatosis, sex)
part2 <-  hc_full %>%
  mutate(blood_donor = "all") %>%
  count(blood_donor, hemochromatosis, sex)
tmp <- bind_rows(part1, part2) %>%
  pivot_wider(names_from=sex, values_from=n) %>%
  mutate(all = female + male) %>%
  pivot_longer(cols=c(female, male, all), names_to = "sex", values_to = "n") %>%
  group_by(sex, blood_donor) %>%
  mutate(total=sum(n), freq = n/total) %>%
  ungroup() %>%
  mutate(label = sprintf("%i\n%.2f%%", n, freq*100))
if (save_tables) {
  # I should save also the denominator
  filename <- file.path(table_path, "fg_prevalences.tsv")
  write_tsv(tmp %>% 
              filter(hemochromatosis == "case") %>% 
              select(-hemochromatosis) %>% 
              mutate(label = stringi::stri_escape_unicode(label)), # escape the newline characters
            filename)
}
tmp %>%
  ggplot(aes(sex, n, fill=sex)) + geom_col() + geom_text(aes(label=label)) +
  scale_y_continuous(expand=expansion(mult=.15)) +
  labs(y="Count", title="Case and control proportions in subgroups") +
  facet_grid(hemochromatosis ~ blood_donor, scales = "free")
```

Test for equality of hemochromatosis prevalences in blood donors vs non-blood donors.

```{r}
donor_row <- tmp %>% filter(blood_donor =="Blood donor", sex=="all")
non_donor_row <- tmp %>% filter(blood_donor =="Non-donor", sex=="all")
counts <- tmp %>% filter(blood_donor %in% c("Blood donor", "Non-donor"), sex=="all")
res <- prop.test(x = counts$n, n = counts$total)
res
```

```{r}
tidy(res)
```

The prevalences are significantly different.

```{r}
df <- dosages %>%
  inner_join(bd, by="sample") %>%
  inner_join(hc %>% select(sample, hemochromatosis), by="sample")
```


```{r}
df_wide <- df %>% 
  select(name, dosage, sample, blood_donor, hemochromatosis) %>% 
  pivot_wider(names_from=name, values_from=dosage)
```

```{r}
all_snips_wide <- hc_full %>%
  inner_join(dosages_wide, by="sample") %>% 
  inner_join(hla_snips, by="sample") %>% 
  inner_join(credible_snips, by="sample")
all_snips <- all_snips_wide %>%
  pivot_longer(cols=-c(sample, blood_donor, hemochromatosis, age, approx_event_date, sex),
               names_to="name", values_to="dosage")
```

### Linkage disequilibrium


```{r}
hla_order <- c("F*01:01", "F*01:03", "G*UTR-4", "G*01:01:01:05",  "G*010101c", "A*03:01", "C*07:02", "B*07:02", snip_order, 
               hla_snip_names, credible_snip_names)
```

#### All

```{r}
ld_input_wide <- all_snips_wide %>% 
  inner_join(hla_dosages_wide, by="sample")
hla_pearson <- ld_input_wide %>% 
  select(-c(sample, hemochromatosis, age, approx_event_date, blood_donor, sex)) %>% 
  drop_na() %>% my_correlation(hla_order)
hla_pearson %>% head()
if (save_tables) {
  filename <- file.path(table_path, "ld_among_all.tsv")
  write_tsv(hla_pearson, filename)
}
g <- plot_correlation(hla_pearson) + theme(geom.text = element_geom(fontsize = 3), axis.text.x = element_text(angle = 45, hjust=1))
if (save_figs) {
  filename <- file.path(figure_path, "ld_among_all.pdf")
  ggsave(filename, g, width=8, height=8)
}
g
```

Should this be done to haplotyped data instead?

#### Cases

```{r}
hla_pearson_cases <- ld_input_wide %>% 
  filter(hemochromatosis == "case") %>%
  select(-c(sample, hemochromatosis, age, approx_event_date, blood_donor, sex)) %>% 
  drop_na() %>% my_correlation(hla_order)
hla_pearson_cases %>% head()
if (save_tables) {
  filename <- file.path(table_path, "ld_among_cases.tsv")
  write_tsv(hla_pearson_cases, filename)
}
g <- plot_correlation(hla_pearson_cases) + 
  labs(subtitle="Among cases") + 
  theme(geom.text = element_geom(fontsize = 3), axis.text.x = element_text(angle = 45, hjust=1))
if (save_figs) {
  filename <- file.path(figure_path, "ld_among_cases.pdf")
  ggsave(filename, g, width=8, height=8)
}
g
```

#### Controls

```{r}
hla_pearson_controls <- ld_input_wide %>% 
  filter(hemochromatosis == "control") %>%
  select(-c(sample, hemochromatosis, age, approx_event_date, blood_donor, sex)) %>% drop_na() %>% my_correlation(hla_order)
hla_pearson_controls %>% head()
if (save_tables) {
  filename <- file.path(table_path, "ld_among_controls.tsv")
  write_tsv(hla_pearson_controls, filename)
}
g <- plot_correlation(hla_pearson_controls) + 
  labs(subtitle="Among controls") + 
  theme(geom.text = element_geom(fontsize = 3), axis.text.x = element_text(angle = 45, hjust=1))
if (save_figs) {
  filename <- file.path(figure_path, "ld_among_controls.pdf")
  ggsave(filename, g, width=8, height=8)
}
g
```


### Minor allele frequencies


```{r}
compute_mafs <- function(df, variables) {
  df %>% 
    group_by(across(all_of(variables))) %>% 
    summarise(minor_allele_count = sum(dosage), allele_count = 2*n(), n = n(), .groups = "drop") %>%
    mutate(minor_allele_freq = minor_allele_count / allele_count)
}
```

```{r Minor allele frequencies}
mafs <- compute_mafs(all_snips, "name")
if (save_tables) {
  filename <- file.path(table_path, "fg_mafs.tsv")
  write_tsv(mafs, filename)
}
mafs
```

Stratify by hc status to enable comparison to UKBB data.

```{r}
group_by_variables2 <- c("name", "hemochromatosis")
mafs_by_status <- compute_mafs(all_snips, group_by_variables2)
if (save_tables) {
  filename <- file.path(table_path, "fg_mafs_by_status.tsv")
  write_tsv(mafs_by_status, filename)
}
mafs_by_status
```


```{r}
mafs_by_status %>%
  ggplot(aes(name, minor_allele_freq, fill=hemochromatosis)) + 
  geom_col(position="dodge") +
  labs(y="Minor allele frequency") + 
  theme(axis.text.x = element_text(angle = 45))
```

Stratify by blood_donor status and hc status.


```{r Minor allele frequencies in subgroups}
group_by_variables <- c("name", "blood_donor", "hemochromatosis")
mafs_by_status_and_bd <- compute_mafs(all_snips, group_by_variables)
if (save_tables) {
  filename <- file.path(table_path, "fg_mafs_by_status_and_bd.tsv")
  write_tsv(mafs_by_status_and_bd, filename)
}
if (allow_individual_level_data)
  mafs_by_status_and_bd
```

The minor allele frequencies are higher among blood donors, as noted in Jonna's paper.

Test the difference in MAFs between bd and non-bd in cases.

```{r}
counts <- mafs_by_status_and_bd %>% filter(name == "C282Y", hemochromatosis == "case")
counts %>% { prop.test(x = .$minor_allele_count, n = .$allele_count) }
```

Test the difference, ignore hemochromatosis status.

```{r}
counts <- mafs_by_status_and_bd %>% filter(name == "C282Y") %>% 
  group_by(blood_donor) %>% summarise(across(c(minor_allele_count, allele_count, n), sum))
counts %>% { prop.test(x = .$minor_allele_count, n = .$allele_count) }
```

```{r}
if (allow_individual_level_data) {
  mafs_by_status_and_bd %>%
    mutate(blood_donor = if_else(blood_donor, "donor", "non-donor")) %>%
    ggplot(aes(name, minor_allele_freq, fill=hemochromatosis)) + 
    geom_col(position="dodge") +
    facet_grid(. ~ blood_donor) +
    labs(y="Minor allele frequency") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1))
}
```

### Dosage distributions

```{r}
compute_dosage_freqs <- function(df, variables) {
  freqs <- df %>% 
    group_by(across(all_of(variables))) %>%
    count(dosage, name = "observed_count") %>%
    ungroup() %>%
    add_missing_combinations2("dosage", !!!variables) %>%   # Make implicit NAs explicit
    replace_na(list(observed_count = 0)) %>%
    group_by(across(all_of(variables))) %>%
    mutate(total = sum(observed_count), observed_freq = observed_count / total) %>%
    ungroup()
  freqs
}

compute_expected_dosage_freqs <- function(maf) tibble(dosage=0:2, expected_freq=c((1-maf)**2, 2*(1-maf)*maf, maf**2))
```

```{r}
expected_dosages <- mafs %>% mutate(value = map(minor_allele_freq, compute_expected_dosage_freqs)) %>% unnest(value) %>%
  mutate(expected_count = expected_freq*n)
all_snips %>% 
  ggplot(aes(dosage)) + 
  geom_bar() +
  geom_point(data=expected_dosages, mapping=aes(x=dosage, y=expected_count), color="red") +
  geom_text(aes(label=after_stat(count)), stat="count", vjust=-0.1) +
  labs(subtitle="Red point is expected count due to MAF") +
  facet_wrap(. ~ name)
```

Dosage counts and frequencies.

```{r Dosage frequencies}
compute_dosage_freqs(all_snips, "name")
```



Don't stratify by blood donor status in the downloadable version of the rendered report.






```{r}
#allow_individual_level_data2 <- allow_individual_level_data
allow_individual_level_data2 <- FALSE   # Exception local to this chunk
group_by_variables <- c("name", "blood_donor", "hemochromatosis")
if (! allow_individual_level_data2)
  group_by_variables <- setdiff(group_by_variables, "blood_donor")

expected_dosages <- compute_mafs(all_snips, group_by_variables) %>% 
  select(all_of(group_by_variables), minor_allele_freq) %>%
  mutate(value = map(minor_allele_freq, compute_expected_dosage_freqs)) %>% 
  unnest(value)
compute_hwe <- function(v) {  # v should be a vector of integers of length 3
  res <- gap::hwe(v, data.type = "count")
  res$p.x2
}
observed_dosages <- compute_dosage_freqs(all_snips, group_by_variables)
hwes <- observed_dosages %>% group_by(across(all_of(group_by_variables))) %>% 
  summarise(hwe = compute_hwe(observed_count)) %>% ungroup() %>%
  mutate(label = sprintf("hwe=%.1e", hwe),
         color = if_else(hwe < 0.05, "red", "black"))

# Save to joined dataframe
filename <- file.path(table_path, "fg_dosage_maf.tsv")
joined <- expected_dosages %>%
  #rename(expected_freq = freq) %>%
  left_join(observed_dosages, by=c("name", "hemochromatosis", "dosage")) %>% 
  inner_join(hwes, by=c("name", "hemochromatosis")) %>%
  filter(hemochromatosis == "case")
if (save_tables)
  write_tsv(joined, filename)
#rm(joined)

observed_dosages %>%
  ggplot(aes(dosage, observed_freq, fill=(if (allow_individual_level_data2) blood_donor))) + 
  geom_col(position="dodge") + 
  geom_point(data = expected_dosages, aes(dosage, expected_freq), color="red", position=position_dodge(width=1)) + 
  geom_text(aes(label=observed_count), vjust=0.5, hjust=0, position=position_dodge(width=1), angle=90) +
  geom_text(data=hwes, mapping=aes(x=I(0.6), y=I(0.95), label = label, color=I(color))) +
  scale_y_continuous(labels=scales::label_percent(), limits=c(0, 1), name="Percentage", expand = expansion(mult=c(0.05, 0.3))) +
  labs(x="Dosage", fill="Is blood donor", subtitle="Red point is expected frequency due to MAF") + 
  #lims(y=c(0, 100)) +
  facet_grid(hemochromatosis ~ name)
```

Check that count are ok.

```{r}
joined %>% filter(observed_count < 5)
```


### HFE mutations among cases

#### as a function in time

```{r}
equal_count_breaks <- function(v, n) {
  v <- as.numeric(v)
  ggplot2:::breaks(v, "numbers", n) %>% as_date()
}
equal_count_cut <- function(v, n) { 
  brk <- equal_count_breaks(v, n)
  cut(v, brk, include.lowest = TRUE)
}
mutations_in_time <- hc %>% 
  filter(hemochromatosis == "case")  %>%
  left_join(dosages_wide %>% select(sample, H63D, C282Y), by="sample") %>%
  #mutate(.bin = cut(APPROX_EVENT_DAY, breaks = 10) %>% fct_relabel(sort)) %>%  # for some reason this did not give equal width bins?!
  #mutate(.bin = cut_interval(APPROX_EVENT_DAY, n = 10)) %>%
  mutate(.bin = equal_count_cut(APPROX_EVENT_DAY, n = 10)) %>%
  mutate(C282Y_H63D_variation = (H63D + C282Y) > 0,
    C282Y_variation = C282Y > 0,
    C282Y_homozygote = C282Y == 2,
    .by = .bin
  )
  # mutate(.bin = ntile(APPROX_EVENT_DAY, n = 10)) %>%
  # group_by(.bin) %>%
  # mutate(min_date = min(APPROX_EVENT_DAY), 
  #   max_date = max(APPROX_EVENT_DAY),
  #   .label = sprintf("%s - %s", min_date, max_date)) %>%
  # ungroup()
```

```{r}
my_levels <- c("C282Y_H63D_variation", "C282Y_variation", "C282Y_homozygote")
mutation_rate <- function(df) {
  df %>%
    pivot_longer(cols = all_of(my_levels), values_to="value", names_to="term") %>%
    summarise(estimate = mean(value), n = n(), .by = c(term, .bin))# %>%
    #mutate(term = "mutation_rate")
}
```

```{r}
mutation_rate(mutations_in_time) %>% arrange(.bin)
```

```{r}
set.seed(1)
time_bootstrap <- rsample::bootstraps(mutations_in_time, times = 1000)

time_stats <-
  time_bootstrap %>%
  mutate(stats = purrr::map(splits, ~ analysis(.x) %>% mutation_rate()))
```

```{r}
time_ci2 <- rsample::int_pctl(time_stats, statistics = stats, alpha = 0.05)
time_ci2
```

```{r}
#| fig-width: 10
#| fig-height: 7

annotations <- mutations_in_time %>% count(.bin) %>% 
  mutate(term = factor("C282Y_H63D_variation", my_levels),
    label = glue("n={n}"))
time_ci <- time_ci2 %>%
  left_join(annotations, by=c(".bin", "term"))
if (save_tables) {
  filename <- file.path(table_path, "variations_along_time.tsv")
  write_tsv(time_ci, filename)
}
g <- time_ci %>%
  #mutate(.bin = fct_rev(.bin)) %>%
  mutate(term = fct_relevel(term, my_levels)) %>%
  ggplot(aes(x = .bin, y = .estimate, ymin = .lower, ymax = .upper)) +
  geom_hline(data=tibble(term = factor("C282Y_homozygote", my_levels)), mapping = aes(yintercept = 0.25), color = "red", ) + 
  geom_pointrange() +
  geom_line(aes(group = term)) +
  geom_text(#data = annotations, 
    mapping = aes(x=.bin, y = I(0.25), label = label), inherit.aes = FALSE) +
  lims(y=0:1) + 
  labs(y = "Variation frequency", x="Date", title = "Variation frequencies in C282Y and H63D SNPs among cases in time",
    subtitle = "Equal-sized, variable-width bins",
    caption = "Relative frequency of individuals among cases with the described variation.") +
  facet_grid(term ~ ., labeller = as_labeller(c(C282Y_homozygote = "C282Y homozygote", 
                                                C282Y_variation = "Any C282Y variant",
                                                C282Y_H63D_variation = "Any C282Y or H63D\nvariant")))
if (save_figs) {
  filename <- file.path(figure_path, "variations_along_time.pdf")
  ggsave(filename, g, width = 10, height = 5, units = "in")
}
g
```

There is not much variation in the mutation fraction.

#### as a function of biobank

```{r}
mutations_in_place <- hc %>% 
  filter(hemochromatosis == "case")  %>%
  left_join(dosages_wide %>% select(sample, H63D, C282Y), by="sample") %>%
  left_join(cohort) %>% 
  mutate(.bin = fct_lump_min(cohort, min=10) %>% fct_relabel(sort)) %>%
  #mutate(.bin = equal_count_cut(APPROX_EVENT_DAY, n = 10)) %>%
  mutate(C282Y_H63D_variation = (H63D + C282Y) > 0,
    C282Y_variation = C282Y > 0,
    C282Y_homozygote = C282Y == 2,
    .by = .bin
  )
```

```{r}
mutation_rate(mutations_in_place) %>% arrange(.bin)
```

```{r}
set.seed(1)
place_bootstrap <- bootstraps(mutations_in_place, times = 1000)

place_stats <-
  place_bootstrap %>%
  mutate(stats = purrr::map(splits, ~ analysis(.x) %>% mutation_rate()))
```

```{r}
place_ci <- int_pctl(place_stats, statistics = stats, alpha = 0.05)
place_ci
```

```{r}
#| fig-width: 10
#| fig-height: 5
annotations <- mutations_in_place %>% count(.bin) %>% 
  mutate(term = factor("C282Y_H63D_variation", my_levels),
    label = glue("n={n}"))
place_ci %>%
  #mutate(.bin = fct_rev(.bin)) %>%
  mutate(term = fct_relevel(term, my_levels)) %>%
  ggplot(aes(x = .bin, y = .estimate, ymin = .lower, ymax = .upper)) +
  geom_pointrange() +
  geom_line(aes(group = term)) +
  geom_text(data = annotations, 
    mapping = aes(x=.bin, y = I(0.25), label = label), inherit.aes = FALSE) +
  lims(y=0:1) + labs(y = "Variation rate", x="Biobank") +
  labs(y = "Variation rate", x="Subcohort", title = "Variation rates of C282Y and H63D among cases in place") +
    #subtitle = "Equal-sized bins") +
  facet_grid(term ~ .) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

### Snip combinations

Cross table. I will not stratify by blood donor status as well since that would result in too small groups.

```{r}
snip_combinations <- function(df, snip1, snip2, filename=NULL) {
  snip_names <- c(snip1, snip2)
  wide <- df %>% 
    filter(name %in% snip_names) %>% 
    select(sample, name, dosage, hemochromatosis) %>% 
    pivot_wider(names_from = name, values_from=dosage)
  #print(wide)
  correlations <- wide %>% group_by(hemochromatosis) %>% summarise(correlation = cor(.data[[snip1]], .data[[snip2]]))
  correlations <- correlations %>% 
    mutate(s = sprintf("%s: r=%.3f r^2=%.3f", hemochromatosis, correlation, correlation**2)) %>% 
    pull(s) %>% 
    paste(collapse="\n")
  tmp <- wide %>%
    mutate(across(-sample, as.factor)) %>%
    count(.data[[snip1]], .data[[snip2]], hemochromatosis) %>%
    group_by(hemochromatosis) %>%
    mutate(total = sum(n),
           freq = n / total,
           label = sprintf("n=%i\n%.1f%%", n, freq*100)) %>%
    ungroup()# %>%
    #print()
  #         label = glue("n={n}\n{format(freq*100, digits=1, scientific=FALSE, nsmall=1)}%"))
  if (! is.null(filename) && save_tables)
    write_tsv(tmp %>% mutate(label=stringi::stri_escape_unicode(label)), filename) # escape the newline characters
  min_count <- min(tmp$n)
  if (allow_individual_level_data || min_count >= 5) {
  tmp %>%
    ggplot(aes(x=.data[[snip1]], y=.data[[snip2]], fill=freq)) + 
    geom_raster() + geom_text(aes(label=label), color="white") +
    scale_fill_continuous(labels=scales::label_percent(), name="Frequency") +
    labs(title="Dosage combinations", subtitle=correlations) +
    facet_grid(~ hemochromatosis)
  } else {
    "Figure not shown as too low counts appear in it!"
  }
}
```

```{r}
filename <- file.path(table_path, "fg_combinations_C282Y_H63D.tsv")
snip_combinations(all_snips, "C282Y", "H63D", filename=filename)
```

```{r}
snip_combinations(all_snips, "C282Y", "H1-2")
```

```{r}
snip_combinations(all_snips, "H63D", "H1-2")
```

Non-C282Y snip carriers versus non-carries and cases and controls

```{r}
tmp <- df_wide %>% 
  rowwise() %>% 
  mutate(other = sum(`H1-2`, H63D, S65C, supersnip)) %>% 
  ungroup() %>% 
  mutate(C282Y = as.integer(C282Y > 0), other = as.integer(other > 0))
tmp <- tmp %>% 
  select(hemochromatosis, C282Y, other) %>%
  mutate(hemochromatosis = as.integer(hemochromatosis=="case"))
```

```{r}
tmp %>%
  count(hemochromatosis, C282Y, other) %>%
  group_by(C282Y) %>%
  mutate(freq = n / sum(n),
         label = sprintf("n=%i\n%.1f%%", n, freq*100)) %>%
  ungroup() %>%
  ggplot(aes(x=hemochromatosis, y=other, fill=freq)) + 
  geom_raster() + 
  geom_text(aes(label=label), color="white") +
  scale_x_continuous(breaks=0:1, labels=c("control", "case")) +
  scale_y_continuous(breaks=0:1) +
  scale_fill_continuous(labels=scales::label_percent(), name="Frequency") +
  labs(y="Carrier of any other snip") +
  #labs(title="Dosage combinations", subtitle=correlations) +
  facet_grid(~ C282Y, labeller = as_labeller(c(`0`="Non-C282Y-carrier", `1`="C282Y-carrier")))
```

### Penetrance

Penetrance by individual variant.

```{r}
all_snips %>% group_by(name, dosage) %>% summarise(penetrance = mean(hemochromatosis=="case")) %>% ungroup()
```

Penetrance by variant pairs.

```{r}
penetrance <- function(df_wide, snp1, snp2) {
  df_wide %>% 
    group_by(.data[[snp1]], .data[[snp2]]) %>% 
    summarise(penetrance = mean(hemochromatosis=="case"), .groups="drop") %>%
    mutate(label = sprintf("%.1f%%", penetrance*100)) %>%
    ggplot(aes(x=.data[[snp1]], y=.data[[snp2]], fill=penetrance)) +
    geom_raster() +
    geom_text(aes(label=label), color="white") +
    scale_fill_continuous(labels=scales::label_percent(), name="Penetrance") +
    labs(title="Penetrance")
}
```

```{r}
penetrance(all_snips_wide, "C282Y", "H63D")
```

```{r}
penetrance(all_snips_wide, "C282Y", "H1-2")
```

```{r}
penetrance(all_snips_wide, "H1-2", "H63D")
```

```{r}
tibble(a=2:5) %>% rowwise() %>% mutate(b=list(1:(a-1))) %>% unnest(cols=b) %>%
  pwalk(function(a, b) plot(penetrance(all_snips_wide, snip_names[a], snip_names[b])))
```

```{r}
df_hla_wide <- df_wide %>% inner_join(hla_dosages_wide)
```


```{r}
#penetrance(df_hla_wide, "C282Y", "F*01:01")
penetrance(ld_input_wide, "C282Y", "F*01:01")
```

```{r}
walk(hla_alleles %>% pull(allele), function(allele) plot(penetrance(ld_input_wide, "C282Y", allele)))
```


### Blood donation

```{r}
sdonation <- donation %>% 
  group_by(sample) %>%
  summarise(count = sum(status=="Whole_blood_donation"),
            mean_Hb = mean(Hb, na.rm=TRUE)) %>% inner_join(sex, by="sample")
```

```{r}
tmp <- sdonation %>% 
  pivot_longer(cols=c(count, mean_Hb)) %>%
  inner_join(hc, by="sample") %>%
  filter(!(name=="mean_Hb" & value < 100)) # drop outlier
breaks <- c("dashed", "solid")
labels <- c("Density", "Frequency polygon")
if (! allow_individual_level_data) {
  breaks <- "dashed"
  labels <- "Density"
}
counts <- tmp %>% count(SEX, hemochromatosis) %>% mutate(label = glue("n={n}")) # counts in facets
tmp %>% ggplot(aes(value, color=SEX)) + 
  (if (allow_individual_level_data) geom_freqpoly(aes(linetype="solid"), key_glyph="path")) +
  geom_density(aes(y=after_stat(count), linetype="dashed"), key_glyph="path") + #c(color=TRUE, linetype=TRUE, fill=FALSE)) +
  geom_text(data=counts %>% filter(SEX=="female"), mapping=aes(label=label), x = -Inf, y = Inf, hjust = 0, vjust = 2) +
  geom_text(data=counts %>% filter(SEX!="female"), mapping=aes(label=label), x = -Inf, y = Inf, hjust = 0, vjust = 1) +
  #annotate(geom = 'text', label = 'sometext', x = -Inf, y = Inf, hjust = 0, vjust = 1) +
  scale_linetype_identity(#values = c("dashed", "solid"),
                        breaks = breaks,
                        labels = labels, 
                        name="Line",
                        #guide="legend"
                        guide = guide_legend(override.aes = list(linetype = breaks))) +
  labs(color="Sex", y=if_else(allow_individual_level_data, "count", "density")) +
  facet_grid(hemochromatosis ~ name, scales = "free")
```



### Diagnoses before and after hemochromatosis

These endpoints are from FinnGen R12 E4_IRON_MET CodeWAS analysis.

```{r}
# codewas <- c("E83.1", "TPH00", "SPAT1074", "M14.5*E83.1", "B99", "K76.0", "JJ2AT", "JN1AE", "TPW99", "K76.9", "K74.6", "E78.5", "FM1EE", "TPH00", "B04", "R74.0", "2750A", "GD1AA", "E87.1", "I85.9", "M14.5", "M16.1", "JN3AE", "B28", "G47.3", "UJD10", "K73.80", "G03BA03", "TPH04", "D75.1", "TJJ00", "R53", "JJ1LE", "XX1XT", "C22.0", "D83.1", "V03AC03", "E83.0", "C83.1", "JN1DG")

```

Read phenocodes and prettier phenotype names.

```{r}
# This contains only core endpoints:
#filename <- "/finngen/library-green/finngen_R12/finngen_R12_analysis_data/finngen_R12_pheno_n.tsv"
#phenotypes <- read_tsv(filename, col_select = c("phenotype", "phenocode", "category"))
# This contains all endpoints, but no categories
filename <- glue("/finngen/library-red/finngen_{Release}/phenotype_1.0/documentation/finngen_{Release}_endpoint_definitions_1.0.txt")
phenotypes <- read_tsv(filename, col_select = c("NAME", "LONGNAME")) %>%
  rename(phenocode = NAME, phenotype = LONGNAME)
#phenotypes %>% head()
```

Read the medical codes and their descriptions.

```{r}
filename <- "/finngen/library-green/finngen_R9/finngen_R9_medical_codes/finngen_R9_medcode_ref.tsv"
medical_codes <- read_tsv(filename)
```

```{r}
tmp <- medical_codes %>% group_by(code) %>% filter(n() > 1) %>% ungroup()
tmp %>% select(code_set) %>% table()
```

Read the drug codes.

```{r}
filename <- "/finngen/library-green/finngen_R6/finngen_R6_medical_codes/ATC_translate_minimal.txt"
drug_names <- read_tsv(filename)
```

```{r}
all_codes <- bind_rows(
  medical_codes %>% select(code_set, CODE1=code, description=name_en) %>% distinct(),
  # The observed code don't contain . in the ICD10 code, hence we drop it.
  medical_codes %>% select(code_set, CODE1=code, description=name_en) %>% 
    mutate(CODE1 = if_else(code_set=="ICD10", str_remove(CODE1, "\\."), CODE1)) %>% distinct(),
  drug_names %>% select(CODE1=Id, description=Longname) %>% mutate(code_set = "ATC")
)  %>% distinct()
```

Load most typical medical codes for hemochromatosis.

```{r}
filename <- file.path(data_path, "codewas_hemochromatosis_r11.tsv")
best_codes <- read_tsv(filename)
codewas <- best_codes$Code
```

```{r}
cases <- hc %>% filter(hemochromatosis =="case")
case_ids <- cases$sample
```

Check that the longitudinal endpoint file is sorted by the finngenids, with the exception of the first line.

```{bash, eval=FALSE}
dir=/finngen/library-red/finngen_R12/phenotype_1.0/data
zcat $dir/finngen_R12_endpoint_longitudinal_1.0.txt.gz | cut -f 1 | tail -n+2 | uniq | sort -c
```

Check that the detailed longitudinal file is sorted, with the exception of the first line.

```{bash, eval=FALSE}
dir=/finngen/library-red/finngen_R12/phenotype_1.0/data
zcat $dir/finngen_R12_detailed_longitudinal_1.0.txt.gz | cut -f 1 | tail -n+2 | uniq | sort -c
```

If nothing is printed, then the file is sorted according to the first field.

Create helper file with all FINNGEN ids in alphabetical order we are interested in.

```{r}
df2 <- tibble(FINNGENID=sort(case_ids))
write_tsv(df2, "/tmp/keys.tsv")
```

Join the two files according the first field. This does not work unless the keys are sorted.

```{bash}
. ../data/bash_init  # Read some bash variables
if [[ $release == "r12" ]] ; then 
  input=/finngen/library-red/finngen_R12/phenotype_1.0/data/finngen_R12_endpoint_longitudinal_1.0.txt.gz
else
  input=/finngen/library-red/finngen_R13/phenotype_1.0/data/finngen_R13_endpoint_longitudinal_1.0.txt.gz
fi
output=$result_path/diagnoses_of_hc_cases.tsv.gz
if [[ ! -f $output ]] ; then
  zcat $input | join -t $'\t' -j 1 /tmp/keys.tsv - | gzip > $output
fi
```

Join the two files according the first field. Now with detailed longitudinal data.

```{bash}
. ../data/bash_init  # Read some bash variables
if [[ $release == "r12" ]] ; then 
  input=/finngen/library-red/finngen_R12/phenotype_1.0/data/finngen_R12_detailed_longitudinal_1.0.txt.gz
else
  input=/finngen/library-red/finngen_R13/phenotype_1.0/data/finngen_R13_detailed_longitudinal_1.0.txt.gz
fi
output=$result_path/detailed_diagnoses_of_hc_cases.tsv.gz
if [[ ! -f $output ]] ; then
  zcat $input | join -t $'\t' -j 1 /tmp/keys.tsv - | gzip > $output
fi
```



Read the diagnoses for the individuals with hemochromatosis cases.

```{r}
filename <- file.path(result_path, "diagnoses_of_hc_cases.tsv.gz")
other_diagnoses <- read_tsv(filename)
other_diagnoses <- other_diagnoses %>% 
  rename(sample=FINNGENID) %>% 
  inner_join(cases %>% select(sample, first_hc_date=APPROX_EVENT_DAY), by="sample") %>%
  left_join(phenotypes, by=c(ENDPOINT="phenocode"))
```

Read the medical codes for the individuals with hemochromatosis cases.
These contain other codes in addition to diagnoses.

```{r}
filename <- file.path(result_path, "detailed_diagnoses_of_hc_cases.tsv.gz")
other_codes_orig <- read_tsv(filename)  %>% 
  rename(sample=FINNGENID)
other_codes <- other_codes_orig %>% 
  inner_join(cases %>% select(sample, first_hc_date=APPROX_EVENT_DAY), by="sample") %>%
  mutate(type = str_remove(CATEGORY, "[0-9]+$"),
         code_set = case_when(SOURCE == "PURCH" ~ "ATC",
                              type == "" | type == "ICD"   ~ glue("ICD{ICDVER}"),
                              TRUE ~ type))
```

```{r}
other_codes_orig %>% filter(is.na(CATEGORY)) %>% select(SOURCE, ICDVER) %>% table()
```

```{r}
diagnoses_before <- other_diagnoses %>% 
  filter(APPROX_EVENT_DAY < first_hc_date)
diagnoses_after <- other_diagnoses %>% 
  filter(APPROX_EVENT_DAY > first_hc_date)
diagnoses_before_ever <- diagnoses_before %>% 
  select(sample, ENDPOINT, phenotype) %>%
  distinct()
diagnoses_after_ever <- diagnoses_after %>% 
  select(sample, ENDPOINT, phenotype) %>%
  distinct()
```

```{r}
codes_before <- other_codes %>% 
  filter(APPROX_EVENT_DAY < first_hc_date)
codes_after <- other_codes %>% 
  filter(APPROX_EVENT_DAY > first_hc_date)
codes_before_ever <- codes_before %>% 
  select(sample, code_set, CODE1) %>%
  distinct()
codes_after_ever <- codes_after %>% 
  select(sample, code_set, CODE1) %>%
  distinct()
```



These functions find for each individual and diagnosis or medical code the min, median and max
distance to the hc diagnosis.

```{r}
helper_diagnosis <- function(df) {
  sign <- if_else(df$APPROX_EVENT_DAY < df$first_hc_date, +1, -1)
  df %>% mutate(distance_in_years = sign*interval(APPROX_EVENT_DAY, first_hc_date) / dyears(1)) %>%
    distinct() %>%
    group_by(sample, ENDPOINT, phenotype) %>%
    summarise(min = min(distance_in_years), max=max(distance_in_years), median=median(distance_in_years),
              .groups="drop") %>%
    ungroup()
}
helper_code <- function(df) {
  sign <- if_else(df$APPROX_EVENT_DAY < df$first_hc_date, +1, -1)
  df %>% mutate(distance_in_years = sign*interval(APPROX_EVENT_DAY, first_hc_date) / dyears(1)) %>%
    distinct() %>%
    group_by(sample, CODE1) %>%
    summarise(min = min(distance_in_years), max=max(distance_in_years), median=median(distance_in_years),
              .groups="drop") %>%
    ungroup()
}
```

```{r}
limit <- if_else(allow_individual_level_data, 0, 5)
```

#### Before

How many individuals have had a diagnosis before the first hc diagnosis?

```{r}
prior_diagnoses_counts <- diagnoses_before_ever %>% 
  count(ENDPOINT, phenotype) %>% 
  mutate(ENDPOINT=fct_reorder(ENDPOINT, n, .desc = TRUE),
         phenotype=fct_reorder(phenotype, n, .desc = TRUE)) %>% 
  arrange(desc(n)) %>% 
  filter(n >= limit)# %>% head()
filename <- file.path(table_path, "prior_diagnoses_counts.tsv")
if (save_tables)
  write_tsv(prior_diagnoses_counts, filename)
most_common_prior_diagnoses <- prior_diagnoses_counts %>% slice_head(n=20)
most_common_prior_diagnoses %>% select(phenotype, n)
```

How many individuals have had a medical code before the first hc diagnosis?

```{r}
prior_codes_counts <- codes_before_ever %>% 
  count(CODE1, code_set) %>% 
  left_join(all_codes %>% select(code_set, CODE1, description), by=c("CODE1", "code_set")) %>%
  mutate(ENDPOINT=fct_reorder(CODE1, n, .desc = TRUE),
         description=fct_reorder(description, n, .desc = TRUE)) %>% 
  arrange(desc(n)) %>% 
  filter(n >= limit) %>% # %>% head()
  left_join(best_codes %>% drop_na(Description) %>% mutate(Code=str_remove(Code, "\\.")), by=c(CODE1="Code")) %>%
  select(code_set, CODE1, description, n, "Odds ratio", mlog10p)
filename <- file.path(table_path, "prior_codes_counts.tsv")
if (save_tables)
  write_tsv(prior_codes_counts, filename)
most_common_prior_codes <- prior_codes_counts %>% slice_head(n=20)
most_common_prior_codes #%>% left_join(best_codes %>% select(Vocabulary, "Odds ratio", mlog10p), by=c(CODE1="Code")) %>% View()
```

```{r}
diagnoses_before_raw <- helper_diagnosis(diagnoses_before)
tmp <- diagnoses_before_raw %>%
  group_by(ENDPOINT, phenotype) %>% 
  summarise(across(c(min, max, median), mean), .groups="drop") %>% 
  pivot_longer(cols=c(min, max, median)) 
tmp %>% ggplot(aes(value, color=name)) + 
  (if (allow_individual_level_data) geom_freqpoly() else geom_density(key_glyph="path")) +
  labs(x="Mean of distance in years", title="Diagnoses prior to the first hemochromatosis",
       color = "Personal distance")
```


```{r, fig.height=8, fig.width=7}
diagnoses_before_raw %>% 
  filter(ENDPOINT %in% most_common_prior_diagnoses$ENDPOINT) %>%
  mutate(phenotype = factor(phenotype, levels=levels(most_common_prior_diagnoses$phenotype))) %>%
#  mutate(ENDPOINT = factor(ENDPOINT, levels=levels(most_common_prior_diagnoses$ENDPOINT)),
#         phenotype = fct_reorder(phenotype, ENDPOINT)) %>%
  #mutate(phenotype = coalesce(phenotype, ENDPOINT)) %>%
  mutate(phenotype = fct_rev(phenotype)) %>%
#  ggplot(aes(y=phenotype, x=value, color=name)) + 
  ggplot(aes(y=phenotype, x=median)) + 
  #geom_point() +
  geom_boxplot(outliers=allow_individual_level_data) +
  scale_y_discrete(labels=scales::label_wrap(40)) +
  labs(title="Diagnoses before first hemochromatosis event", x="Distance in years", y="Phenotype") +
  theme(legend.position = "bottom")
```

```{r}
codes_before_raw <- helper_code(codes_before)
tmp <- codes_before_raw %>%
  group_by(CODE1) %>% 
  summarise(across(c(min, max, median), mean), .groups="drop") %>% 
  pivot_longer(cols=c(min, max, median)) 
tmp %>% ggplot(aes(value, color=name)) + 
  (if (allow_individual_level_data) geom_freqpoly() else geom_density(key_glyph="path")) +
  labs(x="Mean of distance in years", title="Codes prior to the first hemochromatosis",
       color = "Personal distance")
```

```{r, fig.height=8, fig.width=7}
g <- codes_before_raw %>% 
  filter(CODE1 %in% codewas) %>%
  #inner_join(best_codes, by="Description") %>%
  left_join(best_codes %>% drop_na(Description) %>%mutate(Code = str_remove(Code, "\\.")), by=c(CODE1="Code")) %>%
  mutate(Description = fct_reorder(Description, mlog10p)) %>%
#  mutate(ENDPOINT = factor(ENDPOINT, levels=levels(most_common_prior_diagnoses$ENDPOINT)),
#         phenotype = fct_reorder(phenotype, ENDPOINT)) %>%
  #mutate(phenotype = coalesce(phenotype, ENDPOINT)) %>%
  #mutate(phenotype = fct_rev(phenotype)) %>%
#  ggplot(aes(y=phenotype, x=value, color=name)) + 
  ggplot(aes(y=Description, x=median)) + 
  #geom_point() +
  geom_boxplot(outliers=allow_individual_level_data) +
  scale_y_discrete(labels=scales::label_wrap(40)) +
  labs(title="Codes before first hemochromatosis event", x="Distance in years", y="Description") +
  theme(legend.position = "bottom")
g
if (save_figs) {
  filename <- file.path(figure_path, "codes_before_hc.pdf")
  ggsave(filename, g)
}
```

#### After

```{r}
after_diagnoses_counts <- diagnoses_after_ever %>% 
  count(ENDPOINT, phenotype) %>% 
  mutate(ENDPOINT=fct_reorder(ENDPOINT, n, .desc = TRUE),
         phenotype=fct_reorder(phenotype, n, .desc = TRUE)) %>% 
  arrange(desc(n)) %>% 
  filter(n >= limit)# %>% head()
filename <- file.path(table_path, "after_diagnoses_counts.tsv")
if (save_tables)
  write_tsv(after_diagnoses_counts, filename)
most_common_after_diagnoses <- after_diagnoses_counts %>% slice_head(n=20)
most_common_after_diagnoses %>% select(phenotype, n)
```

How many individuals have had a medical code after the first hc diagnosis?

```{r}
after_codes_counts <- codes_after_ever %>% 
  count(CODE1, code_set) %>% 
  left_join(all_codes %>% select(CODE1, code_set, description), by=c("CODE1", "code_set")) %>%
  mutate(ENDPOINT=fct_reorder(CODE1, n, .desc = TRUE),
         description=fct_reorder(description, n, .desc = TRUE)) %>% 
  arrange(desc(n)) %>% 
  filter(n >= limit) %>% # %>% head()
  left_join(best_codes %>% drop_na(Description) %>% mutate(Code=str_remove(Code, "\\.")), by=c(CODE1="Code")) %>%
  select(code_set, CODE1, description, n, "Odds ratio", mlog10p)
filename <- file.path(table_path, "after_codes_counts.tsv")
if (save_tables)
  write_tsv(after_codes_counts, filename)
most_common_after_codes <- after_codes_counts %>% slice_head(n=20)
most_common_after_codes
```

```{r}
diagnoses_after_raw <- helper_diagnosis(diagnoses_after)
tmp <- diagnoses_after_raw %>%
  group_by(ENDPOINT, phenotype) %>% 
  summarise(across(c(min, max, median), mean), .groups="drop") %>% 
  pivot_longer(cols=c(min, max, median)) 
tmp %>% ggplot(aes(value, color=name)) + 
  (if (allow_individual_level_data) geom_freqpoly() else geom_density(key_glyph="path")) +
  labs(x="Mean of distance in years", title="Diagnoses after the first hemochromatosis",
       color = "Personal distance")

```

```{r, fig.height=8, fig.width=7}
diagnoses_after_raw %>% 
  filter(ENDPOINT %in% most_common_after_diagnoses$ENDPOINT) %>%
  mutate(phenotype = factor(phenotype, levels=levels(most_common_after_diagnoses$phenotype))) %>%
#  mutate(ENDPOINT = factor(ENDPOINT, levels=levels(most_common_prior_diagnoses$ENDPOINT)),
#         phenotype = fct_reorder(phenotype, ENDPOINT)) %>%
  #mutate(phenotype = coalesce(phenotype, ENDPOINT)) %>%
  mutate(phenotype = fct_rev(phenotype)) %>%
#  ggplot(aes(y=phenotype, x=value, color=name)) + 
  ggplot(aes(y=phenotype, x=median)) + 
  #geom_point() +
  geom_boxplot(outliers=allow_individual_level_data) +
  scale_y_discrete(labels=scales::label_wrap(40)) +
  labs(title="Diagnoses after first hemochromatosis event", x="Distance in years", y="Phenotype") +
  theme(legend.position = "bottom")
```

```{r}
codes_after_raw <- helper_code(codes_after)
tmp <- codes_after_raw %>%
  group_by(CODE1) %>% 
  summarise(across(c(min, max, median), mean), .groups="drop") %>% 
  pivot_longer(cols=c(min, max, median)) 
tmp %>% ggplot(aes(value, color=name)) + 
  (if (allow_individual_level_data) geom_freqpoly() else geom_density(key_glyph="path")) +
  labs(x="Mean of distance in years", title="Codes after the first hemochromatosis",
       color = "Personal distance")
```

```{r, fig.height=8, fig.width=7}
g <- codes_after_raw %>% 
  filter(CODE1 %in% codewas) %>%
  #inner_join(best_codes, by="Description") %>%
  left_join(best_codes %>% drop_na(Description) %>% 
              mutate(Code = str_remove(Code, "\\.")), by=c(CODE1="Code")) %>%
  mutate(Description = fct_reorder(Description, mlog10p)) %>%
#  mutate(ENDPOINT = factor(ENDPOINT, levels=levels(most_common_prior_diagnoses$ENDPOINT)),
#         phenotype = fct_reorder(phenotype, ENDPOINT)) %>%
  #mutate(phenotype = coalesce(phenotype, ENDPOINT)) %>%
  #mutate(phenotype = fct_rev(phenotype)) %>%
#  ggplot(aes(y=phenotype, x=value, color=name)) + 
  ggplot(aes(y=Description, x=median)) + 
  #geom_point() +
  geom_boxplot(outliers=allow_individual_level_data) +
  scale_y_discrete(labels=scales::label_wrap(40)) +
  labs(title="Codes after first hemochromatosis event", x="Distance in years", y="Description") +
  theme(legend.position = "bottom")
g
if (save_figs) {
  filename <- file.path(figure_path, "codes_after_hc.pdf")
  ggsave(filename, g)
}
```

Save diagnoses/code counts to excel file.

```{r}
dfs <- list(`Prior codes`=prior_codes_counts, `Posterior codes`=after_codes_counts,
            `Prior diagnoses`=prior_diagnoses_counts, `Posterior diagnoses`=after_diagnoses_counts)
filename <- file.path(table_path, "code_counts.xlsx")
if (save_tables)
  openxlsx::write.xlsx(dfs, file = filename, overwrite=TRUE)
```

### Survival models between endpoints before and after, done by FinnGen

This data is downloaded from here
https://r10.risteys.finngen.fi/endpoints/E4_IRON_MET
For later releases the survival results are not downloadable.

```{r}
hazards <- read_csv(file.path(data_path, "E4_IRON_MET_survival-analyses-1.csv"))
```

```{r}
hazards <- hazards %>% filter(outcome_name == "E4_IRON_MET")
```

Let's check top 10 prior phenotypes with highest adjusted p-value.

```{r}
top10 <- hazards %>% filter(hr_lag > 0) %>% mutate(p_adjusted_BY = p.adjust(p, method="BY")) %>% 
  filter(p_adjusted_BY < 0.05) %>% arrange(p) %>% group_by(prior_name) %>% slice_head(n=1) %>% ungroup() %>% head(n=10)
top10
```





```{r}
prior_endpoint_df <- top10
prior_endpoints <- prior_endpoint_df$prior_name
prior_endpoint_df
```

Let's get all the events for these endpoints.

```{r}
input_filename <- if (release == "r12") "/finngen/library-red/finngen_R12/phenotype_1.0/data/finngen_R12_endpoint_longitudinal_1.0.txt.gz" else "/finngen/library-red/finngen_R13/phenotype_1.0/data/finngen_R13_endpoint_longitudinal_1.0.txt.gz"

#input_filename <- "/finngen/library-red/finngen_R12/phenotype_1.0/data/finngen_R12_endpoint_longitudinal_1.0.txt.gz"
output_filename <- file.path(result_path, "prior_endpoint_events_new.tsv.gz")
comparisons <- paste(sprintf('$6 == "%s"', prior_endpoints), collapse=" || ")
cmd <- sprintf('zcat %s | awk -F "\t" -v OFS="\t" \'NR==1{print}NR>1{if (%s) print}\' | gzip > %s', 
               input_filename, comparisons, output_filename)
cat(cmd)
if (! file.exists(output_filename))
  system(cmd)
```

Careful!!!!!! If you copy paste the above, make sure the tabulators are copied
as well!!!!!!!!!!!!!

This took about 15 minutes.

```{r}
prior_endpoint_events <- read_tsv(output_filename, n_max = Inf) %>% rename(sample=FINNGENID)
```

```{r}
prior_endpoint_events <- prior_endpoint_events %>%
  inner_join(prior_endpoint_df %>% select(prior_name, hr_lag), by=c(ENDPOINT="prior_name"))
```

Should I truncate the followup time so that each register would have the same followup time?

I could also compute end of followup based on APPROX_BIRTH_DATE and AGE_AT_DEATH_OR_END_OF_FOLLOWUP. Let's do that!

```{r}
if (release == "r12") {
  filename <- "/finngen/library-red/finngen_R12/analysis_covariates/R12_COV_V2.FID.txt.gz"
} else {
  filename <- "/finngen/library-red/finngen_R13/analysis_covariates/data/R13_COV_V0.FID.txt.gz"
}
covariates <- read_tsv(filename, col_select = c("FID", "APPROX_BIRTH_DATE", "AGE_AT_DEATH_OR_END_OF_FOLLOWUP")) %>%
  rename(sample=FID)
```

```{r}
covariates <- covariates %>% 
  #head(n=100) %>% 
  mutate(end_of_followup = APPROX_BIRTH_DATE + 365*AGE_AT_DEATH_OR_END_OF_FOLLOWUP)
```

```{r}
hc2 <- hc %>% 
  left_join(covariates %>% select(sample, end_of_followup), by="sample") %>%
  mutate(event_or_eof_date = coalesce(APPROX_EVENT_DAY, end_of_followup)) %>%
  drop_na(event_or_eof_date)
```

```{r}
prior_endpoint_events <- prior_endpoint_events %>%
  inner_join(hc2 %>% select(sample, end_of_followup), by="sample")
```

The exposure must start hr_lag years before the first hemochromatosis event.

Here I think on purpose I use end_of_followup for all instead of event_or_eof_date.

```{r}
prior_endpoint_events <- prior_endpoint_events %>% 
  mutate(threshold = end_of_followup - years(hr_lag))
prior_endpoint_events <- prior_endpoint_events %>% filter(APPROX_EVENT_DAY < threshold)
```

```{r}
prior_endpoint_events %>% select(sample, ENDPOINT) %>% distinct() %>% count(ENDPOINT)
```

Count the event only once.

```{r}
prior_endpoint_events2 <- prior_endpoint_events %>% select(sample, ENDPOINT) %>% distinct()
```

```{r}
prior_disease_events <- prior_endpoint_events2 %>% 
  mutate(value = TRUE) %>% 
  pivot_wider(values_from="value", names_from="ENDPOINT") %>% 
  mutate(across(-sample, ~ replace_na(., FALSE)))
filename <- file.path(result_path, "prior_disease_events.tsv")
if (save_tables)
  write_tsv(prior_disease_events, filename)
```

### Event age distribution

```{r}
all <- dosages %>% 
  select(sample, dosage, name) %>% 
  pivot_wider(names_from=name, values_from=dosage) %>% 
  inner_join(hc_full, by="sample")
```

```{r}
tmp <- all %>% 
  filter(hemochromatosis == "case")
if (allow_individual_level_data) {
  #breaks <- tmp %>% filter(C282Y == 2 & sex == "female") %>% pull(age) %>% cut_helper_variable_width(number_of_breaks = 15) %>% .$breaks
  g <- tmp %>%
    ggplot(aes(age, color=sex)) + 
    geom_density(key_glyph = "path") +
    #geom_freqpoly(breaks = breaks) +
    (if (allow_individual_level_data) geom_rug()) +
    geom_text(data=tmp %>% count(C282Y, sex) %>% mutate(label=sprintf("n=%i",n), y=if_else(sex=="male", 0.8, 0.9)),
              mapping=aes(x=I(0.1), y=I(y), label=label),
              hjust = 0, show.legend = FALSE) +
    labs(x="Event age", y="Density", color="Sex", title=sprintf("%i hemochromatosis cases", nrow(tmp))) +
    facet_grid(C282Y ~ ., labeller=label_both)
  #min_bin_size(g)
  g
} else {
  # This looks like a mess, so I won't show it.
  if (FALSE) {
  helper <- function(tmp2, dosage) {
    tmp <- tmp2 %>% filter(C282Y == dosage)
    breaks <- tmp %>% group_by(sex) %>% cut_helper_variable_width("age") %>% pluck("breaks")
    g <- tmp %>%
    ggplot(aes(age, fill=sex)) + 
    geom_histogram(breaks = breaks, position="identity", alpha=0.5) +
    geom_text(data=tmp %>% count(sex) %>% mutate(label=sprintf("n=%i",n), y=if_else(sex=="male", 0.8, 0.9)),
    mapping=aes(x=I(0.1), y=I(y), label=label),
    hjust = 0, show.legend = FALSE) +
    labs(x="Event age", y="Count", color="Sex", title=sprintf("%i hemochromatosis cases", nrow(tmp)))
    g
  }
  g0 <- helper(tmp, dosage = 0)
  g1 <- helper(tmp, dosage = 1)
  g2 <- helper(tmp, dosage = 2)
  g0 / g1 / g2
  }
  "Figure not shown as it contains individual level data"
}

```











