---
title: "FinnGen HC progression"
author: "Jarkko Toivonen"
date: "`r Sys.Date()`"
format: 
  html:
    df-print: kable
    toc: true
    toc-depth: 4
    embed-resources: true
---

## todo

Talleta bigquery-kyselyn tulokset, ettei tarvi joka kerta ajaa.

## Setup

```{r setup}
library(countreg)
library(topmodels)
library(tidyverse)
library(broom)
library(glue)
library(pscl)  # For zero-inflated models
library(patchwork)
library(performance)
library(exclude)
library(UpSetR)
library(bigrquery)
library(statmod)  # For qresid
library(ggvenn)
library(tictoc)

# This can measure time a chunk took to execute.
# Add chunk option time_it=TRUE to each chunk your want to measure.
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- lubridate::now()
    } else {
      # calculate the time difference after a chunk
      res <- lubridate::now() - now
      # return a character string to show the time
      msg <- sprintf("Time for this code chunk to run: %.2f %s", as.numeric(res), units(res))
      message(msg)
      NULL   # Don't return the message so that it won't be printed to the resulting document.
      msg
    }
  }
}))
knitr::opts_chunk$set(echo=TRUE, time_it=TRUE)

source("common.R")

allow_individual_level_data <- TRUE
recompute <- FALSE
save_figs <- FALSE
save_tables <- FALSE

release <- "r12"; Release <- str_to_upper(release)

result_path <- file.path("../results", release, "progression")

#result_path <- "../results/progression"
table_path  <- file.path(result_path, "tables")
figure_path <- file.path(result_path, "figures")
model_path <- file.path(result_path, "models")

c(figure_path, table_path, model_path) %>% purrr::discard(dir.exists) %>% walk(~dir.create(., recursive = TRUE))  # Create dirs if they don't exist
```

## Functions

Some levels combinations in interactions are missing. Hence, I need to convert to model.matrix and then preprocess the data to
exclude constant columns.

I also drop the interaction C282Y1:donations_in_five_years, since it gets huge std error in all models.

```{r}
use_expanded_matrix <- TRUE  # Let's encode the covariates explicitly
responses <- c("n", "n_all", "n_window", "n_full_window")
my_expand <- function(f, data) {
  m <- model.matrix(as.formula(f), data = data)
  non_constant_cols <- m %>% apply(2, sd) %>% keep(~ . != 0) %>% names()       # Drop constant columns
  covariates <- setdiff(non_constant_cols, "donations_in_five_years:C282Y1")   # this causes huge std err
  m <- cbind(m[, covariates], data %>% select(any_of(responses))) %>% as_tibble()
  m
}
```

```{r}
# Compute qqplot of a count model.
# The fitted model must have name fit_{name}
get_qqplot <- function(name, data) {
  model <- get(sprintf("fit_%s", name), rlang::global_env())
  df <- qqrplot(model, newdata = data, plot=FALSE, simint = TRUE, confint = TRUE)
  autoplot(df, main = name)
}
```

```{r}
get_rootogram <- function(name, data) {
  model <- get(sprintf("fit_%s", name), rlang::global_env())
  tryCatch({
    df <- rootogram(model, newdata = data, plot=FALSE, confint = TRUE)
    autoplot(df, main = name)
  }, error = function(e) {  # Some models did not fit. Hence rootogram fails as well
    NULL
  })
}
```


## Load datas

### Primary health care hemochromatosis

```{r}
if (recompute) {
  # Running some authentication code:
  projectid <- "fg-production-sandbox-5" # NB: ADD YOUR SANDBOX NUMBER!
  options(gargle_oauth_cache = FALSE)
  bq_auth(scopes = "https://www.googleapis.com/auth/bigquery")
  
  # Making a connection object that is used to connect to the tables:
  con = dbConnect(
    bigrquery::bigquery(),
    project = "finngen-production-library",
    dataset = if (release == "r12") "sandbox_tools_r12" else "sandbox_tools_r13",
    billing = projectid
  )
  
  dbListTables(con)           # List available tables
  
  if (release == "r12") {
    detailed = tbl(con, "finngen_r12_service_sector_detailed_longitudinal_v1")
  } else {
    detailed = tbl(con, "finngen_r13_service_sector_detailed_longitudinal_v1")
  }
}
```

```{r}
icd9_hc = c("2750", "2750A")
icd10_hc <- c("E831")
filename <- file.path(table_path, "prim_out_hemochromatosis_individuals.txt")
if (recompute || ! file.exists(filename)) {
  prim_out_case_df <- detailed %>%
    select(FINNGENID:CODE2, ICDVER, CATEGORY) %>%  # selecting only columns we need
    filter(                                        # filtering row conditions that define our cases
      (SOURCE %in% c("PRIM_OUT") & (
        (ICDVER == "10" & (CODE1 == icd10_hc | CODE2 == icd10_hc)) |
          (ICDVER == "9" &  (CODE1 %in% icd9_hc |  CODE2 %in% icd9_hc))
      ))
    ) %>%
    collect()
  prim_out_hemochromatosis_individuals <- prim_out_case_df %>% pull(FINNGENID) %>% unique()
  write_lines(prim_out_hemochromatosis_individuals, filename)
} else {
  prim_out_hemochromatosis_individuals <- read_lines(filename)
}
```

### Related diseases

```{bash, eval=FALSE}
filename=/finngen/library-red/finngen_R12/endpoint_cohorts_v1.0/endpoint_cohorts_r12_v1.csv.gz
output=../results/progression/exclusion_diagnoses.tsv.gz
if [[ ! -f $output ]] ; then
  time zcat $filename | awk 'NR==1;NR>1{if ($3 == 1 && ($2 == "POLYCYTVERA" || $2 == "D3_SCNDPOLYCYTHAEMIA" || $2 == "E4_PORCUTTAR")) print}' | gzip > $output 
fi
```

This took 28 minutes. In the future, do this with bigquery!

```{r}
filename <- file.path(result_path, "exclusion_diagnoses.tsv.gz")
if (recompute || ! file.exists(filename)) {
  if (release == "r12") {
    endpoint_cohorts = tbl(con, "endpoint_cohorts_r12_v1")
  } else {
    endpoint_cohorts = tbl(con, "endpoint_cohorts_r13_v3")
  }
  exclusion_diagnoses <- endpoint_cohorts %>% 
    filter(CONTROL_CASE_EXCL == 1 & (ENDPOINT %in% c("POLYCYTVERA", "D3_SCNDPOLYCYTHAEMIA"))) %>%
    rename(sample=FINNGENID) %>%
    collect()
  write_tsv(exclusion_diagnoses, filename)
} else {
  exclusion_diagnoses <- read_tsv(filename, show_col_types = FALSE) %>% rename(any_of(c(sample="FINNGENID")))
}
```

```{r}
polycythemia_individuals <- exclusion_diagnoses %>% filter(ENDPOINT == "POLYCYTVERA") %>% pull(sample) %>% unique() 
secondary_polycythemia_individuals <- exclusion_diagnoses %>% filter(ENDPOINT == "D3_SCNDPOLYCYTHAEMIA") %>% pull(sample) %>% unique() 
porphyria_individuals    <- exclusion_diagnoses %>% filter(ENDPOINT == "E4_PORCUTTAR") %>% pull(sample) %>% unique() 
sprintf("There are %i individuals with polycythemia, %i individuals with secondary polycythemia and %i individuals with porphyria\n", 
        length(polycythemia_individuals), length(secondary_polycythemia_individuals), length(porphyria_individuals))
```

### End of followup age and date

Get end of followup age and date.

```{r}
if (release == "r12") {
  filename <- "/finngen/library-red/finngen_R12/analysis_covariates/R12_COV_V2.FID.txt.gz"
} else {
  filename <- "/finngen/library-red/finngen_R13/analysis_covariates/data/R13_COV_V0.FID.txt.gz"
}
eof_orig <- read_tsv(filename, col_select = c("FID", "APPROX_BIRTH_DATE", 
                                         "AGE_AT_DEATH_OR_END_OF_FOLLOWUP")) %>%
  rename(sample=FID, eof_age = AGE_AT_DEATH_OR_END_OF_FOLLOWUP) %>%
  mutate(eof_date = APPROX_BIRTH_DATE + 365*eof_age)
eof <- eof_orig %>% select(sample, eof_age)
eof_date <- eof_orig %>% select(sample, eof_date)
```

### Venesections

What codes are used for phlebotomy and venesection?

```{bash}
grep -Ei "FG_CODE1|phlebotomy|venesection" /finngen/library-green/medical_codes/fg_codes_info_v7/fg_codes_info.csv > /tmp/procedures.tsv
```

```{r}
procedure_codes <- read_tsv("/tmp/procedures.tsv", show_col_types = FALSE)
procedure_codes %>% select(-c(FG_CODE2:FG_CODE9, extra_info)) # Drop the NA columns
```



```{r}
# Read from the main results directory
venesections <- read_tsv(file.path("../results", release, "venesections.tsv.gz"))
venesections <- venesections %>% 
  mutate(operation = if_else(CODE1=="TPH00", "Phlebotomy", "Venesection")) %>%
  rename(sample = FINNGENID)
if (allow_individual_level_data) {
  venesections %>% select(-sample) %>% head()
}
venesections <- venesections %>% 
  mutate(CATEGORY = fct_relabel(CATEGORY, ~ str_sort(., numeric = TRUE)))
venesections %>% select(-sample) %>% mutate(across(where(is.character), as.factor)) %>% summary()
```



```{r}
if (allow_individual_level_data) {
  venesections %>% 
  select(CATEGORY) %>% table()
}
```

```{r}
if (allow_individual_level_data) {
  venesections %>% select(CODE1, CATEGORY) %>% table()
}
```

The SPAT1074 codes are solely from the primary health care.

```{r}
if (allow_individual_level_data) {
  venesections %>% select(SOURCE, CATEGORY) %>% table()
}
```

```{r}
venesections %>% select(CODE1, SOURCE) %>% table()
```

```{r}
g1 <- venesections %>% ggplot(aes(EVENT_AGE, fill=operation)) + geom_histogram(alpha=0.5,
                 position="identity")
breaks <- venesections %>% 
  group_by(operation) %>% 
  cut_helper_variable_width("EVENT_AGE") %>% 
  pluck("breaks")
g2 <- venesections %>% ggplot(aes(EVENT_AGE, fill=operation)) + 
  geom_histogram(alpha=0.5,
                 position="identity",
                 breaks=breaks) +
  labs(subtitle="Variable-width histogram") + 
  lims(x=c(0,100))
min_bin_size(g2)
g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
```

```{r}
#venesections %>% ggplot(aes(APPROX_EVENT_DAY, color=operation)) + geom_freqpoly()
breaks <- venesections %>% 
  group_by(operation) %>% 
  cut_helper_variable_width("APPROX_EVENT_DAY") %>% 
  pluck("breaks")
g <- venesections %>% ggplot(aes(APPROX_EVENT_DAY, fill=operation)) + 
  geom_histogram(alpha=0.5,
                 position="identity",
                 breaks=breaks) +
  labs(subtitle="Variable-width histogram")
min_bin_size(g)
g
```


Get train ids.

```{r}
filename <- file.path("../results", release, "training_sample_ids.tsv.gz")
#filename <- file.path(result_path, "training_sample_ids.tsv.gz")
training_ids <- read_tsv(filename)
training_ids <- training_ids %>% rename(sample = FID)
```

Read hemochromatosis status.

```{r}
filename <- file.path("../results", release, "hemochromatosis_full.rds")
#filename <- file.path(result_path, "hemochromatosis_full.rds")
hc_full <- readRDS(filename)
hc <- hc_full %>% select(sample, hemochromatosis)
hc_count <- hc %>% count(hemochromatosis)
hc_ids <- hc %>% filter(hemochromatosis == "case") %>% pull(sample)
```

Upsetplot.

```{r}
all_venesection_ids <- unique(venesections$sample)   # This contains both venesections and phlebotomies
real_venesection_ids <- venesections %>% filter(operation == "Venesection") %>% pull(sample) %>% unique()
phlebotomy_ids <- venesections %>% filter(operation == "Phlebotomy") %>% pull(sample) %>% unique()
people <- list(hemochromatosis = hc_ids, 
               extra_hc = setdiff(prim_out_hemochromatosis_individuals, hc_ids),
               venesection = all_venesection_ids, 
               #venesection = real_venesection_ids, 
               #phlebotomy = phlebotomy_ids,
               polycythemia = polycythemia_individuals,
               secondary_polycythemia = secondary_polycythemia_individuals,
               porphyria = porphyria_individuals)
#
#png(filename, width = 180/25.4)
u <- UpSetR::upset(UpSetR::fromList(people),#, order.by = "freq", 
                       set_size.show=TRUE,
                  nsets = 7,
                  nintersects = if (allow_individual_level_data) 40 else 16,
                  order.by = "freq",
                       #empty.intersections = TRUE,
                       set_size.scale_max = 1400, # This doesn't help to explain where donors from eProgesa disappear
                       #nintersects = NA   # Show all intersections. Does not work. Six seems to be the maximum
                       # number of intersections
                       )
if (save_figs) {
  filename <- file.path(figure_path, "venesection_upset.pdf")
  pdf(filename, width = 180/25.4, height=90/25.4)
  print(u)
  invisible(dev.off())
}
grid::grid.newpage()
print(u, newpage = FALSE)
```

Very few of the venesections are explained by the extra hemochromatosis cases from primary health care.

```{r}
venesections <- venesections %>% inner_join(hc, by="sample")
```

## Venesection statistics

Get the number of operations and individuals involved.

```{r}
helper <- function(df, operation_type) {
  tmp <- df %>% filter(operation == operation_type)
  tibble(operation=operation_type, operations=nrow(tmp), individuals=n_distinct(tmp$sample))
}
```

```{r}
map(c("Venesection", "Phlebotomy"), ~ helper(venesections, .)) %>% bind_rows()
```

Statistics stratified by hemochromatosis status.


```{r}
get_statistics <- function(venesections, variables=c("operation", "hemochromatosis")) {
  tmp1 <- venesections %>% select(sample, all_of(variables)) %>% count(across(all_of(variables)), name="operations")
  tmp2 <- venesections %>% select(sample, all_of(variables)) %>% distinct() %>% count(across(all_of(variables)), name="treated")
  tmp <- tmp1 %>%
    inner_join(tmp2, by=variables) %>%
    inner_join(hc_count, by="hemochromatosis") %>%
    mutate(operations_per_treated = operations / treated,
           treated_fraction = treated / n)
  tmp
}
```

```{r}
df <- get_statistics(venesections)
df
```

```{r}
filename <- file.path(table_path, "venesection_statistics.tsv")
if (save_tables)
  write_tsv(df, filename)
```


Check the statistics again. This time unstratified.

```{r}
get_statistics(venesections, "hemochromatosis")
```

## Visualize venesections

When the venesections occur?

```{r}
venesections_extra <- venesections %>% 
  group_by(sample) %>% 
  mutate(n = n(),
         days = as.numeric(max(APPROX_EVENT_DAY) - min(APPROX_EVENT_DAY)),
         years = days / 365,
         diff = c(NA, diff(APPROX_EVENT_DAY)),
         switch = as.integer(diff >= 365) %>% replace_na(0), # New treatment series begins?
         series = 1 + cumsum(switch)) %>%   
  ungroup()
```

```{r}
venesections_extra %>% 
  ggplot(aes(APPROX_EVENT_DAY)) +
  geom_histogram()
```

Distribution of treatment series.

```{r}
venesections_extra %>% 
  group_by(sample, hemochromatosis) %>% summarise(number_of_series = max(series), .groups = "drop") %>%
  mutate(number_of_series = factor(number_of_series)) %>%
  ggplot(aes(number_of_series)) + geom_bar() + facet_wrap(~ hemochromatosis)
```

```{r}
g1 <- venesections_extra %>% filter(hemochromatosis == "case") %>%
  ggplot(aes(x = EVENT_AGE, y = sample, color = operation)) +
  geom_point() + labs(title = "Case") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
g2 <- venesections_extra %>% filter(hemochromatosis == "control") %>%
  ggplot(aes(x = EVENT_AGE, y = sample, color = operation)) +
  geom_point() + labs(title = "Control") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
```

Long careers:

```{r}
venesections_extra %>% filter(years >= 1, hemochromatosis == "case") %>%
  ggplot(aes(x = EVENT_AGE, y = sample, color = operation)) +
  geom_point() + labs(title = "Case") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

New treatment series is defined here to start after a one year break.

```{r}
venesections_extra %>% filter(years >= 1, hemochromatosis == "case") %>%
  mutate(series = as.factor(series)) %>%
  ggplot(aes(x = EVENT_AGE, y = sample, color = series)) +
  geom_point() + labs(title = "Case") +
  labs(color = "Treatment series") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```




Below, colored circle denotes the HC diagnosis and hollow black circle denotes end of followup.

```{r}
#| fig-width: 5
#| fig-height: 30
tmp <- venesections_extra %>% 
  filter(hemochromatosis == "case", n >= 4) %>%
  left_join(hc_full %>% select(sample, sex, diagnosis_date = approx_event_date, diagnosis_age = age), by="sample") %>%
  left_join(eof, by="sample") %>%
  mutate(sample = fct_reorder(sample, diagnosis_age))
n <- n_distinct(tmp$sample)
tmp %>%
  ggplot(aes(EVENT_AGE, sample)) + 
  geom_point(aes(x = diagnosis_age, color = sex)) +
  geom_point(aes(x = eof_age), color="black", shape=21) +
  geom_violin(scale = "count", position = "identity", width = 25) + 
  coord_cartesian(y=c(0, n+1)) +
  labs(title = sprintf("%i cases with venesections", n)) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  NULL
```


```{r}
#| fig-width: 5
#| fig-height: 6
n <- 10
controls <- venesections_extra %>% 
  filter(hemochromatosis == "control", n >= 2) %>% pull(sample) %>% unique() %>% sample(10)
tmp <- venesections_extra %>% 
  #filter(hemochromatosis == "control", n >= 4) %>%
  filter(sample %in% controls) %>%
  #left_join(hc_full %>% select(sample, sex, diagnosis_date = approx_event_date, diagnosis_age = age), by="sample") %>%
  left_join(eof, by="sample") %>%
  mutate(sample = fct_reorder(sample, eof_age))
#limits <- unique(tmp$sample)
#n <- n_distinct(tmp$sample)
tmp %>%
  ggplot(aes(EVENT_AGE, sample)) + 
  #geom_histogram()
  #geom_point(aes(x = diagnosis_age, color = sex)) +
  geom_point(aes(x = eof_age), color="black", shape=21) +
  geom_violin(scale = "count", position = "identity", width = 20) + 
  #geom_violin(scale = "count", position = "identity") + 
  #scale_y_discrete(expand = expansion(add = -8)) + #, limits=limits) +
  coord_cartesian(y=c(0, n+1)) +
  #lims(y=c(1, n)) +
  #geom_violin(scale = "area") +
  labs(title = sprintf("Controls, sample of %i individuals", n)) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  NULL
```

Are there some people who have gotten both venesections and phlebotomies.

```{r}
setlist <- list(venesections = venesections %>% filter(operation == "Venesection") %>% pull(sample) %>% unique(),
                phlebotomy = venesections %>% filter(operation == "Phlebotomy") %>% pull(sample) %>% unique())
ggvenn(setlist)
```

Distribution of treatment lengths.

```{r}
venesections_extra %>% select(sample, years, hemochromatosis) %>% distinct() %>% ggplot(aes(years)) + geom_histogram() +
  facet_wrap(~ hemochromatosis)
```


## Form data

Compute the combined number of operations. Take only the venesections from the first two years.

```{r}
tmp <- venesections %>% 
  left_join(hc_full %>% select(sample, sex, diagnosis_date = approx_event_date, diagnosis_age = age), by="sample") %>%
  group_by(sample) %>%
  mutate(first_venesection_date = min(APPROX_EVENT_DAY)) %>%
  ungroup()
tmp1 <- tmp %>%
  count(sample, hemochromatosis, first_venesection_date, name = "n_all") 
tmp2 <- tmp %>%
  #filter(hemochromatosis == "case") %>%
  filter(first_venesection_date <= APPROX_EVENT_DAY, APPROX_EVENT_DAY <= first_venesection_date + years(2)) %>% 
  count(sample, hemochromatosis, first_venesection_date, name = "n_window")  # Window might be less than two years of length
tmp3 <- tmp %>%
  filter(first_venesection_date <= APPROX_EVENT_DAY, APPROX_EVENT_DAY <= first_venesection_date + years(2)) %>% 
  count(sample, hemochromatosis, first_venesection_date, name = "n_full_window") %>% # Exactly two-year window
  left_join(eof_date, by="sample") %>%
  mutate(diff = eof_date - first_venesection_date) %>% 
  mutate(n_full_window = if_else(diff < 2*365, -n_full_window, n_full_window))
venesection_counts <- hc %>% 
  left_join(tmp1, by = c("sample", "hemochromatosis")) %>%
  left_join(tmp2, by = c("sample", "hemochromatosis", "first_venesection_date")) %>%
  left_join(tmp3, by = c("sample", "hemochromatosis", "first_venesection_date")) %>%
  replace_na(list(n_all = 0, n_window = 0)) %>%
  mutate(n = n_all) %>%
  select(-c(eof_date, diff))
  #mutate(n = replace_na(n, 0))
rm(tmp, tmp1, tmp2, tmp3)
```




```{r}
filename <- file.path("../results", release, "basic.rds")
basic <- readRDS(filename)
```

Combined the datas.

```{r}
combined <- inner_join(
  venesection_counts, 
  basic %>% select(-starts_with("int")), 
  by=c("sample", "hemochromatosis"))
```

Take all individuals in the training set.

```{r}
train_data2 <- combined %>% semi_join(training_ids)
```



```{r}
g <- train_data2 %>%
  #filter(n <= 10) %>%
  ggplot(aes(n)) +
  #geom_density() +
  geom_bar() +
  facet_grid(hemochromatosis ~ ., scales="free") +
  #scale_x_log10() +
  scale_y_log10() +
  labs(x="Number of operations")
min_bin_size(g)
g
```

### Exclude individuals with certain diagnoses.


Do the exclusion.

```{r}
helper <- function(df, x) { df %>% filter(hemochromatosis==x) %>% pull(n) %>% sum() }
statistics <- function(df) list(cases=sum(df$hemochromatosis=="case"), controls=sum(df$hemochromatosis=="control"),
                                case_donations=helper(df, "case"), control_donations=helper(df, "control"))
```

#### Train data

```{r}
filename <- file.path(result_path, "train_data.rds")
exclusions_filename <- file.path(table_path, "train_exclusions.tsv")
if (! file.exists(filename) || recompute) {
  train_data <- train_data2 %>%
    init_exclude(statistics=statistics) %>%
    filter(! sample %in% polycythemia_individuals) %>%
    exclude("Polycythemia") %>%
    filter(! sample %in% porphyria_individuals) %>%
    exclude("Porphuria cutanea tarda")
  saveRDS(train_data, filename)
  # Save exclusion tibble to file
  e <- get_exclude()
  train_exclusions <- as_tibble(e)
  write_tsv(train_exclusions, exclusions_filename)
} else {
  train_data <- readRDS(filename)
  train_exclusions <- read_tsv(exclusions_filename)
}
```


```{r}
rm(train_data2)
```

```{r}
train_exclusions %>%
  exclude::plot_flow() %>% 
  DiagrammeR::grViz()
```

#### Full data

```{r}
filename <- file.path(result_path, "full_data.rds")
exclusions_filename <- file.path(table_path, "full_exclusions.tsv")
if (! file.exists(filename) || recompute) {
  full_data <- combined %>%
    init_exclude(statistics=statistics) %>%
    filter(! sample %in% polycythemia_individuals) %>%
    exclude("Polycythemia") %>%
    filter(! sample %in% porphyria_individuals) %>%
    exclude("Porphuria cutanea tarda")
  saveRDS(full_data, filename)
  # Save exclusion tibble to file
  e <- get_exclude()
  full_exclusions <- as_tibble(e)
  write_tsv(full_exclusions, exclusions_filename)
} else {
  full_data <- readRDS(filename)
  full_exclusions <- read_tsv(exclusions_filename)
}
if (save_figs) {
  filename <- file.path(figure_path, "full_exclusions.dot")
  full_exclusions %>%
    exclude::plot_flow() %>%
    cat(file = filename)
}

```

```{r}
rm(combined)
```

```{r}
full_exclusions %>%
  exclude::plot_flow() %>% 
  DiagrammeR::grViz()
```

```{r}
sprintf("Train data has %i individuals\n", nrow(train_data)) %>% cat()
sprintf("Full data has %i individuals\n", nrow(full_data)) %>% cat()
```


### Check the counts again.

```{r}
train_data %>%
  #filter(n <= 10) %>%
  ggplot(aes(n)) +
  #geom_density() +
  geom_bar() +
  geom_text(data = train_data %>% count(hemochromatosis), mapping = aes(x = I(0.5), y = I(0.5), label=sprintf("n=%i", n))) +
  facet_grid(hemochromatosis ~ ., scales="free") +
  #scale_x_log10() +
  scale_y_log10() +
  labs(x="Number of operations", y="Number of individuals")
```

```{r}
cases <- train_data %>% filter(hemochromatosis == "case") %>% select(-hemochromatosis)
n <- nrow(train_data)   # Number of individuals
```

```{r}
cases %>% 
  ggplot(aes(n, fill=sex2)) +
  geom_bar(position="dodge") +
  #scale_y_log10() +
  labs(x="Number of operations", title="Number of venesections in hemochromatosis cases",
       subtitle = sprintf("%i cases", nrow(cases))) +
  theme(legend.position = "bottom")
```

One of the outliers got removed after filtering out individual with other diseases for which phlebotomy is a treatment.

### Data with exposure

Take only venesections after the first hemochromatosis diagnosis.

```{r}
helper <- function(df, x) { df %>% filter(hemochromatosis==x) %>% pull(n) %>% sum() }
helper2 <- function(df, x) { df %>% filter(hemochromatosis==x) %>% pull(sample) %>% n_distinct() }
statistics <- function(df) list(venesections = sum(! is.na(df$venesection_date)),
                                cases = helper2(df, "case"), 
                                controls = helper2(df, "control"))
```

```{r}
combined_with_exposure <- hc_full %>% 
  left_join(eof_date, by="sample") %>% 
  select(sample, hemochromatosis, sex, diagnosis_date = approx_event_date, diagnosis_age = age, eof_date) %>%
  left_join(venesections %>% rename(venesection_age = EVENT_AGE, venesection_date = APPROX_EVENT_DAY) %>% select(-c(CODE2:CODE4, ICDVER, INDEX)), 
             by=c("sample", "hemochromatosis")) %>%
  #left_join(basic %>% select(-starts_with("int")), by=c("sample", "hemochromatosis")) %>%
  mutate(exposure = as.numeric((eof_date - diagnosis_date) / 365)) %>%
  init_exclude(statistics=statistics) %>%
  filter(hemochromatosis == "case") %>% 
  exclude("Controls") %>%
  filter(is.na(venesection_date) | (venesection_date >= diagnosis_date)) %>%
  exclude("Drop venesections before diagnosis") %>%
  filter(! sample %in% polycythemia_individuals) %>%
  exclude("Polycythemia") %>%
  filter(! sample %in% porphyria_individuals) %>%
  exclude("Porphuria cutanea tarda") %>%
  filter(diagnosis_date <= eof_date) %>%
  exclude("Diagnosis after EOF") %>%
  nest(venesection_df = c(SOURCE, venesection_age, venesection_date, CODE1, CATEGORY, operation)) %>%
  mutate(n = map_int(venesection_df, function(df) { sum(! is.na(df$venesection_date)) } )) %>%
  select(- venesection_df) %>%
  left_join(basic %>% select(-starts_with("int")), by=c("sample", "hemochromatosis", "sex"))
  #count(sample, hemochromatosis, exposure, name = "n")
```

For some reason, some people have HC diagnosis after end of follow-up.

Save and plot exclusions.

```{r}
# Save exclusion tibble to file
e <- get_exclude()
full_exposure_exclusions <- as_tibble(e)
if (save_tables) {
  exclusions_filename <- file.path(table_path, "full_exposure_exclusions.tsv")
  write_tsv(full_exclusions, exclusions_filename)
}
if (save_figs) {
  filename <- file.path(figure_path, "full_exposure_exclusions.dot")
  full_exposure_exclusions %>%
    exclude::plot_flow() %>%
    cat(file = filename)
  filename <- file.path(figure_path, "full_exposure_exclusions.svg")
  #plot(e) %>%
  full_exposure_exclusions %>%
    exclude::plot_flow() %>% 
    DiagrammeR::grViz() %>%
    DiagrammeRsvg::export_svg() %>% 
    cat(file = filename)
}

full_exposure_exclusions %>%
  exclude::plot_flow() %>% 
  DiagrammeR::grViz()
```

```{r, eval=FALSE}
r"(digraph graphname {node [shape="box"];labeljust = l P1 [label = "Original data\lvenesections=7,104\lcases=477\l"]})" %>%
  #cat() %>%
  DiagrammeR::grViz()
```

```{r}
g1 <- combined_with_exposure %>% ggplot(aes(n)) + geom_bar() + labs(x="Number of venesections")
```

```{r}
g2 <- combined_with_exposure %>% pivot_longer(cols=c(diagnosis_age, exposure)) %>% ggplot(aes(value)) + geom_histogram() + 
  facet_wrap(~name, scales="free", labeller = as_labeller(c(diagnosis_age = "HC diagnosis age", exposure="Exposure to venesections in years")))
#labs(x="Exposure to venesections in years")
```

```{r}
g3 <- combined_with_exposure %>% ggplot(aes(diagnosis_date)) + geom_histogram()
```

```{r}
g1 + g3 / g2 + plot_annotation(title = "Exposure model data", 
                               subtitle = glue("{nrow(combined_with_exposure)} hemochromatosis patients, {sum(combined_with_exposure$n)} venesections"))
```

## Who are the outliers?

No outliers visible anymore.

```{r, eval=FALSE}
outliers <- cases %>% filter(sex2 == "pre") %>% filter(n > 50)
#View(outliers)
outlier_ids <- outliers %>% pull(sample)
```

```{r, eval=FALSE}
conv <- set_names(outlier_ids, paste0("pseudonym", 1))
outlier_venesections <- venesections %>% 
  filter(sample %in% outlier_ids) %>%
  mutate(sample = fct_recode(sample, !!!conv))
outlier_venesections %>%
  ggplot(aes(x=EVENT_AGE, y=1)) + geom_point() + 
  facet_grid(sample ~ ., scales = "free")
  #ggh4x::facet_grid2(sample ~ ., scales = "free", independent=TRUE)
outlier_venesections %>%
  ggplot(aes(x=APPROX_EVENT_DAY, y=1)) + geom_point() + 
  facet_grid(sample ~ ., scales = "free")
```

They are not relatives and they are not born in the same place.


#### Pedigree.

No outliers anymore.

```{r, eval=FALSE}
filename <- "/finngen/library-red/finngen_R12/kinship_1.0/data/finngen_R12_pedigree.fam"
cmd <- glue('grep -E "{conv["pseudonym1"]}" {filename}')
#print(cmd)
system(cmd)
```



```{r, eval=FALSE}
filename <- "/finngen/library-red/finngen_R12/kinship_1.0/data/finngen_R12.kin0"
cmd <- glue('grep -E "{conv["pseudonym1"]}" {filename}')
system(cmd)
```

They are not related.

#### Stratified summary statistics of venesection counts

```{r}
cases %>% group_by(sex2) %>% summarise(individuals = n(), count_mean = mean(n), count_var = var(n)) %>%
  bind_rows(
    cases %>% summarise(individuals = n(), count_mean = mean(n), count_var = var(n)) %>% mutate(sex2="all")
  )
```


## Covariates

Plot age and number of donations.

```{r}
breaks1 <- cut_helper_variable_width(cases, "age")$breaks
g1 <- cases %>% #pivot_longer(cols=c(age, donations_in_five_years)) %>%
  ggplot(aes(age)) + geom_histogram(breaks = breaks1)# + facet_wrap(~name, scales="free")
g2 <- cases %>% 
  mutate(across(donations_in_five_years, as.factor)) %>% 
  ggplot(aes(donations_in_five_years)) + geom_bar() + scale_y_log10()
g1 + g2
```

```{r}
sprintf("There are %s cases with blood donations", sum(cases$donations_in_five_years > 0))
```



Check the time from first venesection to eof.

```{r}
tmp <- full_data %>% left_join(eof_date, by="sample")
tmp2 <- tmp %>% 
  mutate(diff = eof_date - first_venesection_date) %>% 
  select(hemochromatosis, diff) %>% filter(! is.na(diff)) %>%
  mutate(inside = diff < 2*365)
tmp2 %>% ggplot(aes(as.numeric(diff/365))) + geom_histogram() + labs(x="Years", title="Time from first venesection to eof")
res <- tmp2 %>%
  summarise(s = sum(inside), f = mean(inside), n = n())
```

```{r}
sprintf("%.1f%% (%i / %i) of venesectees are still in their initial two-year window at the end of followup\n", 
        res$f * 100,
        res$s, res$n) %>% 
  cat()
```

## Modeling

### Scale and transform

```{r}
train_scaled <- train_data %>% 
  mutate(age = as.numeric(scale(age)),
         across(c(H63D, C282Y), as.factor))
```

```{r}
full_scaled <- full_data %>% 
  mutate(age = as.numeric(scale(age)),
         across(c(H63D, C282Y), as.factor))
```

### Fit models

```{r}
#step_interact(terms = ~ starts_with("C282Y") : c(starts_with("H63D"), S65C, supersnip, donations_in_five_years, sex2_pre, sex2_post
variables5 <- c("sex2", "age", "donations_in_five_years", 
                #"H1_2", # is correlated with C282Y
                "H63D", "S65C", "C282Y", 
                "supersnip",
                "credible1", "credible2", 
                "credible3",
                "rs3094093", "rs114179634", #, "rs9261354"  # HLA snips
                # Interactions
                "C282Y:H63D", "C282Y:S65C", "C282Y:supersnip", "C282Y:donations_in_five_years", "C282Y:sex2"
                )
f5v <- sprintf("n ~ %s", paste(variables5, collapse=" + "))
f5v_zi <- paste(f5v, "| 1")
f5v
f5v_zi
```




```{r}
m <- my_expand(f5v, train_scaled)
m_full <- my_expand(f5v, full_scaled)
stopifnot(all(colnames(m) == colnames(m_full)))   # Make sure same set of variables are used for both train and full data
summary(m)
```

Drop unneeded data.

```{r}
rm(train_data, train_scaled)
rm(full_data, full_scaled)
```

Final formula:

```{r}
expanded_variables <- setdiff(colnames(m_full), responses)
f_basic <- sprintf("n ~ %s", expanded_variables %>% paste(collapse = " + "))
f_basic
f_basic1 <- sprintf("n ~ %s | 1", expanded_variables %>% paste(collapse = " + "))
f_basic1
```

#### Poisson

Train data.

Warning: glm.fit: fitted rates numerically 0 occurred

```{r}
filename <- file.path(model_path, "poisson.rds")
if (! file.exists(filename) || recompute) {
  tic("poisson")
  if (use_expanded_matrix)
    fit_poisson <- glm(as.formula(f_basic), data = m, family = poisson(link="log"))
  else 
    fit_poisson <- glm(as.formula(f5v), data=train_scaled, family = poisson(link="log"))
  toc()
  saveRDS(fit_poisson, filename)
} else {
  fit_poisson <- readRDS(filename)
}
summary(fit_poisson)
```

Full data.

```{r}
filename <- file.path(model_path, "poisson_full.rds")
if (! file.exists(filename) || recompute) {
  tic("poisson_full")
  if (use_expanded_matrix)
    fit_poisson_full <- glm(as.formula(f_basic), data = m_full, family = poisson(link="log"))
  else
    fit_poisson_full <- glm(as.formula(f5v), data = full_scaled, family = poisson(link="log"))
  toc()
  saveRDS(fit_poisson_full, filename)
} else {
  fit_poisson_full <- readRDS(filename)
}
summary(fit_poisson_full)
```

```{r}
my_ggdiagnostics(fit_poisson_full)
```

Does not look very good. Probably the peak in Cook's distance is due to those observations with venesections.

```{r}
variable_plots(fit_poisson_full)
```

#### Negative binomial

The NegBin(mu, theta) distribution is parameterized by two parameters. The mean mu is positive real number
estimated for each observation (the response of the model). The shape parameter theta is common for all observation
and is estimated during fitting with MASS::glm.nb. If one fits with glm function, the theta is not estimated and
must be given as parameter.

Does not work after removing the restriction to cases only. It does work again after resplitting the data!

Warning: glm.fit: fitted rates numerically 0 occurred

```{r, eval=TRUE}
filename <- file.path(model_path, "nb.rds")
if (! file.exists(filename) || recompute) {
  tic("nb")
  if (use_expanded_matrix)
    fit_nb <- MASS::glm.nb(as.formula(f_basic), data=m, control=glm.control(maxit=200))  # Default number of iterations is 25
  else
    fit_nb <- MASS::glm.nb(as.formula(f5v), train_scaled, control=glm.control(maxit=200))  # Default number of iterations is 25
  toc()
  saveRDS(fit_nb, filename)
  } else {
  fit_nb<- readRDS(filename)
}
summary(fit_nb)
```

Full data.

```{r}
filename <- file.path(model_path, "nb_full.rds")
if (! file.exists(filename) || recompute) {
  tic("nb_full")
  if (use_expanded_matrix)
    fit_nb_full <- MASS::glm.nb(as.formula(f_basic), data = m_full, control=glm.control(maxit=200))
  else
    fit_nb_full <- MASS::glm.nb(as.formula(f5v), data = full_scaled, control=glm.control(maxit=200))
  toc()
  saveRDS(fit_nb_full, filename)
} else {
  fit_nb_full <- readRDS(filename)
}
summary(fit_nb_full)
```

"Dispersion parameter for Negative Binomial(7e-04) family taken to be 1"
The difference 0.00030745 != 7e-4 is probably due to the algorithm failing to converge.

See also
https://stats.stackexchange.com/questions/583437/dispersion-parameter-in-negative-binomial/583798#583798

Theta is the shape parameter of the negative binomial distribution, which the glm.nb function estimates. This is sometimes also
confusingly called "dispersion parameter", but not here. The summary method by default assumes that dispersion is 1.
But it can be also estimated with

```{r, eval=TRUE}
sum(residuals(fit_nb, type = "pearson")**2) / summary(fit_nb)$df.residual
```

The summary method can be instructed to compute and use the dispersion parameter with

```{r, eval=TRUE}
summary.glm(fit_nb, dispersion = NULL)$dispersion
```

Not sure why there is a small difference. But anyway they are both close to 1.

```{r}
my_ggdiagnostics(fit_nb_full)
```


#### Zero-inflated poisson

Train data.

Warning: glm.fit: fitted rates numerically 0 occurred

```{r}
filename <- file.path(model_path, "zip.rds")
if (! file.exists(filename) || recompute) {
  tic("zip")
  if (use_expanded_matrix)
    fit_zip <- zeroinfl(as.formula(f_basic1), data = m, dist="poisson")
  else
    fit_zip <- zeroinfl(as.formula(f5v_zi), data = train_scaled, dist="poisson")
  toc()
  saveRDS(fit_zip, filename)
  } else {
  fit_zip <- readRDS(filename)
}
summary(fit_zip)
```

Full data.

```{r}
filename <- file.path(model_path, "zip_full.rds")
if (! file.exists(filename) || recompute) {
  tic("zip_full")
  if (use_expanded_matrix)
    fit_zip_full <- zeroinfl(as.formula(f_basic1), data = m_full, dist="poisson")
  else
    fit_zip_full <- zeroinfl(as.formula(f5v_zi), data = full_scaled, dist="poisson")
  toc()
  saveRDS(fit_zip_full, filename)
  } else {
  fit_zip_full <- readRDS(filename)
}
summary(fit_zip_full)
```

#### Zero-inflated negative binomial

Train data.

Warning: glm.fit: fitted rates numerically 0 occurred

```{r}
filename <- file.path(model_path, "zinb.rds")
if (! file.exists(filename) || recompute) {
  tic("zinb")
  if (use_expanded_matrix)
    fit_zinb <- zeroinfl(as.formula(f_basic1), data = m, dist="negbin")
  else
    fit_zinb <- zeroinfl(as.formula(f5v_zi), data = train_scaled, dist="negbin")
  toc()
  saveRDS(fit_zinb, filename)
} else {
  fit_zinb <- readRDS(filename)
}
summary(fit_zinb)
```

Full data.

```{r}
filename <- file.path(model_path, "zinb_full.rds")
if (! file.exists(filename) || recompute) {
  tic("zinb_full")
  if (use_expanded_matrix)
    fit_zinb_full <- zeroinfl(as.formula(f_basic1), data = m_full, dist="negbin")
  else
    fit_zinb_full <- zeroinfl(as.formula(f5v_zi), data = full_scaled, dist="negbin")
  toc()
  saveRDS(fit_zinb_full, filename)
} else {
  fit_zinb_full <- readRDS(filename)
}
summary(fit_zinb_full)
```

These models use the same predictors for both the count and zero components.

#### ZIP2

Train data.

Warning: glm.fit: fitted probabilities numerically 0 or 1 occurredWarning: glm.fit: fitted rates numerically 0 occurredWarning: system is computationally singular: reciprocal condition number = 9.63268e-20FALSE

```{r}
filename <- file.path(model_path, "zip2.rds")
if (! file.exists(filename) || recompute) {
  tic("zip2")
  if (use_expanded_matrix)
    fit_zip2 <- zeroinfl(as.formula(f_basic), data = m, dist="poisson")
  else
    fit_zip2 <- zeroinfl(as.formula(f5v), data = train_scaled, dist="poisson")
  toc()
  saveRDS(fit_zip2, filename)
} else {
  fit_zip2 <- readRDS(filename)
}
summary(fit_zip2)
```

Testing capturing warnings.



```{r, eval=FALSE}
f <- function() {
  Sys.sleep(1)
  warning("Eka\n")
  Sys.sleep(1)
  warning("Toka\n")
  Sys.sleep(1)
  warning("Kolmas\n")
  Sys.sleep(1)
  warning("Kolmas\n")
  "return value"
}
```



```{r, eval=FALSE}
options(warn = 1)  # Show warnings immediately
res <- capture_warnings(f())
c(result, warnings) %<-% capture_warnings(f())  # using zeallot package 

options(warn = o)  # Show all warnings only at the end
res$warnings %>% map_chr(~ .$message) %>% fct_count()
res2 <- capture_warnings(zeroinfl(as.formula(f_basic), data = m, dist="poisson"))
res2$warnings %>% map_chr(~ .$message) %>% fct_count()
```

Full data.

```{r}
filename <- file.path(model_path, "zip2_full.rds")
if (! file.exists(filename) || recompute) {
  tic("zip2_full")
  if (use_expanded_matrix)
    fit_zip2_full <- zeroinfl(as.formula(f_basic), data = m_full, dist="poisson")
  else
    fit_zip2_full <- zeroinfl(as.formula(f5v), data = full_scaled, dist="poisson")
  toc()
  saveRDS(fit_zip2_full, filename)
} else {
  fit_zip2_full <- readRDS(filename)
}
summary(fit_zip2_full)
```

```{r}
rootogram(fit_zip2_full)
```

Following does not work. Member family is not part of zero-inflated model.

```{r}
#y_ggdiagnostics(fit_zip2_full)
```

Original full data. Check that the result is same as above

```{r, eval=FALSE}
filename <- file.path(model_path, "zip2_full_orig.rds")
if (! file.exists(filename) || recompute) {
  tic("zip2_full_orig")
  if (use_expanded_matrix)
    fit_zip2_full_orig <- zeroinfl(n ~ ., data = m_full %>% select(-c(n_all, n_window, n_full_window)), dist="poisson")
  else
    fit_zip2_full_orig <- zeroinfl(as.formula(f5v), data = full_scaled, dist="poisson")
  toc()
  saveRDS(fit_zip2_full_orig, filename)
} else {
  fit_zip2_full_orig <- readRDS(filename)
}
summary(fit_zip2_full_orig)
```

The result looks the same.

Venesection count in the 2-year window:

```{r, eval=FALSE}
filename <- file.path(model_path, "zip2_full_window.rds")
if (! file.exists(filename) || recompute) {
  tic("zip2_full_window")
  fit_zip2_full_window <- zeroinfl(n_window ~ ., data = m_full %>% select(-c(n_all, n, n_full_window)), dist="poisson")
  toc()
  saveRDS(fit_zip2_full_window, filename)
} else {
  fit_zip2_full_window <- readRDS(filename)
}
summary(fit_zip2_full_window)
```

Venesection count in the strict 2-year window. The window must not contain the eof date.

```{r, eval=FALSE}
# Note, here I should do the scaling of age differently, since I filter out individuals.
filename <- file.path(model_path, "zip2_full_full_window.rds")
if (! file.exists(filename) || recompute) {
  tic("zip2_full_full_window")
  fit_zip2_full_full_window <- zeroinfl(n_full_window ~ ., 
                                        data = m_full %>% select(-c(n_all, n, n_window)) %>% replace_na(list(n_full_window=0)) %>% filter(n_full_window >= 0), 
                                        dist="poisson")
  toc()
  saveRDS(fit_zip2_full_full_window, filename)
} else {
  fit_zip2_full_full_window <- readRDS(filename)
}
summary(fit_zip2_full_full_window)
```

#### ZINB2

Train data.

Warning: glm.fit: fitted probabilities numerically 0 or 1 occurredWarning: glm.fit: fitted rates numerically 0 occurredWarning: system is computationally singular: reciprocal condition number = 9.85978e-19FALSE

```{r}
filename <- file.path(model_path, "zinb2.rds")
if (! file.exists(filename) || recompute) {
  tic("zinb2")
  if (use_expanded_matrix)
    fit_zinb2 <- zeroinfl(as.formula(f_basic), data = m, dist="negbin")
  else
    fit_zinb2 <- zeroinfl(as.formula(f5v), data = train_scaled, dist="negbin")
  toc()
  saveRDS(fit_zinb2, filename)
} else {
  fit_zinb2 <- readRDS(filename)
}
summary(fit_zinb2)
```

Full data.

```{r}
filename <- file.path(model_path, "zinb2_full.rds")
if (! file.exists(filename) || recompute) {
  tic("zinb2_full")
  if (use_expanded_matrix) {
    fit_zinb2_full <- zeroinfl(n ~ ., data = m_full %>% select(-c(n_all, n_window, n_full_window)), dist="negbin")
    #fit_zinb2_full <- zeroinfl(as.formula(f_basic), data = m_full, dist="negbin")
  } else
    fit_zinb2_full <- zeroinfl(as.formula(f5v), data = full_scaled, dist="negbin")
  toc()
  saveRDS(fit_zinb2_full, filename)
} else {
  fit_zinb2_full <- readRDS(filename)
}
summary(fit_zinb2_full)
```

### Fit with exposure

In the exposure model, instead of counts we model the incidents in a time unit.
Here exposure starts from the diagnosis of hemochromatosis and will continue until the end of follow-up.
The exposure time (in years) is individual specific.

Exposure and offset:
https://en.wikipedia.org/wiki/Poisson_regression#%22Exposure%22_and_offset
Incidence rate ratio, exposure:
https://stats.stackexchange.com/a/17010

```{r}
# variables_exposure <- c("sex2", "age", "donations_in_five_years", 
#                 #"H1_2", # is correlated with C282Y
#                 "H63D", "S65C", "C282Y", 
#                 "supersnip",
#                 "credible1", "credible2", 
#                 "credible3",
#                 "rs3094093", "rs114179634", #, "rs9261354"  # HLA snips
#                 # Interactions
#                 "C282Y:H63D", "C282Y:S65C", "C282Y:supersnip", "C282Y:donations_in_five_years", "C282Y:sex2"
#                 )
# Number of donations causes problems, I will drop them here.
variables_exposure <- c("sex2", "age", #"donations_in_five_years", 
                "H63D", "S65C", "C282Y", 
                "supersnip",
                "credible1", "credible2", 
                "credible3",
                "rs3094093", "rs114179634", #, #"rs9261354"  # HLA snips
                # Interactions
                "C282Y:H63D" #"C282Y:S65C", "C282Y:supersnip", "C282Y:donations_in_five_years", "C282Y:sex2"
                )
f_exposure2 <- sprintf("n ~ %s", paste(variables_exposure, collapse=" + "))
f_exposure2
```

Some levels combinations in interactions are missing. Hence, I need to convert to model.matrix and then preprocess the data to
exclude constant columns.


```{r}
sprintf("Exposure data has %i individuals\n", nrow(combined_with_exposure)) %>% cat()
```

I also drop the interaction C282Y1:donations_in_five_years, since it gets huge std error in all models.

```{r}
combined_with_exposure_scaled <- combined_with_exposure %>% 
  mutate(age = as.numeric(scale(age)),
         across(c(H63D, C282Y), as.factor))
```

```{r}
m_exposure <- my_expand(f_exposure2, combined_with_exposure_scaled)
m_exposure %>% mutate(across(-c(age, n), as.factor)) %>% summary()
```

Final formula:

```{r}
expanded_variables <- setdiff(colnames(m_exposure), responses)
f_exposure <- sprintf("n ~ %s", expanded_variables %>% paste(collapse = " + "))
f_exposure
```

Drop unneeded data.


#### ZIP2 

```{r}
filename <- file.path(model_path, "zip2_exposure.rds")
if (! file.exists(filename) || recompute) {
  tic("zip2_exposure")
  fit_zip2_exposure <- zeroinfl(as.formula(f_exposure), offset = log(combined_with_exposure_scaled$exposure), data = m_exposure, dist="poisson")
  toc()
  saveRDS(fit_zip2_exposure, filename)
} else {
  fit_zip2_exposure <- readRDS(filename)
}
summary(fit_zip2_exposure)
```

#### ZINB2

```{r}
filename <- file.path(model_path, "zinb2_exposure.rds")
if (! file.exists(filename) || recompute) {
  tic("zinb2_exposure")
  fit_zinb2_exposure <- zeroinfl(as.formula(f_exposure), offset = log(combined_with_exposure_scaled$exposure), data = m_exposure, dist="negbin")
  toc()
  saveRDS(fit_zinb2_exposure, filename)
} else {
  fit_zinb2_exposure <- readRDS(filename)
}
summary(fit_zinb2_exposure)
```

Theta is quite far away from 1.

#### Confidence intervals

```{r}
# Count components
cis_zip2_exposure    <- cis_helper(fit_zip2_exposure,    "cis_zip2_exposure.tsv",  model="count")
cis_zinb2_exposure   <- cis_helper(fit_zinb2_exposure,   "cis_zinb2_exposure.tsv", model="count")

# Zero components
cis_zero_zip2_exposure    <- cis_helper(fit_zip2_exposure,         "cis_zero_zip2_exposure.tsv",  model="zero")
cis_zero_zinb2_exposure   <- cis_helper(fit_zinb2_exposure,        "cis_zero_zinb2_exposure.tsv", model="zero")
```

```{r}
g1 <- bind_rows(zip2 = cis_zip2_exposure, zinb2 = cis_zinb2_exposure, .id = "model") %>%
  plot_cis2("Count component") + 
  labs(x="Incidence rate ratio", color="Model") + theme(axis.title.y = element_blank()) +
  guides(colour = guide_legend(reverse = TRUE, override.aes = list(shape = 16)))
g2 <- bind_rows(zip2 = cis_zero_zip2_exposure, zinb2 = cis_zero_zinb2_exposure, .id = "model") %>%
  plot_cis2("Zero component") + 
  labs(x="Incidence rate ratio", color="Model") + theme(axis.title.y = element_blank(), axis.text.y = element_blank()) +
  guides(colour = guide_legend(reverse = TRUE, override.aes = list(shape = 16)))
g1 + g2 + plot_layout(guides = "collect") + plot_annotation(title = "Exposure model")
```

Don't really know how to interpret these. Why is the effect of C282Y negative in the count component?
Not many significant effect, if we consider zinb2 to be the correct model.

#### Diagnostics

```{r}
g1 <- get_qqplot("zip2_exposure", m_exposure)
g2 <- get_qqplot("zinb2_exposure", m_exposure)
g1 + g2
```

Poisson version is really bad!

```{r}
g1 <- get_rootogram("zip2_exposure", m_exposure)
g2 <- get_rootogram("zinb2_exposure", m_exposure)
g1 + g2
```

### Score model for nb

Idea here was to combine several genetic variables into one to increase power and make
confidence intervals narrower.

```{r}
score_components <-  c(
                "H63D", "S65C", "C282Y", 
                "supersnip",
                "credible1",
                "credible3",
                "rs3094093", "rs114179634" )  # HLA snips
full_genetic_model <- TRUE
if (full_genetic_model) {
  score_components <- c(setdiff(score_components, c("H63D",  "C282Y")),
                        paste0("H63D", 1:2),
                        paste0("C282Y", 1:2))
}
variables_score <- c("sex2", "age", "donations_in_five_years", "genetic_score")

f_score <- sprintf("n ~ %s", paste(variables_score, collapse=" + "))
f_score
```

Form score as the weighted sum of all genetic predictors.

```{r, eval=FALSE}
M <- model.matrix(fit_nb)[,score_components]
v <- coef(fit_nb)[score_components]
score <- as.numeric(M %*% v)
cases_score <- cases_scaled %>% mutate(genetic_score = score)
```

```{r, eval=FALSE}
fit_nb_score <- MASS::glm.nb(as.formula(f_score), cases_score, control=glm.control(maxit=200))  # Default number of iterations is 25
summary(fit_nb_score)
```


```{r, eval=FALSE}
cis_nb_score <- compute_cis(fit_nb_score)
cis_nb_score %>%
  plot_cis("Regression of venesection counts", subtitle=sprintf("n=%i", nrow(cases))) +
  labs(x="Incidence ratio") + #theme_minimal() + 
  scale_x_log10() #+ theme_classic() + lims(x=c(0.25, 30))
```

### Read models

```{r}
model_ids <- c("poisson", "poisson_full", "nb", "nb_full", "zip", "zip_full",
                 "zinb", "zinb_full", "zip2", "zip2_full", "zinb2", "zinb2_full")
get_model <- function(model_id) {
  model_filename <- sprintf("%s.rds", model_id)
  model_name <- sprintf("fit_%s", model_id)
  cat(sprintf("Read model %s from file %s\n", model_name, model_filename))
  mod <- readRDS(file.path(model_path, model_filename))
  assign(model_name, mod, envir = rlang::global_env())
}
```

```{r, eval=FALSE}
walk(model_ids, get_model)
```

### Diagnostics

The following two are extremely slow.

```{r, eval=FALSE}
filename <- file.path(figure_path, "check_poisson.pdf")
tic("check_poisson")
check_poisson_g <- fit_poisson %>% performance::check_model() %>% plot()  # Requires the 'see' package
toc()
if (save_figs)
  ggsave(filename, check_poisson_g)
#check_poisson_g
```

Error in loadNamespace(j <- i[[1L]], c(lib.loc, .libPaths()), versionCheck = vI[[j]]) : 
  namespace ‘insight’ 1.2.0 is already loaded, but >= 1.4.1 is required
  
Why is this? 

> packageVersion("insight")
[1] ‘1.4.2’

```{r, eval=FALSE}
filename <- file.path(figure_path, "check_nb.pdf")
check_nb_g <- fit_nb %>% performance::check_model() %>% plot()
if (save_figs)
  ggsave(filename, check_nb_g)
```

#### qqplots

To make thing faster and use less memory I only take a small sample of the full data.

```{r}
set.seed(123)
new_data <- m %>% slice_sample(n=10000)
```

```{r}
new_data %>% pull(n) %>% table()
```






```{r, eval=FALSE}
#get_qqplot("zip2", new_data)
qqplots <- map(model_ids, function(name) get_qqplot(name, new_data))
g_qq <- reduce(qqplots, `+`)   # Combine all plots with patchwork
filename <- file.path(figure_path, "qqplots_2year_strict.pdf")
ggsave(filename, g_qq, width = 7, height = 7)
g_qq
```

#### Rootograms



```{r, eval=FALSE}
#new_data <- m %>% slice_sample(n=1000)
#get_rootogram("nb", new_data)
rootograms <- model_ids %>% #discard(~ str_starts(., "nb")) %>% 
  set_names() %>% map(function(name) get_rootogram(name, new_data))
g_root <- reduce(rootograms, `+`)   # Combine all plots with patchwork
filename <- file.path(figure_path, "rootograms_2year_strict.pdf")
ggsave(filename, g_root, width = 7, height = 7)
g_root
```

The negative binomial models distributions have very long tail!

### Glance

```{r}
dispersion <- function(fit) {
  sum(residuals(fit, type = "pearson")^2) / df.residual(fit)
}
# Computes the maximum standard error of the predictor coefficients
maximum_se <-function(fit) max(sqrt(diag(vcov(fit))))
```

```{r}
compute_stats <- function(fit) {
  n <- if ("glm" %in% class(fit)) length(fit[["y"]]) else fit[["n"]] 
  tibble(df.null=summary(fit)$df.null, logLik=as.numeric(logLik(fit)), AIC=AIC(fit), BIC=BIC(fit),
         df.residual=df.residual(fit), nobs=n, converged=fit$converged)
  
}
filename <- file.path(table_path, "progression_model_stats.tsv")
if (recompute || ! file.exists(filename)) {
  models <- list(
    poisson = fit_poisson, 
    nb = fit_nb, 
    zip = fit_zip, 
    zinb = fit_zinb, 
    zip2 = fit_zip2, 
    zinb2 = fit_zinb2, 
    
    poisson_full = fit_poisson_full, 
    nb_full = fit_nb_full, 
    zip_full = fit_zip_full, 
    zinb_full = fit_zinb_full, 
    zip2_full = fit_zip2_full, 
    zinb2_full = fit_zinb2_full)
  stats <- map(models, compute_stats) %>% bind_rows(.id = "model")
  stats <- stats %>% mutate(dispersion = map_dbl(models, dispersion),
                            max_se = map_dbl(models, maximum_se))
  unrealistic_values <- map(models, exact_zeros_or_ones) %>% bind_rows(.id = "model")
  stats <- stats %>% inner_join(unrealistic_values, by="model")
  #if (save_tables)
  write_tsv(stats, filename)
} else {
  stats <- read_tsv(filename)
}
stats
```

```{r}
rm(models)
```

All model fitted on train data failed and those trained on full data succeeded (with the exception of zinb_full).


Check zero inflation.

```{r}
performance::check_zeroinflation(fit_poisson)
```

```{r}
performance::check_zeroinflation(fit_nb)
```

Check overdispersion.

```{r}
performance::check_overdispersion(fit_poisson)
```

```{r, eval=FALSE}
performance::check_overdispersion(fit_nb)
```

These functions don't work for zero-inflated models!

What does this mean?

https://stats.stackexchange.com/questions/230730/negative-binomial-for-underdispersed-data

### Confidence intervals

```{r}
cis_helper <- function(fit, filename2, ...) {
  tryCatch(
    error = function(cnd) {message(conditionMessage(cnd)); NULL},
    {
      filename = file.path(table_path, filename2)
      message(sprintf("Filename is %s", filename))
      if (! file.exists(filename) || recompute) {
        message("recomputing")
        cis <- compute_cis(fit, ...)
        write_tsv(cis, filename)
        cis
      } else{
        message(paste("Loading from file", filename))
        read_tsv(filename, col_types = cols(
          term = col_character(),
          conf.low = col_double(),
          conf.high = col_double(),
          estimate = col_double()
        ))
      }
    }
  )
}
```


```{r}
cis_poisson      <- cis_helper(fit_poisson,      "cis_poisson.tsv")   # failed, glm.fit: fitted rates numerically 0 occurredError: no valid set of coefficients has been found: please supply starting values
cis_poisson_full <- cis_helper(fit_poisson_full, "cis_poisson_full.tsv")
cis_nb           <- cis_helper(fit_nb,           "cis_nb.tsv")       # Warning: glm.fit: algorithm did not converge
cis_nb_full      <- cis_helper(fit_nb_full,      "cis_nb_full.tsv")
cis_zip          <- cis_helper(fit_zip,          "cis_zip.tsv")
cis_zip_full     <- cis_helper(fit_zip_full,     "cis_zip_full.tsv")
#cis_zinb         <- cis_helper(fit_zinb,         "cis_zinb.tsv")  # slow
cis_zinb_full    <- cis_helper(fit_zinb_full,    "cis_zinb_full.tsv")  # slow
# Count components
cis_zip2         <- cis_helper(fit_zip2,         "cis_zip2.tsv",       model="count")
cis_zip2_full    <- cis_helper(fit_zip2_full,    "cis_zip2_full.tsv",  model="count")
#cis_zinb2       <- cis_helper(fit_zinb2,        "cis_zinb2.tsv",      model="count")
cis_zinb2_full   <- cis_helper(fit_zinb2_full,   "cis_zinb2_full.tsv", model="count")

# Zero components
cis_zero_zip2    <- cis_helper(fit_zip2,         "cis_zero_zip2.tsv",  model="zero")
cis_zero_zinb2   <- cis_helper(fit_zinb2,        "cis_zero_zinb2.tsv", model="zero")
cis_zero_zip2_full    <- cis_helper(fit_zip2_full,         "cis_zero_zip2_full.tsv",  model="zero")
cis_zero_zinb2_full   <- cis_helper(fit_zinb2_full,        "cis_zero_zinb2_full.tsv", model="zero")

# Alternative versions of models
cis_zip2_full_window         <- cis_helper(fit_zip2_full_window,         "cis_zip2_full_window.tsv",  model="count")
cis_zip2_full_full_window    <- cis_helper(fit_zip2_full_full_window,    "cis_zip2_full_full_window.tsv",  model="count")

```

Compare model zip2 on train data and full data.

```{r}
cis_tmp <- bind_rows(train = cis_zip2, full = cis_zip2_full, .id="model") %>% mutate(model = factor(model))
cis_tmp %>% plot_cis2("zip2 comparison") + labs(x="Incidence ratio", color="Data") + theme(axis.title.y = element_blank())
rm(cis_tmp)
```

Compare three definitions of response for model zip2 on full data.

```{r, eval=FALSE}
#cis_zip2_full_old <- read_tsv(file.path(table_path, "cis_zip2_full_old.tsv"))
cis_tmp3 <- bind_rows(n = cis_zip2_full, n_window = cis_zip2_full_window,  n_full_window = cis_zip2_full_full_window, 
                      .id="model") %>% mutate(model = factor(model, levels = c("n", "n_window", "n_full_window")))
cis_tmp3 %>%
  mutate(model = fct_rev(model)) %>%
  plot_cis2("zip2 comparison") + 
  labs(x="Incidence ratio", color="Data") + 
  #scale_color_discrete(breaks = c("old", "new")) +
  guides(colour = guide_legend(reverse = TRUE, override.aes = list(shape = 16))) +
  theme(axis.title.y = element_blank())
#rm(cis_tmp3)
```

Compare just the two definitions of response for model zip2 on full data.

```{r, eval=FALSE}
cis_tmp3 %>%
  filter(model != "n_full_window") %>%
  droplevels() %>%
  mutate(model = fct_rev(model)) %>%
  plot_cis2("zip2 comparison") + 
  labs(x="Incidence ratio", color="Data") + 
  #scale_color_discrete(breaks = c("old", "new")) +
  guides(colour = guide_legend(reverse = TRUE, override.aes = list(shape = 16))) +
  theme(axis.title.y = element_blank())
```

Compare old and new model zip2 on full data.

```{r, eval=FALSE}
cis_zip2_full_old <- read_tsv(file.path(table_path, "cis_zip2_full_old.tsv"), col_types = cols(
  term = col_character(),
  conf.low = col_double(),
  conf.high = col_double(),
  estimate = col_double()
))
cis_tmp2 <- bind_rows(old = cis_zip2_full_old, new = cis_zip2_full, .id="model") %>% mutate(model = factor(model))
cis_tmp2 %>% plot_cis2("zip2 comparison") + 
  labs(x="Incidence ratio", color="Data") + 
  scale_color_discrete(breaks = c("old", "new")) +
  theme(axis.title.y = element_blank())
rm(cis_tmp2)
```

```{r}
mylevels <- c("poisson", "nb", "zip", "zinb", "zip2", "zinb2",
              "poisson_full", "nb_full", "zip_full", "zinb_full", "zip2_full", "zinb2_full")
cis_count <- bind_rows(#poisson=cis_poisson, nb=cis_nb, 
                       poisson_full=cis_poisson_full, nb_full=cis_nb_full, 
                       zip_full=cis_zip_full, zinb_full=cis_zinb_full, 
                       #zip2=cis_zip2, zinb2=cis_zinb2, 
                       zip2_full=cis_zip2_full, zinb2_full=cis_zinb2_full, 
                       .id="model") %>%
  mutate(model = factor(model, levels=mylevels))
if (save_tables) {
  filename <- file.path(table_path, "full_models_forest.tsv")
  write_tsv(cis_count, filename)
}
```

```{r}
g <- cis_count %>%
  #filter(term == "age") %>% # REMOVE THIS LINE!!!!!!!!!!!!!!!!!!!!!!
  mutate(model = fct_drop(model)) %>% # Drop unused levels
  mutate(model = fct_rev(model)) %>%
  plot_cis2("Regression of venesection counts (count component)", subtitle=sprintf("n=%i", nrow(m_full)), na.rm = FALSE) +
  #geom_point(data = tibble(x=c(0.3, 0.5, 1.5, 2), y=1.5), mapping=aes(x, y), inherit.aes = FALSE) +
  labs(x="Incidence ratio") + #theme_minimal() + 
  #facet_wrap(~ model) +
  #scale_color_discrete(breaks=mylevels) +
  #scale_fill_discrete(breaks=mylevels) +
  guides(colour = guide_legend(reverse = TRUE, override.aes = list(shape = 16))) +
  coord_cartesian(xlim = c(1e-2, 1e+2)) +
  #scale_color_manual(breaks=c("zip", "poisson", "nb"), labels="") +
  #scale_x_log10(minor_breaks=c(1.5, 2.0)) #+ theme_classic() + lims(x=c(0.25, 30))
  NULL
if (save_figs) {
  filename <- file.path(figure_path, "full_models_forest.pdf")
  ggsave(filename, g)
}
g
```

"If a Poisson model is fitted under overdispersion of the response, then the standard errors of the estimated coefficients are underestimated." 




```{r}
#cis_zero <- bind_rows(zip2=cis_zip2, #zinb2=cis_zinb2, 
cis_zero <- bind_rows(#poisson_full=cis_poisson_full, nb_full=cis_nb_full, 
                       #zip_full=cis_zip_full, zinb_full=cis_zinb_full, 
                       zip2_full=cis_zip2_full, zinb2_full=cis_zinb2_full, 
                      .id="model") %>%
  mutate(model = factor(model, levels=mylevels))
```

```{r}
cis_zero %>%
  mutate(model = fct_rev(model)) %>%
  plot_cis2("Regression of venesection counts (zero component)", subtitle=sprintf("n=%i", nrow(m_full))) +
  #geom_point(data = tibble(x=c(0.3, 0.5, 1.5, 2), y=1.5), mapping=aes(x, y), inherit.aes = FALSE) +
  labs(x="Incidence ratio") + #theme_minimal() + 
  #facet_wrap(~ model) +
  scale_color_discrete(breaks=mylevels) +
  #scale_color_manual(breaks=c("zip", "poisson", "nb"), labels="") +
  scale_x_log10(minor_breaks=c(1.5, 2.0)) #+ theme_classic() + lims(x=c(0.25, 30))
```

Both components in the same plot.

```{r}
cis <- bind_rows(count = cis_count, zero = cis_zero, .id="component")
```

```{r}
g <- cis %>%
  mutate(model = fct_rev(model)) %>%
  filter(model != "zinb") %>%
  plot_cis2("Regression of venesection counts", subtitle=sprintf("n=%i", nrow(m))) +
  #geom_point(data = tibble(x=c(0.3, 0.5, 1.5, 2), y=1.5), mapping=aes(x, y), inherit.aes = FALSE) +
  labs(x="Incidence ratio") + #theme_minimal() + 
  facet_wrap(~ component, scales = "free_x") +
  scale_color_discrete(breaks=mylevels) +
  #scale_x_log10(minor_breaks=c(1.5, 2.0)) + #+ theme_classic() + lims(x=c(0.25, 30))
  theme(legend.position="bottom")
if (save_figs) {
  filename <- file.path(figure_path, "progression_cis.pdf")
  ggsave(filename, g, width=7, heigh=7)
}
g
```

### Probabilities

#### Mixture model

These probabilities use both components of the zero-inflation model.

```{r}
get_probabilities <- function(fit) {
  max_count <- max(fit$y)
  if ("zeroinfl" %in% class(fit)) {
    tmp <- predict(fit, type="prob") %>% as_tibble()  # Includes both zero and count components
  } else if ("negbin" %in% class(fit)) {
    theta <- fit$theta
    tmp <- map(0:max_count, function(k) dnbinom(k, mu = fitted(fit), size = theta)) %>% bind_cols()
    colnames(tmp) <- 0:max_count
  } else {
    tmp <- map(0:max_count, function(k) dpois(k, lambda = fitted(fit))) %>% bind_cols()
    colnames(tmp) <- 0:max_count
  }
  tmp
}
```

```{r}
# Plots the probability distribution of counts for each individual.
# Specifying n chooses only the first n individuals
plot_probabilities <- function(fit, model, n = Inf) {
  max_n <- length(fit$y) # How many individuals in total
  max_count <- max(fit$y)
  tmp <- get_probabilities(fit)
  tmp2 <- tmp %>% mutate(id = 1:n()) %>% pivot_longer(cols=-id, names_to = "count", values_to = "probability") %>% mutate(count = as.numeric(count))
  tmp2 <- tmp2 %>% 
    filter(id <= n) %>% 
    mutate(id = as.factor(id)) 
  tmp2 %>% 
    ggplot(aes(count, probability, #color=id, 
               group=id)) + 
    (if (is.infinite(n)) geom_line(alpha = 0.1) else geom_line(aes(color = id))) +
    labs(y="probability", title=model, subtitle=sprintf("%i individuals", n_distinct(tmp2$id)))
}
```

All individuals. These seem to be taking too much memory, so I disabled them.

```{r, eval=FALSE}
g1 <- plot_probabilities(fit_poisson, "Poisson")
```


```{r, eval=FALSE}
#g2 <- plot_probabilities(fit_nb, "Negative binomial")
```


```{r, eval=FALSE}
g3 <- plot_probabilities(fit_zip, "Zero-inflated Poisson")
```


```{r, eval=FALSE}
g4 <- plot_probabilities(fit_zinb, "Zero-inflated negative binomial")
```


```{r, eval=FALSE}
g1 + #g2 + 
  g3 + g4 + plot_layout(guides = "collect")
```

First five individuals.

```{r}
g1 <- plot_probabilities(fit_poisson, "Poisson", n = 5)
#g2 <- plot_probabilities(fit_nb, "Negative binomial", n = 5)
g3 <- plot_probabilities(fit_zip, "Zero-inflated Poisson", n = 5)
g4 <- plot_probabilities(fit_zinb, "Zero-inflated negative binomial", n = 5)
g1 + #g2 + 
  g3 + g4 + plot_layout(guides = "collect")
```

```{r}
rm(g1, g2, g3, g4)
```

#### Zero model

The zinb2 fitting failed to get std errors, so no wonder the following plot looks a mess.

```{r}
zero_probabilities <- tibble(zip2 = predict(fit_zip2, type="zero"), zinb2 = predict(fit_zinb2, type="zero"))
g1 <- zero_probabilities %>% ggplot(aes(zip2, zinb2)) + 
  geom_abline(slope=1, intercept = 0, color="red") +
  geom_point() + labs(title="Zero model probabilities")
g2 <- zero_probabilities %>% 
  pivot_longer(cols=everything(), names_to="model", values_to="probability") %>% 
  ggplot(aes(probability, color=model)) + 
  geom_freqpoly() +
  labs(title="Zero model probability")
g1 + g2
```


### Counts

#### Expected counts

Observed zeros and expect number of zeros predicted by models.

```{r}
round(c("Observed" = sum(m$n < 1),
        "Poisson"  = sum(dpois(0, fitted(fit_poisson))),
#        "NB"   = sum(dnbinom(0, mu = fitted(fit_nb), size = fit_nb$theta)),
        "ZIP"  = sum(predict(fit_zip, type = "prob")[,1]),
        "ZINB" = sum(predict(fit_zinb, type = "prob")[,1]),
        "ZIP2"  = sum(predict(fit_zip2, type = "prob")[,1]),
        "ZINB2" = sum(predict(fit_zinb2, type = "prob")[,1])
        ))
```


```{r}
get_expected_counts <- function(fit) {
  tmp <- get_probabilities(fit)
  tmp2 <- tmp %>% 
    summarise(across(everything(), sum)) %>% 
    as.list() %>% enframe() %>% unnest(value) %>% 
    rename(count=name, expected_count=value) %>% 
    mutate(count = as.numeric(count))
  tmp2
}
tmp <- list(poisson=fit_poisson, #nb=fit_nb, 
            zip=fit_zip, zinb=fit_zinb, zip2=fit_zip2, zinb2=fit_zinb2) %>%
  map(get_expected_counts) %>% 
  bind_rows(.id="model")
tmp %>% pivot_wider(names_from=model, values_from=expected_count)
observed_counts <- m %>% count(n) %>% rename(count = n, expected_count = nn) %>% mutate(model = "observed")
expected_counts <- bind_rows(tmp, observed_counts) %>%
  mutate(model = factor(model, levels=c("observed", mylevels)))
expected_counts %>% 
  ggplot(aes(count, expected_count, color=model, group=model)) + 
  geom_line() + lims(x=c(0, 40)) +
  guides(color = guide_legend(override.aes = list(linewidth = 2) ) ) +
  labs(title="Expected number of counts")
```


#### Predicted counts

```{r}
result_poisson <- augment(fit_poisson, type.predict = "response") %>% select(n, poisson=.fitted)
#result_nb <- augment(fit_nb, type.predict = "response") %>% select(nb=.fitted)
#result_nb <- predict(fit_nb, type="response")
result_zip <- predict(fit_zip)
result_zinb <- predict(fit_zinb)
result <- bind_cols(result_poisson, #nb=result_nb, 
                    zip = result_zip, zinb = result_zinb)
result <- result %>% pivot_longer(cols=everything())
```

```{r}
tmp <- result %>% 
  mutate(name = factor(name, levels=c("poisson", "nb", "n", "zip", "zinb"))) %>%
  #mutate(value = round(value)) %>%
  mutate(value = as.integer(value))
tmp %>%
  ggplot(aes(value)) + 
  #geom_histogram(position="dodge") + 
  geom_bar(position="dodge") + 
  lims(x=c(-1, 50)) +   # For some reason, if lower limit is zero, then the zero isn't shown
  #scale_fill_discrete(labels=c("Predicted", "Observed")) + #, values=c(".fitted", "n")) +
  labs(x="Number of operations", fill=NULL) +
  facet_wrap(~name)
```

Note, that the above figure model counts tell, for example, for how many individuals count zero was predicted, which is not the same 
as the expected number of zeros.

```{r}
result %>% ggplot(aes(value, color=name)) + 
  #geom_histogram() + 
  geom_density(key_glyph = "path") +
  #scale_color_discrete(labels=c("Observed", "NB", "Poisson")) + #, values=c(".fitted", "n")) +
  labs(x="Number of operations", color=NULL) +
  #facet_wrap(~ name, scale="free_x") + 
  #scale_y_log10()
  NULL
```

After we removed the restriction to cases, this plot came more uninformative.











