---
title: "Hemochromatosis model in UKBB"
author: "Jarkko Toivonen"
date: "`r Sys.Date()`"
format: 
  html:
    df-print: kable
    toc: true
    toc-depth: 4
    embed-resources: true
---

## Setup

```{r, eval=FALSE}
install.packages("probably")
install.packages("biomaRt")
install.packages("ggh4x")
renv::snapshot()
```

```{r setup}
library(tidyverse)
library(tidymodels)
library(glue)
library(glmnet)
library(broom)
library(fmsb)  # for NagelkerkeR2
library(survival)
library(ciTools)
library(forestplot)
library(tableone)
library(probably)  # for calibration plots
library(statmod)   # for quantile residuals
library(patchwork)
library(splines)
library(survminer)

#ukbb_dir <- "/media/ukbb/UKBB"
qctool <- "~/bin/qctool"
bgenix <- "~ubuntu/progs/bgen/bin/bgenix"
result_path <- "../results"
figure_path <- file.path(result_path, "figures")
table_path <- file.path(result_path, "tables")

recompute <- FALSE
recompute_genotypes <- FALSE
save_figs <- FALSE
save_tables <- FALSE

#boot.n <- 2000  # Not needed
```

## Functions

```{r}
pretty_conv2 <- tibble::tribble(
     ~phenocode,    ~phenotype,
          "age",         "Event age",
     "assessment_age", "Baseline age",
     "event_age", "Diagnosis age",
     "hemochromatosis", "Hemochromatosis",
          "sex_female",         "Female",
          "sexfemale",         "Female",
     "sex", "Sex",
     "rs181949568", "Causal1 (rs181949568)",
  "rs114179634", "HLA1 (rs114179634)",
    "rs3094093",   "HLA2 (rs3094093)",
    "supersnip",   "HLA3 (rs9272324)",
         "H1_2",        "H1_2",
         "H63D",        "H63D",
         "S65C",        "S65C",
        "C282Y",       "C282Y",
        "C282Y_X1",       "C282Y heterozygote",
        "C282Y_X2",       "C282Y homozygote",
        "C282Y1",       "C282Y heterozygote",
        "C282Y2",       "C282Y homozygote",
         "H63D_X1",        "H63D heterozygote",
         "H63D_X2",        "H63D homozygote",
         "H63D1",        "H63D heterozygote",
         "H63D2",        "H63D homozygote",
         "H1_2_X1",        "H1-2 heterozygote",
         "H1_2_X2",        "H1-2 homozygote",
  ) %>% deframe()
# Form pretty interaction names
create_interactions <- function(x, y, input_sep="_x_", output_sep=" x ") {
  interactions <- bind_rows(   # Make interactions in both ways
    expand_grid(a=x, b=y),
    expand_grid(a=y, b=x)
  )
  interactions %>% 
    mutate(interaction = paste(a, b, sep=input_sep), pretty=paste(pretty_conv2[a], pretty_conv2[b], sep=output_sep)) %>% 
    select(interaction, pretty) %>%
    deframe()
}
# Tidymodels encoding
interactions1 <- create_interactions(c("C282Y_X1", "C282Y_X2"),
                                    c("H63D_X1", "H63D_X2", "S65C", "supersnip",
                                      "sex_female", "rs114179634", "rs3094093"),
                                    input_sep = "_x_",
                                    output_sep = " x ")
# Base R encoding
interactions2 <- create_interactions(c("C282Y1", "C282Y2"),
                                    c("H63D1", "H63D2", "S65C", "supersnip",
                                      "sexfemale", "rs114179634", "rs3094093"),
                                    input_sep = ":",
                                    output_sep = " x ")
pretty_conv <- c(pretty_conv2, interactions1, interactions2)
```

```{r}
model_stats <- function(fit) {
  #y <- fit[["fit"]][["fit"]][["fit"]][["data"]] %>% pull(..y) 
  y <- extract_fit_engine(fit)[["data"]] %>% pull(..y) 
  
  res <- fit %>% extract_fit_engine() %>% fmsb::NagelkerkeR2()
  sprintf("N=%i, prevalence=%.2f%%, R2=%.2f", res$N, mean(y == "case")*100, res$R2)
}
```

```{r}
compute_cis <- function(model) {
  cis <- model %>%
    tidy(conf.int = TRUE, exponentiate = TRUE) %>%
    filter(term != "(Intercept)")
  cis
}
```

```{r}
plot_cis <- function(cis, title, subtitle=NULL) {
  cis %>%
    mutate(term = fct_relevel(term, intersect(names(pretty_conv), term)),
           term = fct_relabel(term, ~ coalesce(pretty_conv[.x], .x)),
           term = fct_rev(term),
           hollow_color = if_else(conf.low < 1 & conf.high > 1, NA, "black")) %>%
    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, y=term, fill=I(hollow_color))) +
    geom_vline(xintercept = 1, color="gray") +
    geom_linerange() +
    geom_point(shape=21, size=3) +
    scale_fill_identity() +
    scale_x_log10() +
    #scale_fill_discrete(na.value = NA, guide="none") +
    labs(title=title, subtitle=subtitle, x="Odds ratio", y="Variable")
}
```

```{r}
fill_names <- function(v) {
  old_names <- names(v) %>% na_if("")
  new_names <- coalesce(old_names, unname(v))
  set_names(v, new_names)
}
swap_names_and_values <- function(v) {
  set_names(names(v), unname(v))
}
```

```{r}
is_wholenumber <- function(x, tol = .Machine$double.eps^0.5) abs(x - round(x)) < tol
get_fixed_predictors <- function(data, important_predictors) {
  helper2 <- function(value) {
    if (is.factor(value) || is_wholenumber(value) || is.logical(value)) as.character(value) else sprintf('%.2f', value)
  }
  helper <- function (value, name) {
    glue("{name}={map(value, helper2)}")
  }
  data %>%
    select(-all_of(important_predictors)) %>%
    distinct() %>%
    as.list() %>%
    imap(helper) %>%
    paste(collapse=" ")
}
```

```{r}
risk_table <- function(lf, data, important_predictors, filename, title=NULL, n=16) {
  variables <- important_predictors %>% unname()
  fit <- lf %>% extract_fit_engine()
  prediction_cis <- lf %>%
    extract_recipe() %>%
    bake(data) %>%
    ciTools::add_ci(fit, names = c("lower", "upper"), yhatName = "mean", alpha = 0.5) %>%
    mutate(type = "parametric")
  tmp <- bind_cols(data %>% select(all_of(variables)),
                   prediction_cis %>% select(mean, lower, upper))
  counts <- train %>% count(across(all_of(variables)))
  result <- tmp %>%
    left_join(counts, by=variables) %>%
    replace_na(list(n=0)) %>%
    filter(n >= 5) %>%
    arrange(desc(mean))
  if (save_figs) {
    pdf(filename, width=12, height=7, title=basename(filename))
  }
  result2 <- result %>%
    slice_max(order_by = mean, n=n) %>%
    mutate(probability = sprintf("%.3f", mean))
  important_predictors2 <- c(important_predictors,
                             Probability="probability", `Train count`="n")
  variables <- important_predictors2 %>% unname()
  header <- important_predictors2 %>% swap_names_and_values()
  rlang::inject(
    p <- result2 %>%
      forestplot(labeltext = !!variables,
                 boxsize = 0.2,
                 title = title,
                 new_page = FALSE,
                 xlab = "Hemochromatosis probability") %>%
      fp_add_lines() %>%
      fp_add_header(!!!header) %>%
      fp_set_zebra_style("#EFEFEF")
  )
  print(p)
  if (save_figs) dev.off()
  p
}
```

```{r}
compute_risk_cis <- function(lf, newdata, important_predictors, count_data = train, conf.level = 0.95) {
  variables <- important_predictors %>% unname()
  fit <- lf %>% extract_fit_engine()
  prediction_cis <- lf %>% 
    extract_recipe() %>%
    bake(newdata) %>% 
    ciTools::add_ci(fit, names = c("lower", "upper"), yhatName = "mean", alpha = 1 - conf.level) %>%
    mutate(type = "parametric")
  counts <- count_data %>% count(across(all_of(variables)))
  result <-
    bind_cols(
      newdata %>% select(all_of(variables)),
      prediction_cis %>% select(mean, lower, upper)) %>%
    left_join(counts, by=variables) %>%
    replace_na(list(n=0))
  result
}

compute_risk_table <- function(model, newdata, important_predictors ,count_data, col="sex2", conf.level = 0.95) {
  cis <- compute_risk_cis(model, newdata, important_predictors, count_data = count_data, conf.level = conf.level)
  tmp_variables <- setdiff(unname(important_predictors), c("sex2", "sex", "donation_in_five_years"))
  tmp2 <- cis %>% 
    group_by(across(all_of(tmp_variables))) %>%
    mutate(max_mean = max(mean), max_n = max(n)) %>%
    group_by(max_mean, max_n, .add=TRUE) %>% nest() %>% ungroup() %>%
    filter(max_n >= 10) %>%
    slice_max(order_by=max_mean, n=8) %>%
    unnest(data)
  if (col == "sex") {
    tmp2 <- tmp2 %>%
      mutate(value = sprintf("%.3f (%i)", mean, n))
  } else {
      tmp2 <- tmp2 %>%
        mutate(value = sprintf("%.3f (%s)", mean, if_else(donations_in_five_years == 0, as.character(n), "NA")))
  }
  labels <- tmp2 %>%
    select(-c(max_mean, max_n, mean, lower, upper, n)) %>%
    pivot_wider(names_from=starts_with("sex"), values_from=value)
  
  if (col == "sex2") {
    result <- tmp2 %>% inner_join(labels, by=setdiff(unname(important_predictors), c("sex2", "sex")))
  } else {
    result <- tmp2 %>% inner_join(labels, by=setdiff(unname(important_predictors), c("sex2", "sex", "donations_in_five_years")))
  }
  result %>% mutate(conf.level = conf.level)
  
}

```

```{r}
plot_risk_table <- function(result, important_predictors, title = NULL, col="sex2", conf.level) 
{
  max_x <- result %>% pull(upper) %>% max()
  max_x <- as.integer(max_x/0.05 + 1)*0.05
  if (col == "sex2") {
    important_predictors2 <- c(important_predictors, Male="male", Pre="pre", Post="post") %>%
      discard_at("Sex2")
    colors <- c("blue", "darkred", "green")
    stripes <- c("#ffffff", "#efefef", "#dfdfdf")
  } else {
    important_predictors2 <- c(important_predictors, Male="male", Female="female") %>%
      discard_at("Sex")
    colors <- c("blue", "darkred")
    stripes <- c("#ffffff",  "#dfdfdf")
  }
  variables <- important_predictors2 %>% unname()
  header <- important_predictors2 %>% swap_names_and_values()
  colgap <- convertX(unit(6, "mm"), "npc")
  xlab <- "Hemochromatosis probability"
  xticks <- seq(0, max_x, 0.05)
  confidence_line_color <- "black"
  txt_gp <- fpTxtGp(xlab = gpar(cex = 0.7),
                    ticks = gpar(cex = 0.6)
                    )
  txt_gp2 <- fpTxtGp(
      xlab = gpar(
        cex = 0.7,
        alpha = 0
      ),
      label = gpar(alpha = 0),
      legend = gpar(alpha = 0),
      ticks = gpar(alpha = 0, cex = 0.6)
      )
  boxsize <- 0.2
  rlang::inject(
    p1 <- result %>%
      filter(conf.level == .env$conf.level[[1]]) %>%
      group_by(across(all_of(col))) %>%
      forestplot(labeltext=!!variables,
                 title = title,
                 new_page = FALSE,
                 colgap = colgap,
                 xlab = xlab,
                 xticks = xticks,
                 txt_gp = txt_gp,
                 col = fpColors(lines = confidence_line_color),
                 lwd.ci = 1.5,
                 boxsize = boxsize) %>%
      fp_set_style(box = colors |> lapply(function(x) gpar(fill = x, col = "#555555")),
                   default = gpar(lineend = "square", vertices = TRUE)) %>%
      fp_add_header(!!!header) %>%
      fp_set_zebra_style(!!!stripes)
  )
  
  rlang::inject(
    p2 <- result %>%
      filter(conf.level == .env$conf.level[[2]]) %>%
      group_by(across(all_of(col))) %>%
      forestplot(labeltext=!!variables,
                 title = "",
                 new_page = FALSE,
                 colgap = colgap,
                 zero = NA,
                 xlab = " ",
                 xticks = xticks,
                 txt_gp = txt_gp2,
                 col = fpColors(zero = NA, lines = confidence_line_color),
                 lwd.ci = 6,
                 boxsize = boxsize) %>%
      fp_set_style(box = colors |> lapply(function(x) gpar(fill = x, col = "#555555")),
                   default = gpar(lineend = "square", vertices = TRUE)) %>%
      fp_add_header(!!!header) #%>%
      #fp_set_zebra_style(!!!stripes)
  )
  list(p1, p2)
}
```

## Load data

```{r}
basic <- readRDS("../results/basic.rds")
```

### Withdrawals

Get withdrawals.

```{r, eval=FALSE}
# This file was last changed 2023-09-11
filename <- file.path(ukbb_dir, "data/withdrawals/all_withdrawals.csv")
withdrawals <- read_tsv(filename, col_names = FALSE) %>% pull(X1)
length(withdrawals)
```











Only one credible snip was found and only 50 persons are heterozygotes.

## Form data

```{r}
filename <- "../results/basic_combined.rds"
if (!file.exists(filename) || recompute) {
  basic_combined <- basic %>%
    rename(H1_2="H1-2") %>%
    mutate(age = eof_age) %>%
    mutate(sex = factor(sex, levels=c("male", "female")),
           sex2 = factor(case_when(sex == "male" ~ "male",
                                   age <= 50 ~ "pre",
                                   age > 50  ~ "post"),
                         levels=c("male", "pre", "post")))
  saveRDS(basic_combined, filename)
} else {
  basic_combined <- readRDS(filename)
}
basic_combined <- basic_combined %>%
  mutate(survival_age = coalesce(event_age, eof_age))   # This is for the survival model
```

## Table 1

Create table 1.

```{r}
genetic_variables <- c("C282Y", "H63D", "S65C",
                
                "rs181949568",
                "H1_2",
                "rs114179634", "rs3094093", "supersnip")

vars <- c("hemochromatosis", "sex", "assessment_age", "event_age",
          genetic_variables)
renamer <- function(x) { coalesce(pretty_conv[x], x) %>% unname()}
#basic_combined <- basic_combined %>% rename(rs181949568 = credible1) # Korjaa tämä eksploraatiossa!!!!!
tmp <- basic_combined %>%
  mutate(across(c("rs181949568", "rs114179634"), function(v) { v %>% round()}),
         across("rs181949568", function(v) { v %>% as.factor() %>% fct_collapse(dominant = c("1", "2")) } ),
         hemochromatosis = fct_relevel(hemochromatosis, c("control", "case"))) %>%
  rename_with(renamer)
```

```{r}
continuous_variables <- c("Baseline age", "Diagnosis age")
tmp %>% 
  select(eid, Sex, all_of(continuous_variables)) %>% 
  pivot_longer(cols = all_of(continuous_variables)) %>%
  ggplot(aes(value)) +
  geom_histogram() +
  ggh4x::facet_grid2(Sex ~ name, scales = "free", independent = TRUE)
```

```{r}
tableone <- tableone::CreateTableOne(vars = vars %>% map_chr(renamer), 
                                     strata = "Sex", data = tmp,
                           factorVars = c("Hemochromatosis", genetic_variables) %>% map_chr(renamer),
                           addOverall = TRUE)
#rm(tmp)
# Save to a TSV file
if (save_tables) {
  filename <- file.path(table_path, "tableone_ukbb.tsv")
  tableone_matrix <- print(tableone, quote=FALSE, noSpaces = TRUE, printToggle = FALSE,
                           nonnormal = continuous_variables)
  write_tsv(tableone_matrix %>% as_tibble(rownames = "rownames"), file = filename)
}
print(tableone, nonnormal = continuous_variables)
```

```{r}
tmp %>% select(all_of(continuous_variables)) %>% summary()
```

How many non-NA values?

```{r}
tmp %>% select(all_of(continuous_variables)) %>% summarise(across(everything(), function(v) sum(!is.na(v))))
```

## Split

Divide the data into train and test sets.

```{r}
seed <- 142
set.seed(seed)
helper <- function(n) {
  v <- rep(c("train", "test"), length.out=n)
  sample(v, n)
}
use_tidymodels <- TRUE
if (use_tidymodels) {
  # See
  # https://github.com/tidymodels/rsample/issues/473
  basic_split <- basic_combined %>%
    initial_split(strata = "hemochromatosis", prop = 0.5) #, pool = 0.0010) # causes bias
  # set.seed(seed)
  # basic_split_wild <- basic_combined %>%
  #   filter(C282Y == 0) %>%
  #   initial_split(strata = "hemochromatosis", prop = 0.5, pool = 0.0010)
  train <- training(basic_split)
  test <- testing(basic_split)
  # train_wild <- training(basic_split_wild)
  # test_wild <- testing(basic_split_wild)
} else {
  tmp <- basic %>%
    group_by(hemochromatosis, sex) %>%
    mutate(group = helper(n())) %>%
    ungroup()
    
  train <- tmp %>% filter(group=="train") %>% select(-group)
  test <- tmp %>% filter(group=="test") %>% select(-group)
  rm(tmp)
}
```


## Logistic model 1




```{r}
known_snip_names <- c("H1_2",  "H63D",  "S65C", "C282Y", "supersnip")
hla_snip_names <- c("rs114179634",  #"rs9261354",  
                    "rs3094093")
credible_snip_names <- c("rs181949568",  "rs1284018747", "rs1300021441")
hla_snip_names2 <- intersect(colnames(basic_combined), hla_snip_names)
#credible_snip_names2 <- setdiff(colnames(credible_snips), "eid")
credible_snip_names2 <- intersect(colnames(basic_combined), credible_snip_names)
variables1 <- c("sex", "age",  "H1_2",  "H63D", "S65C", "C282Y", 
                # "H63D:C282Y", # Done in the recipe 
                "supersnip", "rs3094093", "rs114179634")
                #credible_snip_names2)
f1 <- sprintf("hemochromatosis ~ sex + age + %s",
             paste(variables1, collapse = " + "))
f1
```

```{r}
logistic_mod_1 <- parsnip::logistic_reg()
logistic_mod_1
```

```{r}
logistic_recipe_1 <-
  recipe(as.formula(f1), data=train) %>%
  #step_relevel(hemochromatosis, ref_level = "control") %>%
  step_normalize(age) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_interact(terms = ~ H63D:C282Y)
logistic_recipe_1
```

```{r}
logistic_wflow_1 <-
  workflow() %>%
  add_model(logistic_mod_1) %>%
  add_recipe(logistic_recipe_1)
logistic_wflow_1
```

```{r}
filename <- file.path(result_path, "logistic_lf_1.rds")
if (! file.exists(filename) || recompute) {
  logistic_fit_1 <- logistic_wflow_1 %>%
    fit(data = train)
  saveRDS(logistic_fit_1, filename)
} else {
  logistic_fit_1 <- readRDS(filename)
}
logistic_fit_1
```



```{r}
broom::glance(logistic_fit_1)
```

Variance inflation factors.

```{r}
logistic_fit_1 %>% extract_fit_engine() %>% car::vif() %>% sort(decreasing = TRUE)
```

```{r}
logistic_fit_1 %>% extract_fit_engine() %>% plot()
```

```{r}
logistic_fit_1 %>% extract_fit_engine() %>% fmsb::NagelkerkeR2()
```

```{r}
filename <- file.path(table_path, "logistic_cis_1.tsv")
if (! file.exists(filename) || recompute) {
  logistic_cis_1 <- compute_cis(logistic_fit_1)
  write_tsv(logistic_cis_1, filename)
} else
  logistic_cis_1 <- read_tsv(filename)
```




```{r}
filename <- file.path(figure_path, "logistic_cis_1.pdf")
g <- plot_cis(logistic_cis_1, "Logistic model 1", subtitle=model_stats(logistic_fit_1))
ggsave(filename, g)
g
```

Calibration.

```{r}
logistic_fit_1_test_pred <- augment(logistic_fit_1, new_data = test)
logistic_fit_1_test_pred %>%
  cal_plot_windowed(truth = hemochromatosis, estimate = .pred_case, step_size = 0.025, event_level = "second")
```

Try with quantile residuals instead.

```{r}
constant_information_transformation <- function(fit) {
  family <- fit$family$family
  # Constant-information scale transformation of the linear predictor.
  # See page 308 of Generalized linear models with examples in R.
  transformation <- if (family == "binomial") {
    function(x) asin(sqrt(x))
  } else if (family == "poisson") {
    function(x) sqrt(x)
  } else if (family == "gamma") {
    function(x) log(x)
  } else {
    function(x) { stop("not implemented")}
  }
}
my_diagnostics <- function(fit) {
  transformation <- constant_information_transformation(fit)
  par(mfrow = c(2, 2))
  qr <- qresid(fit)
  qqnorm(qr, las=1); qqline(qr)
  plot(qr ~ asin(sqrt(fitted(fit))), las=1)   # sin^-1(sqrt(mu)) is the constant-information scale transformation for Binomial distribution    
  plot(cooks.distance(fit), type="h", las=1)
  plot(hatvalues(fit), type="h", las=1)
}
fit <- logistic_fit_1 %>% extract_fit_engine()
my_diagnostics(fit)
```

```{r}
my_diagnostics2 <- function(fit) {
  transformation <- constant_information_transformation(fit)
  par(mfrow = c(1, 2))
  use_deviance_residuals <- TRUE
  if (use_deviance_residuals) {
    label <- "Standardized deviance residuals"
    standardized_residuals <- rstandard(fit)
  } else {
    label <- "Standardized quantile residuals"
    standardized_residuals <- qresid(fit) / sqrt(1 - hatvalues(fit))
  }
  scatter.smooth(standardized_residuals ~ transformation(fitted(fit)), las=1,
                 ylab=label,
                 xlab="Constant information fitted")
  z <- resid(fit, type="working") + fit$linear.predictors
  plot(z ~ fit$linear.predictors, las=1, xlab="Working responses",
       ylab="Linear predictor")
  abline(0, 1)
}
my_diagnostics2(fit)
```


```{r}
my_ggdiagnostics <- function(fit) {
  transformation <- constant_information_transformation(fit)
  #par(mfrow = c(2, 2))
  qr <- qresid(fit)
  #qqnorm(qr, las=1); qqline(qr)
  use_deviance_residuals <- TRUE
  if (use_deviance_residuals) {
    label <- "Standardized\ndeviance residuals"
    standardized_residuals <- rstandard(fit)
  } else {
    label <- "Standardized\nquantile residuals"
    standardized_residuals <- qresid(fit) / sqrt(1 - hatvalues(fit))
  }
  df <- tibble(qr = qr, 
               standardized_residuals = standardized_residuals,
               cit_fitted = transformation(fitted(fit)),
               linear_predictor = fit$linear.predictors,
               cook = cooks.distance(fit),
               hat  = hatvalues(fit),
               z = resid(fit, type="working") + fit$linear.predictors,   # Working responses
               index = 1:length(qr))
  g1 <- df %>% ggplot(aes(sample=qr)) + 
    geom_abline(slope=1, intercept = 0, color="red") +
    geom_qq() + labs(x="Theoretical quantiles", y="Sample quantiles")
  #plot(qr ~ asin(sqrt(fitted(fit))), las=1)   # sin^-1(sqrt(mu)) is the constant-information scale transformation for Binomial distribution    
  g2 <- df %>%
    ggplot(aes(x=cit_fitted, y=qr)) +
    geom_point() + geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs")) +
    labs(x="Constant information fitted", y="Quantile residual")
  g3 <- df %>% 
    ggplot(aes(x=index, y=cook, xend=index, yend=0)) +
    geom_segment() + labs(x="Index", y="Cook's distance")
  g4 <- df %>% 
    ggplot(aes(x=index, y=hat, xend=index, yend=0)) +
    geom_segment() + labs(x="Index", y="Hatvalue")
  
  # Why does the fitted curve look different from the one created by base graphics?
  g5 <- df %>%
    ggplot(aes(x=cit_fitted, y=standardized_residuals)) +
    geom_point() + 
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs")) +
    #geom_smooth(method = "loess", span=2/3, method.args=list(span = 2/3, degree = 1, family="gaussian"), se=FALSE) +
    labs(y=label, x = "Constant information fitted")
  # scatter.smooth(standardized_residuals ~ transformation(fitted(fit)), las=1,
  #                ylab=label,
  #                xlab="Constant information fitted")
  
  g6 <- df %>%
    ggplot(aes(x=linear_predictor, y=z)) +
    geom_point() + geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs")) +
    labs(x="Working responses", y="Linear predictor")
  # plot(z ~ fit$linear.predictors, las=1, xlab="Working responses",
  #      ylab="Linear predictor")
  # abline(0, 1)
  #plot(cooks.distance(fit), type="h", las=1)
  #plot(hatvalues(fit), type="h", las=1)
  (g1 + g2) / (g3 + g4) / (g5 + g6)
}
```


```{r}
my_ggdiagnostics(fit)
```

```{r}
variable_plots <- function(fit, variables="age") {
  df1 <- fit$data %>% select(all_of(variables)) %>%
    mutate(y = rstandard(fit)) %>% 
    pivot_longer(cols = variables, values_to="x", names_to = "variable")
  # df1 <- tibble(standardized_residuals = rstandard(fit),
  #               value = fit$data$age)
  # g1 <- df1 %>% ggplot(aes(x=value, y=standardized_residuals)) + geom_point() +
  #   labs(x="age", y="Standardized deviance residuals")
  # Partial residual plot
  # This probably does not make sense when model contains interactions
  tmp <- termplot(fit, data=fit$data, 
                  partial.resid = TRUE, terms = variables, las=1, plot = FALSE)
  df2 <- bind_rows(tmp, .id="variable") %>% as_tibble()
  df <- bind_rows(full=df1, partial=df2, .id="type")
  #g2 <- df2 %>% ggplot(aes(x, y)) + geom_point() + labs(x="age")
  g <- df %>% ggplot(aes(x, y)) + geom_point() +
    labs(x="value", y="residual") +
    ggh4x::facet_grid2(rows=vars(variable), cols=vars(type),
                       #variable ~ type, 
                       scales="free", independent=TRUE) 
  #g1 + g2
  g
}
```

```{r}
variable_plots(fit)
```

## Model 5 C282Y wild only


```{r}
variables5_wild <- c("sex", "age",
                "H63D", "S65C",
                "C282Y",
                "H1_2",
                "supersnip",
                #"credible1",
                "rs3094093", "rs114179634"
                )
f5_wild <- sprintf("hemochromatosis ~ %s",
             paste(variables5_wild, collapse = " + "))
f5_wild
```


```{r}
logistic_mod_5 <- parsnip::logistic_reg()
logistic_mod_5
```

```{r}
logistic_recipe_5_wild <-
  recipe(as.formula(f5_wild), data=train) %>%
  step_filter(C282Y == 0) %>%
  step_rm(C282Y) %>%
  #step_relevel(hemochromatosis, ref_level = "control") %>%
  step_normalize(age) %>%
  step_mutate(H1_2 = as.factor(round(H1_2)),
              H63D = as.factor(round(H63D))) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_lincomb(all_numeric_predictors())
logistic_recipe_5_wild
```

```{r}
logistic_wflow_5_wild <-
  workflow() %>%
  add_model(logistic_mod_5) %>%
  add_recipe(logistic_recipe_5_wild)
logistic_wflow_5_wild
```

```{r}
filename <- file.path(result_path, "logistic_lf_5_wild.rds")
if (! file.exists(filename) || recompute) {
  logistic_fit_5_wild <- logistic_wflow_5_wild %>%
    fit(data = train)
  saveRDS(logistic_fit_5_wild, filename)
} else {
  logistic_fit_5_wild <- readRDS(filename)
}
logistic_fit_5_wild
```

```{r}
broom::glance(logistic_fit_5_wild)
```

Variance inflation factors.

```{r}
logistic_fit_5_wild %>% extract_fit_engine() %>% car::vif() %>% sort(decreasing = TRUE)
```

```{r}
logistic_fit_5_wild %>% extract_fit_engine() %>% plot()
```

```{r}
logistic_fit_5_wild %>% extract_fit_engine() %>% fmsb::NagelkerkeR2()
```


```{r}
filename <- file.path(table_path, "logistic_cis_5_wild.tsv")
if (! file.exists(filename) || recompute) {
  logistic_cis_5_wild <- compute_cis(logistic_fit_5_wild)
  write_tsv(logistic_cis_5_wild, filename)
} else
  logistic_cis_5_wild <- read_tsv(filename)
```




```{r}
filename <- file.path(figure_path, "logistic_cis_5_wild.pdf")
g <- plot_cis(logistic_cis_5_wild, "Logistic model 5 wild UKBB", subtitle=model_stats(logistic_fit_5_wild))
ggsave(filename, g)
g
```

## Model 6

Comparison model between FinnGen and UKBB.

### Logistic regression

```{r}
variables6 <- c("sex", "age",
                "H63D", "S65C", "C282Y", "supersnip",
                "rs3094093", "rs114179634"
                #credible_snip_names2
                )
f6 <- sprintf("hemochromatosis ~ %s",
             paste(variables6, collapse = " + "))
f6
```


```{r}
logistic_mod_6 <- parsnip::logistic_reg()
logistic_mod_6
```

```{r}
logistic_recipe_6 <-
  recipe(as.formula(f6), data=train) %>%
  #step_relevel(hemochromatosis, ref_level = "control") %>%
  step_normalize(age) %>%
  step_mutate(C282Y = as.factor(round(C282Y)),
              H63D = as.factor(round(H63D))) %>%
  step_dummy(all_nominal_predictors(), H63D, C282Y) %>%
  step_interact(terms = ~ starts_with("C282Y") :
                  c(starts_with("H63D"), S65C, supersnip, starts_with("sex"))) %>%
  step_lincomb(all_numeric_predictors())
logistic_recipe_6
```

```{r}
logistic_wflow_6 <-
  workflow() %>%
  add_model(logistic_mod_6) %>%
  add_recipe(logistic_recipe_6)
logistic_wflow_6
```

#### Fit 

```{r}
filename <- file.path(result_path, "logistic_lf_6.rds")
if (! file.exists(filename) || recompute) {
  logistic_fit_6 <- logistic_wflow_6 %>%
    fit(data = train)
  saveRDS(logistic_fit_6, filename)
} else {
  logistic_fit_6 <- readRDS(filename)
}
logistic_fit_6
```

Fit on full data.

```{r}
filename <- file.path(result_path, "logistic_full_6.rds")
if (! file.exists(filename) || recompute) {
  logistic_full_6 <- logistic_wflow_6 %>%
    fit(data = basic_combined)
  saveRDS(logistic_full_6, filename)
} else {
  logistic_full_6 <- readRDS(filename)
}
logistic_full_6
```

#### Diagnostics

```{r}
broom::glance(logistic_fit_6)
```

Variance inflation factors.

```{r}
logistic_fit_6 %>% extract_fit_engine() %>% car::vif() %>% sort(decreasing = TRUE)
```

```{r}
logistic_fit_6 %>% extract_fit_engine() %>% plot()
```

```{r}
logistic_fit_6 %>% extract_fit_engine() %>% my_ggdiagnostics()
```

```{r}
#par(mfrow = c(3, 3))
fit <- logistic_fit_6 %>% extract_fit_engine()
termplot(fit, data=fit$data, partial.resid = TRUE, terms = "age", las=1, rug=TRUE)
```



```{r}
#fit %>% variable_plots(c("age", "sex_female"))
fit %>% variable_plots(c("age"))
```

Calibration.

```{r}
logistic_fit_6_test_pred <- augment(logistic_fit_6, new_data = test)
logistic_fit_6_test_pred %>%
  cal_plot_windowed(truth = hemochromatosis, estimate = .pred_case, step_size = 0.025, event_level = "second")
```

```{r}
logistic_fit_6 %>% extract_fit_engine() %>% fmsb::NagelkerkeR2()
```

#### Forest plots

```{r}
filename <- file.path(table_path, "logistic_cis_6.tsv")
if (! file.exists(filename) || recompute) {
  logistic_cis_6 <- compute_cis(logistic_fit_6)
  write_tsv(logistic_cis_6, filename)
} else
  logistic_cis_6 <- read_tsv(filename)
```

```{r}
filename <- file.path(table_path, "logistic_cis_full_6.tsv")
if (! file.exists(filename) || recompute) {
  logistic_cis_full_6 <- compute_cis(logistic_full_6)
  write_tsv(logistic_cis_full_6, filename)
} else
  logistic_cis_full_6 <- read_tsv(filename)
```



```{r}
filename <- file.path(figure_path, "logistic_cis_6.pdf")
g <- plot_cis(logistic_cis_6, "Logistic model 6 UKBB", subtitle=model_stats(logistic_fit_6))
ggsave(filename, g, width=7, height=7)
g
```

```{r}
filename <- file.path(figure_path, "logistic_cis_full_6.pdf")
g <- plot_cis(logistic_cis_full_6, "Logistic model 6 full UKBB", subtitle=model_stats(logistic_full_6))
ggsave(filename, g, width=7, height=7)
g
```

#### Risk table

Compute confidence intervals for predictions for multiple predictor combinations.

```{r}
data6 <- expand_grid(sex = c("male", "female"), age=mean(train$age), H63D=0:2, S65C=0, C282Y=0:2, 
                     supersnip=0:2, rs3094093=2, rs114179634=0, hemochromatosis = "case") %>%
  mutate(across(c(sex), as.factor),
         hemochromatosis = factor(hemochromatosis, levels=c("case", "control")))
important_predictors6 <- c(Sex="sex", "H63D", "C282Y", HLA3="supersnip") %>% fill_names()

cat("The following variables were fixed during the computation of probabilities\n")
fixed_predictors6 <- get_fixed_predictors(data6, important_predictors6)
fixed_predictors6
```

```{r}
# filename <- file.path(figure_path, "risk_table_6.pdf")
# p <- risk_table(logistic_fit_6, data6, important_predictors6, filename, title=sprintf("UKBB 6, %s", fixed_predictors6))
# plot(p, new_page=TRUE)
```

```{r, fig.width = 12, fig.height = 7}
title <- "Model 6 train data"
filename <- file.path(table_path, "risk_table_6_train.tsv")
if (! file.exists(filename) | recompute) {
  result  <- compute_risk_table(logistic_fit_6, data6, important_predictors6, count_data = train, col="sex", conf.level = 0.95)
  result2 <- compute_risk_table(logistic_fit_6, data6, important_predictors6, count_data = train, col="sex", conf.level = 0.80)
  result <- bind_rows(result, result2)
  write_tsv(result, filename)
} else {
  result <- read_tsv(filename)
}
p <- plot_risk_table(result, important_predictors6, title=title, col = "sex", conf.level = c(0.95, 0.80))
if (save_figs) {
  filename <- file.path(figure_path, "risk_table_6_train.pdf")
  pdf(filename, width=12, height=7, title=title)
  plot(p[[1]], new_page=TRUE)
  plot(p[[2]], new_page=FALSE)
  invisible(dev.off())
}
#dev.new()
plot(p[[1]], new_page=TRUE)
plot(p[[2]], new_page=FALSE)
```

```{r, fig.width = 12, fig.height = 7}
title <- "Model 6 full data"
filename <- file.path(table_path, "risk_table_6_full.tsv")
if (! file.exists(filename) | recompute) {
  result  <- compute_risk_table(logistic_full_6, data6, important_predictors6, count_data = basic_combined, col="sex", conf.level = 0.95)
  result2 <- compute_risk_table(logistic_full_6, data6, important_predictors6, count_data = basic_combined, col="sex", conf.level = 0.80)
  result <- bind_rows(result, result2)
  write_tsv(result, filename)
} else {
  result <- read_tsv(filename)
}
p <- plot_risk_table(result, important_predictors6, title=title, col = "sex", conf.level = c(0.95, 0.80))
if (save_figs) {
  filename <- file.path(figure_path, "risk_table_6_full.pdf")
  pdf(filename, width=12, height=7, title=title)
  plot(p[[1]], new_page=TRUE)
  plot(p[[2]], new_page=FALSE)
  invisible(dev.off())
}
#dev.new()
plot(p[[1]], new_page=TRUE)
plot(p[[2]], new_page=FALSE)
```

### Cox regression

```{r}
variables6_cox <- c("sex",
                "H63D", "S65C", "supersnip",
                "rs3094093", "rs114179634",
                "C282Y", 
                #credible_snip_names2
                "C282Y:(H63D + S65C + supersnip + sex)"
                )
f6_cox <- sprintf("Surv(survival_age, hemochromatosis) ~ %s",
             paste(variables6_cox, collapse = " + "))
f6_cox
```

#### Fit

```{r}
cox_train <- train %>% mutate(hemochromatosis = as.integer(hemochromatosis == "case"),
                              across(c(C282Y, H63D), as.factor))
filename <- file.path(result_path, "cox_fit_6.rds")
if (! file.exists(filename) || recompute) {
  cox_fit_6 <- survival::coxph(as.formula(f6_cox), data=cox_train)
  saveRDS(cox_fit_6, filename)
} else {
  cox_fit_6 <- readRDS(filename)
}
summary(cox_fit_6)
```

Fit on full data.

```{r}
cox_full <- basic_combined %>% mutate(hemochromatosis = as.integer(hemochromatosis == "case"),
                              across(c(C282Y, H63D), as.factor))
filename <- file.path(result_path, "cox_full_6.rds")
if (! file.exists(filename) || recompute) {
  cox_full_6 <- survival::coxph(as.formula(f6_cox), data=cox_full)
  saveRDS(cox_full_6, filename)
} else {
  cox_full_6 <- readRDS(filename)
}
summary(cox_full_6)
```

#### Diagnostics.

Test the proportinal hazards assumption.

```{r}
cox_ftest_6 <- cox.zph(cox_fit_6)
cox_ftest_6
```

Only rs3094093 (HLA2) seems to deviate from the proportional hazards assumption.
After fixing survival age, more deviations appeared.

```{r, fig.width=14, fig.height=14}
# w <- tryCatch(
#   warning = function(w) {invokeRestart("muffleWarning")}, 
#   quiet=TRUE,
#   ggcoxzph(cox_ftest_6)
# )
# w <- withCallingHandlers(warning = function(w) {},
#   ggcoxzph(cox_ftest_6)
# )
# logWarning <- function(w)
# {
#   #cat("Hola\n")
# }
# x <- tryCatch(
#   withCallingHandlers(ggcoxzph(cox_ftest_6), 
#                       warning = function(w) logWarning(w),
#                       quiet = TRUE),
#   quiet=TRUE,
#   error=function(cnd) return(cnd)
# )
grid.newpage()
print(suppressWarnings(ggcoxzph(cox_ftest_6)), newpage=FALSE)
```

Following does not work for some reason. Causes R to hang.

```{r, eval=FALSE}
ggcoxdiagnostics(cox_fit_6,
                 type = "deviance",
                 ox.scale = "linear.predictions")
```

Full data.

```{r}
cox_ftest_full_6 <- cox.zph(cox_full_6)
cox_ftest_full_6
```

```{r, fig.width=14, fig.height=14}
grid.newpage()
print(suppressWarnings(ggcoxzph(cox_ftest_full_6)), newpage=FALSE)
```

#### Forest plots

Extract confidence intervals.

```{r}
filename <- sprintf("%s/cox_cis_6.tsv", table_path)
if (recompute || ! file.exists(filename)) {
  cox_cis_6 <- tidy(cox_fit_6, conf.int = TRUE, exponentiate = TRUE) %>% 
    drop_na(estimate)    # Some interacting levels don't exist in the date, we drop them here
  write_tsv(cox_cis_6, filename)
} else {
  cox_cis_6 <- read_tsv(filename)
}
```

```{r}
filename <- sprintf("%s/cox_cis_full_6.tsv", table_path)
if (recompute || ! file.exists(filename)) {
  cox_cis_full_6 <- tidy(cox_full_6, conf.int = TRUE, exponentiate = TRUE) %>% 
    drop_na(estimate)    # Some interacting levels don't exist in the date, we drop them here
  write_tsv(cox_cis_full_6, filename)
} else {
  cox_cis_full_6 <- read_tsv(filename)
}
```

Make forest plots.

```{r}
g <- plot_cis(cox_cis_6, "Cox model 6 UKBB") +
  labs(x="Hazard ratio") +
  theme(axis.title.y = element_blank())
if (save_figs) {
  filename <- sprintf("%s/cox_cis_6.pdf", figure_path)
  ggsave(filename,  g, width=180, height = 120, units="mm", dpi=300, scale=1.0, device="pdf")
}
g
```

```{r}
g <- plot_cis(cox_cis_full_6, "Cox model 6 full UKBB") +
  labs(x="Hazard ratio") +
  theme(axis.title.y = element_blank())
if (save_figs) {
  filename <- sprintf("%s/cox_cis_full_6.pdf", figure_path)
  ggsave(filename,  g, width=180, height = 120, units="mm", dpi=300, scale=1.0, device="pdf")
}
g
```

## Old stuff

### Precision-recall

#### Training set.

```{r, eval=FALSE}
compute_helper <- function(model, newdata=NULL, boot.n=2000) {
  pred <- augment(logistic_mod_1, type.predict = "link")
  g <- pred %>% ggplot(aes(.fitted)) + geom_histogram()
  distribution_df <- pred %>% select(hemochromatosis, .fitted) %>% summary()
  res <- create_precision_recall_new(pred$hemochromatosis, pred$.fitted, boot.n=boot.n)
  list(distribution=g, distribution_df=distribution_df, pr_plot=res$pr_plot, pr_ci=res$pr_ci)
}
```

```{r, eval=FALSE}
filename <- file.path(result_path, "res_train_1.rds")
if (! file.exists(filename)) {
  res_train_1 <- compute_helper(logistic_mod_1, boot.n=boot.n)
  saveRDS(res_train_1, filename)
} else 
  res_train_1 <- readRDS(filename)
res_train_1
```





#### Test set.

```{r, eval=FALSE}
filename <- file.path(result_path, "res_test_1.rds")
if (! file.exists(filename)) {
  res_test_1 <- compute_helper(logistic_mod_1, model_test_data, boot.n=boot.n)
  saveRDS(res_test_1, filename)
} else 
  res_test_1 <- readRDS(filename)
res_test_1
```









